### 2019-03-06 Proposal.Rmd
small, so small

---
title: "Statistical methods in cancer bioinformatics and tumor evolution"
author: "Brian Mannakee"
date: "`r format(Sys.time(), '%d %B, %Y')`"

header-includes:
  - \usepackage{ragged2e}\RaggedRight
  - \usepackage{wrapfig}

output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    keep_tex: yes

documentclass: article
mainfont: Arial
fontsize: 12pt
geometry: margin=1in
bibliography: variantcalling.bib
---

# Introduction

Recent advances in sequencing technologies and increasing availability of sequenced tumors present an opportunity to expand our understanding of the characteristic processes acting in cancer [@Hu2017].
Tumors are generally considered to arise from a single cell which acquires metastatic potential and is the ancestor of all cells in the tumor [@Nowell1976;@Fearon1989].
Clonal evolutionary processes are well studied in cancer [@Bozic2010;@Bozic2016], and the theory leads to three potential models for tumor evolution.
Tumors can evolve as a terminal expansion subject to little or no selective pressure, leading to a so-called *Big Bang* which is characterized by a large number of heterogeneous subclones[@Sottoriva2015].
Tumors evolving under selective pressure can have one of two dynamics: either they evolve in such a way that advantageous mutations arise and compete with each other leading to multiple dominant subclones in the tumor (branched evolution) [@Yates2012;@Gerlinger2012;@Burrell2013;@Bozic2016]; or they evolve such that each new advantageous mutation out-competes all previous mutations leading to a single dominant clone (sequential evolution) [@Hu2017].
These processes lead directly to the mutational profile of the tumor, i.e. mutations (variants), their genomic contexts, and their frequencies.

Identification of the mutations present in a tumor can be critical in optimizing the treatment regime for an individual patients disease [@Ding2012;@Mardis2012;@Chen2013;@Borad2014;@Findlay2016].
Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by NGS.
@Griffith2015 demonstrate conclusively that identification of all major resistance mutations present in a tumor is essentially impossible with current sequencing practice.
They find that rather than the current standard of sequencing a tumor exome at 75X-100X depth, sequencing of up to 400X is required to identify important subclonal mutations in a heterogeneous tumor.
In addition, they find that to identify the mode and tempo of evolution in a tumor, whole genome sequences of 300X are required [@Griffith2015].
Unfortunately, modern variant callers suffer from excessive false positive rates at extreme sequencing depths [@Cibulskis2013].
@Griffith2015 found that in order to generate a reliable set of variant calls for a whole genome sequenced at 312X depth they needed to combine the calls from eight different variant callers and then resequence 200,000 individual variants.
In this proposal I develop two methods; a novel variant calling algorithm with a significantly reduced false positive rate for deep sequencing data, and an algorithm that identifies the timing of one of the characteristic processes operating in tumor evolution.
<!-- This assumption leads directly to the inference that tumors evolve through a process of branched evolution from a single clone[@Burrell2013;@Yates2012]. -->
<!-- Branched evolutionary processes are well studied from a both a theoretical and experimental standpoint so that genetic information from the tumor, in particular distribution of observed variant allele frequencies, can be used to analyze the evolutionary trajectory of the tumor and test hypotheses regarding the mutation rate, mutational processes, and the strength of natural selection, operating in the tumor[@Somarelli2017]. -->
<!-- Important applications of this include a better understanding of the drivers behind tumor initiation and progression, the prediction of treatment resistance mechanism, and treatment strategies for combatting the evolution of resistance. -->
<!-- The quality and applicability of these predictions depends heavily on the sensitivity and specificity of the algorithms used to identify variants in tumor samples. -->

<!-- For any given tumor, the key to discerning the evolutionary process, as well as identifying resistant clones at low frequency, is variant calling. -->
<!-- Variant calling is the process of converting the output of next generation sequencing (NGS) to a set of mutations that are present in the tumor. -->

<!-- The current statistical paradigm in variant calling is to assume that the data generating process for tumor mutations is uniform, i.e. the mutation rate, allele frequency, and probability of a mutation at a given site are all assumed to be uniform. -->
<!-- These are known to be incorrect assumptions. -->
<!-- Mutations are generated by mutational processes that have specific preferences for sites depending on their genomic context, and result in varying but identifiable mutation rates at those preferred sites [@Nik-Zainal2012a;@Alexandrov2013;@Alexandrov2013a;@Helleday2014]. -->
<!-- Mutation allele frequencies are determined by the mode and tempo of evolution in the tumor. -->
<!-- As a result, the data generating process that results in the collection of mutations present in a tumor is far from uniform, and this prior information can be of significant value in interpreting the results of NGS sequencing data. -->

<!-- Here I propose to more fully integrate the idea of a *Generative Distribution* for observed mutations in tumors, and create a collection of methods that support the identification of the processes driving a tumor. -->
<!-- These methods simultaneously reinforce each other, and will lead to better resolution of the generative distribution (Figure \ref{figmain}). -->
<!-- \begin{wrapfigure}{r}{.6\textwidth}  --> 
<!-- \includegraphics[width=.6\textwidth]{./proposal_main.png}  --> 
<!-- \caption{Theory and method advances toward a generative model of tumor mutation.}  --> 
<!-- \label{figmain}  --> 
<!-- \end{wrapfigure}  --> 
<!-- The first will be a variant caller that uses the empirical distribution of mutations with high allele frequency to generate an informative prior that can be used to increase the sensitivity of the variant caller for low frequency mutations. -->
<!-- The second method will use these improved variant calls to investigate the evolution of the mutation spectrum over time in the tumor, allowing finer grained estimates of mutation rate and the effects of natural selection. -->
<!-- The third method will recast the procedure for identifying mutational processes from a classification context ([@Lee2000;@Alexandrov2013]) to a topic modelling context (cite), generating sharper prior estimates for the probability of a mutation at a give site. -->
<!-- Finally, the fourth method will combine the concepts of mutational processes and tumor evolution to create software to *Generate*, rather then *simulate*, tumor genomes. -->

# Background

## Sequencing
Next generation sequencing technology combines three separate processes:

1. DNA extraction and library preparation
    - DNA is extracted from cells, fragmented, and 5' and 3' adapters are ligated to each fragment. DNA fragments prepared in this way are called the library.
    - The library is washed onto a flow cell where fragments hybridize to the surface.
    - Hybridized fragments are amplified into clusters of identical fragments on the flow cell surface.
2. Sequencing
    - Sequencing reagents and flourescent-tagged DNA bases are washed over the flow cell in cycles.
    - In each cycle one DNA base is added to the hybridized *templates* and flourescence is measured for each cluster.
    - Repeating this cycle results in a *read* of the DNA template at each cluster, and the number of cycles corresponds to the read length.
3. Alignment and Data Analysis
    - Following sequencing, reads are aligned to a reference genome.
    - These alignments are then used for a variety of purposes, including somatic nucleotide and copy number variant calling

Each of these steps is a source of error in the process.
The errors in steps one and two are inherent to the sequencing technology used, and the manufacturers of each have done extensive testing to develop an error model for these two steps taken together.
The sequencer computes the probability that each base in a read has been called in error, and reports a *base quality score* $q$, such that $q= -10*log_{10}(error\ probability)$.
In step three uncertainty arises from the difficulty in aligning short reads (~100-150 base pairs) to the genomic sequence, due to the likelihood of highly similar sequences of this length occuring at more than one place in the genome.
Each alignment algorithm generates a mapping quality score (MAPQ) for every read.
The score ranges from 0-40 and is typically simply used as a heuristic filter, such that reads with a MAPQ score less than some value are ignored.
An excellent overview of the NGS sequencing process is available on the [Illumina website](https://www.illumina.com/content/dam/illumina-marketing/documents/products/illumina_sequencing_introduction.pdf) and also in [@Escalona2016].

## Mutation spectrum

Cells that are functioning normally have mechanisms that monitor the DNA during replication and correct potential errors so that they are not transmitted to daughter cells.
As a result the accumulation of mutations during somatic cell division is low in healthy cells.
In addition cells have mechanisms that control their rate of division so that tumors do not arise.
Oncogensesis is thought to involve the loss of both of these mechanisms, leading to a proliferation of cells that accumulate significant numbers of mutations [@Fearon1989].
Failure of individual repair mechanisms, as well as carcinogen exposure, tend to be associated with specific patterns of mutation.

Using the pyrimidine of each Watson-Crick base pair we can label all possible nucleotide substitutions, resulting in 6 subtypes: C>A,C>G,C>T,T>A,T>C,T>G.
As many repair mechanisms and mutagens are also affected by the identify of the bases flanking the substituted nucleotide, we can further sub-divide mutations in terms of their 5' and 3' flanking bases, to create a total of 96 different mutation types (4 5' bases * 6 subtypes * 4 3' flanking bases).
These mutation subtypes are often described as mutation *contexts*.
The mutation spectrum of a particular tumor is represented by the proportion of mutations occuring in each of the 96 contexts.
Signatures of specific mechanisms and mutagens can be extracted from large collections of variant calls taken from thousands of tumors [@Nik-Zainal2012a;@Alexandrov2013;@Alexandrov2013a;@Helleday2014].
The complete set of mutational signatures collected by the COSMIC project is available at [http://cancer.sanger.ac.uk/cosmic/signatures](http://cancer.sanger.ac.uk/cosmic/signatures).
An example of one of the signatures is shown in Figure \ref{fig1}.

\begin{figure} 
\includegraphics{./Signature1.png} 
\caption{COSMIC mutational process signature number 1. This signature is found in nearly all tumors, and is generated by spontaneous deamination of 5-methylcytosine. The deamination turns a cytosine to uracil, and during the next replication cycle the uracil is read as a thymine and paired with an adenine, leading to C>T transitions in specific genomic contexts. This signature is strongly associated with age at diagnosis.} 
\label{fig1} 
\end{figure}

# Projects
## Somatic single nucleotide variant calling 
<!-- Variant callers show significant performance degradation as sequencing depth increases, so there are very few whole exome sequences at depths > ~150X. -->
<!-- The goal of the variant calling project is to allow deeper sequencing with better control over the false positive rate, leading to better resolution of low frequency mutations.  -->
<!-- As costs come down, it may become affordable, and thus desirable, to go much deeper. -->

The data available to call variants are NGS reads from matched tumor/normal samples.
Several statistical methods for estimating the probability that an apparent somatic variant in the tumor is truly present have been developed.
One popular algorithm computes a simple one-sided Fisher's exact test comparing the read counts in the tumor and normal samples with those expected under a model where sequencing error is 0.01\% at every position, and calls the site variant if the P-value from the test is less than 0.10 [@Koboldt2012].
Other models jointly compute Bayesian posterior probabilities for the tumor and normal genotype with models of varying complexity [@Larson2012;@Saunders2012;@Hansen2013;@Christoforides2013].

Here I extend the model used by MuTect, developed by the Broad Institute, which balances complexity and simplicity while performing as well or better than other algorithms over a broad range of tumor types and sequencing depths [@Cibulskis2013].
First I will describe the MuTect model, and then propose an extension using the mutation spectrum of the tumor to construct a prior for the probability of mutation at a given site.

Notation:
\[
  \begin{aligned}
   r &\in \{A,C,G,T\} &\quad \textrm{reference allele}\\
   m &\in \{A,C,G,T\}, m \ne r &\quad \textrm{variant allele}\\
   f &\in [0,1] &\quad \textrm{variant allele fraction}\\
   b_i &\quad &\textrm{base called at read i}\\
   e_{b_i} &= 10^{-\frac{q_{b_i}}{10}},  q \in [0,\infty] &\quad \textrm{the probability of error at base } b_i\\
   i &= (1 \dots d) &\quad \textrm{index over d reads}.\\\\
  \end{aligned}
\]

The probability that a given base is correctly called can be written as

\[ 
  P(b_i \mid e_i, r, m, f) = \left\{
    \begin{array}{cr}
      f \frac{e_{b_i}}{3} + (1-f)(1-e_{b_i}) & b_i = r\\
      f(1-e_{b_i}) + (1-f) \frac{e_{b_i}}{3} & b_i = m\\
      \frac{e_{b_i}}{3} & otherwise.
    \end{array}
    \right.
\]

Now consider two models for the data. Model $M_0$ in which there are no variants at a site, and $M^{m}_{f}$ where allele $m$ is present at allele fraction $f$.
Assuming reads are independent the likelihood of the model given the data is

\[
  \mathcal{L}(M^{m}_{f}) = P(\{ b_i \} \mid \{ e_{b_i} \}, r, m, f) = 
  \prod_{i=1}^{d} P(b_i \mid e_{b_i}, r, m, f)
\]

and the probability of $M^{m}_{f}$ can be written

\[ 
  P(m, f \mid \{ b_i \}, \{ e_{b_i} \}, r) = 
      \mathcal{L}(M^{m}_{f}) \frac{P(m,f)}{P({\{ b_i \}} \mid {\{ e_{b_i} \}}, r)}.
\]

Because we are ultimately going to evaluate the odds in favor of $M^{m}_{f}$ given the data, it is useful to note that we can write this probability in terms of the model $M_0$

\[ 
  1 - P(m, f \mid \{ b_i \}, \{ e_i \}, r) = 
      \mathcal{L}(M_{0}) \frac{1 - P(m,f)}{P({\{ b_i \}} \mid {\{ e_{b_i} \}}, r)}.
\]

Taking the log of the ratio of the two previous equations gives the log odds in favor of $M^{m}_{f}$, and some cancellation yields the following expression
\[
  LOD_{T}(m,f) = \textrm{log}_{10} \left(\frac{\mathcal{L}(M^{m}_{f})P(m,f)}{\mathcal{L}(M^{m}_{0})(1-P(m,f))} \right).
\]

We can construct a classifier for variants by selecting an odds threshold $\delta_T$ and labeling variants satisfying the condition
\[
  LOD_{T}(m,f) = \textrm{log}_{10} \left(\frac{\mathcal{L}(M^{m}_{f})P(m,f)}{\mathcal{L}(M^{m}_{0})(1-P(m,f))} \right)
    \ge \textrm{log}_{10} \delta_T
\]
as true variants, and rejecting them otherwise.
Note that the expression for $LOD_{T}(m,f)$ can be further factorized as the sum of the log-likelihood ratio of the two models and the log odds of the prior for $M^{m}_{f}$.
Current variant callers calculate this prior by assuming the allele and its frequency are independent, and that $f \sim \textrm{U}(0,1)$, so that $P(f) = 1$.
If all substitutions are equally likely, then $P(m) = \mu/3$ where $\mu = 3 \times 10^{-6}$, the estimated per-base mutation rate in tumors.
Given these assumptions the log prior odds is a constant, and the classifier can be re-written as

\[
  LOD_{T}(m,f) = \textrm{log}_{10} \left(\frac{\mathcal{L}(M^{m}_{f})}{\mathcal{L}(M^{m}_{0})} \right) \ge
    \textrm{log}_{10} \delta_T - \textrm{log}_{10} \left(\frac{P(m)}{1 - P(m)} \right) \ge \theta_T.
\]

If $\delta_T = 2$, i.e the odds in favor of $M^{m}_{f}$ is 2, then $\theta_T = 6.3$.

\begin{figure} 
\begin{center}
\includegraphics[width=\textwidth]{./proposal_empirical} 
\caption{Empirical mutation spectrum of a pancreatic ductal adenocarcinoma, whole exome sequenced to 50X mean depth.} 
\label{fig2} 
\end{center}
\end{figure}

Figure \ref{fig2} shows the empirical mutation spectrum from a single tumor.
It is clear that the assumption that the mutation probability $P(m)$ is the same at every base in the genome is not supported by the data.
We can better describe the probability $P(m)$ as $P(m,M)$, the probability that there is a mutation $M$ and that the mutation is to allele $m$. 
Further, Figure \ref{fig2} makes it clear that this probability $P(m,M)$ depends strongly on the genomic context in which we find the mutation.
Thus, what we would like to use as a prior is $P(m,M \mid C)$, the probability of a mutation to a particular allele at a locus given the sequence context of that locus.
By the laws of conditional probability we can rewrite $P(m,M \mid C)$ as 
\[
  P(m,M \mid C) = P(m \mid M,C)P(M \mid C).
\]

The conditional probability $P(m \mid M,C)$ can be computed from the empirical data in Figure \ref{fig2}, but $P(M \mid C)$ can not be.
Using Bayes rule we can rewrite$P(m,M \mid C)$ as

\[
  P(m,M \mid C) = P(m \mid M,C)P(C \mid M) \frac{P(M)}{P(C)}.
\]

Now $P(C \mid M)$ can also be computed from the data in Figure \ref{fig2}. 
$P(C)$ can be estimated as $1/96$ where 96 is the total number of contexts, and $P(M)$ can be estimated as above as $\mu$.
The new expression for the log odds is

\[
  LOD_{T}(m,f) = \textrm{log}_{10} \left(\frac{\mathcal{L}(M^{m}_{f})P(m,M \mid C)}{\mathcal{L}(M^{m}_{0})(1-P(m,M \mid C))} \right).
\]

As a proof of concept, I have implemented an algorithm that computes $P(C \mid M)$ and $P(m \mid M,C)$ directly from the empirical distribution of the tumor represented by Figure \ref{fig2}.

\begin{wrapfigure}{r}{.5\textwidth} 
\includegraphics[width=.4\textwidth]{./mutect_prior_performance} 
\caption{Comparing the level of evidence seen by MuTect against the corrected level of evidence from the algorithm including the empirical prior, binned by allele frequency. The horizontal line is the evidence threshold required to call a variant using my algorithm, and the vertical line is the threshold used by MuTect. Variants in quadrant I are called only by my algorithm, those in quadrant II are called by neither, those in quadrant III are called by both, and variants in quadrant IV are called by MuTect, but rejected by the new algorithm.} 
\label{fig3} 
\end{wrapfigure} 
The performance of this method is summarized in Figure \ref{fig3}.
For allele frequencies above 20\% the two algorithms return identical results.
At allele frequencies below 20\% the algorithm with the empirical prior probability rejects a large number of variants called by MuTect (quadrant IV), and accepts a few variants that are rejected by MuTect (quadrant 1).
<!-- These results suggest that the method has promise both in reducing the number of false positives, and increasing the evidence for some low frequency variants that are supported by the prior. -->
Note the banding of points along the Y-axis, which is due to the implementation of the non-uniform prior.
The empirical prior tends to reduce the level of evidence for most variants compared to the uniform prior, pushing a large number of points into quadrant IV, and potentially reducing the false positive rate.

Using the observed mutation spectrum as a prior, while better than assuming a uniform prior, is far from ideal.
Whole exome sequencing (WES) is the most common form of NGS in cancer sequencing both because it is cheaper than sequencing the whole genome and the genomic changes of interest to researchers and clinicians are those that actually change protein composition or expression patterns, and thus might be targetable by some therapy. 
In a typical WES between 100 and 300 single nucleotide variants are identified, and both biological and measurement noise are high.
This leads to a high probability that some mutation contexts will have zero observations in the data (300 observations spread over 96 bins with an expectation of high concentration in a subset of bins).
Since the prior can't enforce zero probability of a mutation in a given context I have renormalized the observed spectrum so that every bin has at least $10^{-3}$ probability, which is somewhat arbitrary and probably not representative.
Additionally, the prior here is in some sense intended to represent the underlying data generating process that leads to mutations, but the small amount of data available means the observed mutation spectrum a likely a poor estimate.

The work of the COSMIC consortium cataloging mutation signatures provides another means of generating the mutation prior [@Rosenthal2016].
The prior can be generated by identifying the linear combination of mutation signatures that most closely approximate the true data, and using this model to generate a mutation spectrum. 
Let the vector $M (96 \times 1)$ contain the frequencies of each of the 96 mutation types (i.e. the data in Figure \ref{fig2}) in the tumor.
Let $P (96 \times N)$ be the frequency profiles of $N$ pre-determined mutation signatures ($N$ = 30 if we use the COSMIC signatures).
We wish to estimate the vector $E (N \times 1)$ of exposures to each mutational signature.
The least squares estimate for $E$ is given by

\[
min_E(M - PE)^T(M-PE)
\]

subject to the constraint that the rows of E sum to one, i.e. the rows of $PE$ contain the linear combination of the K mutation types that most closely approximates $M$.
When $P$ is full rank, as it is here, there is an exact solution to this constrained optimization using quadratic programming [@Huang2017].
The optimization can also be accomplished via simulated annealing, which allows for investigation of a distribution of solutions that are within say 1\% of the optimum in order to estimate the stability of the solution.
I will extend the current preliminary algorithm to compute the mutation prior either from the normalized empirical spectrum or the estimated spectrum, and determine if there is any performance difference between the two methods.
The performance of these methods will be evaluated using simulations as well as gold standard variant call sets as described below.

## Evolution of the mutation spectrum
Cancer is an evolutionary process and modelling the mode and tempo of evolution in initiation and progression, as well as the role of evolution in the development of treatment resistance, are crucial to developing treatments [@Findlay2016].
In particular, an understanding of the nature of tumor heterogeneity and selection on subclones is crucial for understanding the risk of mutation driven escape in drug treatment regimes[@Greaves2015].
There is significant interest in using mutation spectra to catalogue the roles of particular mutational process exposures in initiation and progression of cancer [@Hollstein2017].

\begin{figure} 
\begin{center}
\includegraphics[height=.3\textheight]{./proposal_sigs} 
\caption{Changes in mutation signatures between the full collection of variants (Full), and only those with allele frequency below 20\% (Subclonal)} 
\label{fig4} 
\end{center}
\end{figure}

<!-- One way to assess heterogeneity and look for signatures of clonal selection is simply to generate a set of high-confidence variant calls for the tumor sample and see if they can be clustered by allele frequency in such a way that the data is consistent with the emergence of defined sub-clones. -->
<!-- The drawback to this method is that it requires measuring allele frequencies at a resolution that is out of reach at typical sequencing depth. -->
<!-- In addition, in many tumors allele frequencies are confounded by amplifications and deletions in such a way that it can be difficult to determine the fraction of cells that carry a mutation from the NGS allele frequency estimates. -->

Two recent studies in lung cancer found signatures of cigarette smoking and APOBEC3B hyper-activity at similar exposure intensities in multiple samples.
On closer inspection however, they found that the APOBEC signature was absent in the early (clonal) mutations, and dominant in the later (subclonal) mutations [@DeBruin2014;@Zhang2014].
The implication of this work is that smoking is the main source of mutational insult prior to initiation, but during progression APOBEC3B drives the accumulation of mutations.
We have preliminary results, shown in Figure \ref{fig4}, suggesting that in pancreatic cancers we also see shifts in mutational process as we look at mutations of different frequency.
In a recent preprint, @Rubanova2018 show similar results on a large dataset of tumor mutations in a variety of cancer types.
All of these results were generated simply by using the matrix factorization method of @Rosenthal2016 described above, applied to variants subsetted according to their allele frequency.

<!-- If tumors are evolving neutrally, with important drivers and mutational process present at initiation, we would expect to uncover the same mutation signatures whether we look at the entire set of mutations, or only the subset of low frequency mutations. -->
<!-- We see both new signatures, and changes in the balance of signatures, arising at lower allele frequencies. -->
<!-- This initial work provides evidence for the presence of selection pre-treatment, and against the Big Bang model, but suffers from serious data and methodological limitations. -->

There are far more sophisticated methods available to look for changes in the mutational processes during the course of tumor evolution.
As a first approximation it is reasonable to consider the allele frequency as a proxy for the time at which a mutation arose, such that mutations at low frequency occurred later in the tumors evolution.
This opens up the possibility of applying time series methods, such as change point detection, to determine whether the underlying data generating (i.e. mutational) processes have changed over the course of the tumors growth [@Zou2014].
Even more promising is to think of the problem in light of the document classification and multiple author identification literature.
The tumor mutation spectrum is the result of one or more mutational processes, and thinking about those different processes combining as a mixture of multinomial distributions opens up a substantial body of literature on detection of change points in categorical data [@Wolfe1990;@Stoffer1993;@Giron2005;@Batsidis2013;@Wang2017].
While the NMF procedure used by the COSMIC consortium to uncover the mutation signatures is nominally a document classification algorithm, I am unaware of any work in this field in which the mutation spectrum of the tumor is explicitly treated as sequential draws from a possibly changing mixture of multinomial distributions, and statistical methods developed for these processes applied.
Work evaluating these methods in this context will be a contibution to cancer informatics, and be very useful in the effort to understand the evolutionary processes responsible for the heterogeneity we see in cancer.
These statistical methods will be evaluated using simulations as well as published sequence and mutation spectrum heterogeneity data described below.

<!-- ## Identification of mutation spectra -->
<!-- The COSMIC mutation signatures we use to construct a prior for variant calling above were generated using non-negative matrix factorization (NMF)[@Lee2000;@Alexandrov2013]. -->
<!-- Using NMF to identify mutational signatures casts the problem in the topic modelling context used to classify documents in a corpus by the similarity of the topics they include. -->
<!-- It has been shown that NMF is mathematically approximately equivalent to a specific parameterization of Latent Dirichlet Analysis (LDA), the most common topic model[@Faleiros2016]. -->
<!-- In particular, NMF algorithms which minimize KL divergence are an instance of LDA where the dirichlet priors on the distributions of both words over topics and topics over documents are uniform (symmetric with concentration parameters = 1). -->
<!-- This construction imposes several drawbacks in the context of using these mutation signatures for inference in the variant calling context, that I propose to use more sophisticated statistical methods to alleviate. -->

<!-- The mutation signatures in Figure~\ref{fig1} are thought to be indicative of the activity of various mutational process, as describe in the background section. -->
<!-- While they are widely used these mutation signatures have several properties which may affect the quality of inferences using them. -->
<!-- We would expect mutational signatures to be sparse in some sense, i.e. a real mutational "process" should be operating on specific contexts. -->
<!-- While we see several signatures that have this sort of sparseness, many are not as specific as one would assume. -->
<!-- For instance, signatures 3,4,5,8, and 9 are all signatures where nearly every context has positive weight, yet none has more than 5%. -->
<!-- There is certainly an underlying random mutational process that occurs in all somatic cell divisions, and it appears that this random process is being distributed among many signatures. -->
<!-- This is analagous to what have been called "Stop" words in the topic modelling literature. -->
<!-- If we naively feed a corpus of documents into the LDA algorithm, every topic will include words such as "the" and "and", so the general practice is simply to remove these common words from the final topics. -->
<!-- In the case of mutational signatures, this is not a reasonable thing to do because it is not possible to know a priori which contexts are truly extraneous to the signature. -->
<!-- Wallach and colleagues have proposed a modification of the LDA algorithm that allows for the estimation of dirichlet priors that are both asymmetric and non-uniform [-@Wallach2009]. -->
<!-- They show that this tends to alleviate the "Stop" word problem by concentrating commonly used words in one or two common topics. -->
<!-- In the mutation signature context this would likely equate to confining the random mutation probabilities to Signature 1, the aging signature, while making the rest of the signatures more sparse. -->

<!-- There must be some correlation structure between mutation signatures. -->
<!-- Signatures 2 and 13 are signatures of the activity of the APOBEC family of cytidine deaminases involved in viral defense, and are often found together in cervical and bladder cancers. -->
<!-- These two signatures are obviously strongly correlated. -->
<!-- Conversely, Signature 4 (Tobacco smoking) and Signature 7 (UV radiation) should be anti-correlated to the point of mutual exclusion because UV radiation doesn't penetrate past the skin layer and tobacco smoke effects only the internal organs. -->
<!-- The LDA framework can not estimate the correlation structure among signatures, and as a result they are not available for use when attempting to dissect the signature contributions of a single tumori -->
<!-- Aitchison identified this problem in the compositional data literature long ago and suggested the logistic normal distribution as a method for modeling the correlation structure of mixed member models[@Aitchison1982].  -->
<!-- Subsequently, @Paisley2012 built on Aitchison's work and conceived the Discrete Infinite Logistic Normal distribution, which for the estimation of correlation structure in mixed member models, including topic models. -->

<!-- I propose to develop an algorithm that utilizes the @Paisley2012 DILN model with assymetric and non-uniform dirichlet process priors to estimate mutation spectra from the same TCGA data used by @Alexandrov2013. -->
<!-- I am not aware of any literature considering this approach to mutation signature identification, and it will be useful in both evolutionary and variant calling methods described above. -->

## Simulation method for validating variant calling and mutation signature evolution methods

\begin{figure} 
\begin{center}
\includegraphics[height=.3\textheight]{./simulation} 
\caption{Branching process simulation scheme, with examples of some of the scenarios it can create. 
A: Every node divides at every time step with probability 1. $N$ mutations drawn from a mutational signature $M$ are added to each leaf, where $N$ is drawn from a Poission distibution with parameter $\lambda$ and $M$ is a combination of the COSMIC mutation signatures. 
B: A new mutation signature arises at time step $T_3$ but there is no corresponding advantageous mutation, so the number of mutations and the probability of division at each time step remain the same. 
C:  Two lineages gain the same reproductive advantage at time step $T_3$, and form a final population with two major clones. 
D:  A  new mutational process arises alongside an advantageous mutation at time step $T_3$ and the mutation and mutational process sweep to fixation.} 
\label{figsim} 
\end{center}
\end{figure} 
Distinguishing low frequency resistance mutations requires sequencing tumors to high depth [@Griffith2015].
Sequencing whole genomes to high depth, and/or sampling multiple tumor regions, is necessary to identify evolutionary processes [@Ding2012;@Sun2017;@Somarelli2017;@VandenEynden2017].
Deep whole genome sequencing of tumors is rare because the cost can not typically be justified by clinicians and researchers, due to high false positive rates at high depth and the lack of actionable mutations in the non-exonic portions of the genome [@Griffith2015].
Tumor simulations provide the opportunity to test variant calling and mutation signature inference methods at varying depths and genome coverages in a setting where the ground truth is known.
Many tumor genome simulators exist, but none generates clonal structure in a realistic way, and none allow for changing mutation signatures during the course of tumor development.
I will develop a tumor simulation method that generates mutation sets with clonal structure for both mutations and mutation signatures.
<!-- Simulation of synthetic tumor/normal genome pairs will be necessary to characterize the performance and attributes of both the tumor variant calling and mutation signature evolution methods. -->

Clonal evolution from a single cell can be simulated via a simple branching process [@Durrett2014].
This branching process can then be represented as a graph where each node representing a cell carries various attributes, and leaves inherit attributes as they are added.
Figure \ref{figsim} shows the way these graphs can generate various clonal architectures.
Once the graph is generated, selecting a sample of adjacent leaves (representing the small fractions that are typically sequenced from bulk tumor samples) and traversing the graph back to the origin yields the full set of mutations and their frequencies in the sample.
This set of mutations and frequencies can be used to simulate NGS reads at arbitrary mean coverage depth using WGSIM [@Li2009].
Finally, these NGS reads can be aligned to a reference and variants called using a standard bioinformatics pipeline.
<!-- In Figure A every node divides at every time step with probability 1, $N$ mutations drawn from a mutational signature $M$ and added to each leaf, where $N \dist Pois(\lambda)$ and $M$ is a combination of the COSMIC mutation signatures. -->
<!-- This structure represents the *Big Bang* model of clonal evolution. -->
<!-- In Figure B the mutation signature of a single node is modified at time step $T_3$, while the number of mutations and the probability of division at each time step remain the same. -->
<!-- This construction facilitates testing the ability of our algorithm to detect changes in mutation signature across the evolution of the tumor. -->
<!-- In Figure C two lineages gain the same reproductive advantage at time step $T_3$, and form a final population with two major clones. -->
<!-- In Figure D a single lineage acquires a mutational process at time step $T_3$ that gives it a large reproductive advantage and its progeny sweep to fixation. -->
<!-- This branching process, where multiple parameters can be changed at any time step in specific nodes, provides a flexible method of simulating sets of mutations arising from complex evolutionary trajectories. -->

This simulation scheme is designed to test the effect of sequencing depth and breadth on the sensitivity and specificity of the methods developed here.
However, it will be useful in variant calling and mutation signature identification methods development generally.
It is important to note that these simulations will not be proper population genetics simulations.
The changes in reproductive rate do not correspond to changes in *selection*, and because the population is allowed to expand infinitely there is no interference between clones.
The method is narrowly focused on generating realistic NGS read data representing the outcome of interesting evolutionary scenarios, in order to determine the effect of sequencing depth and breadth on the methods devised here.

## Validation

Sensitivity and specificity as well as the effects of sequencing depth for both projects will be assessed using the simulation scheme described above.
Publishing in this field requires real data comparisons in addition to simulations [@Alioto2015], and I have identified four data sets that provide high quality sequence data with curated variant call sets.

1. The dataset published by @Griffith2015 is an acute myeloid leukemia (AML) sequenced pre-treatment as well as post-relapse to greater than 300X depth.
They also called variants on these tumors with 8 different variant callers, and curated a set of 200,000 mutations by ultra-deep sequencing.
This dataset provides very deep sequenced whole genomes, with a set of gold standard variant calls, to test the variant calling algorithm.
2. @Alioto2015 sequenced a chronic lymphocytic leukemia whole genome to approximately 300X depth, with the specific intention of generating a gold standard set of variant calls to use in testing variant calling pipelines.
As with the AML dataset, this will allow for testing of the variant calling algorithm against a gold standard variant call set, and comparing the performance to other variant callers.
3. @Chen2013 acguired 42X whole genome sequence for 16 rhabdomyosarcoma tumors from 13 patients, and resequenced 22,123 variants for validation.
This dataset is important because rhabdomyosarcomas are very rare tumors that occur in muscle tissue and are frequently diagnosed in children.
Thus these tumors have very distinct mutational signatures that are quite different from other tumor types, and provide an excellent opportunity to test the variant calling algorithm on tumors with atypical mutational signatures.
4. The dataset published by @DeBruin2014 consists of 25 spatially separated regions from seven non-small cell lung cancers sequenced to 107X.
They validated almost 2000 mutations via ultra-deep sequencing, with the goal of investigating heterogeneity of mutational processes in these tumors.
The dataset will be useful for validating the variant calling algorithm, as well as comparing the mutational process results against the more sophisticated statistical methods described above.
This dataset from the Swanton lab at University College London has been the basis for a large number of publications and has been analyzed in a huge variety of contexts, making it perfect for validating new methods.

All four sets are publicly available, although the sequence and variant data are restricted.
We have begun the process of gaining access to the datasets.

## Timeline

- Variant caller (7 months beginning March 2018)
    - Algorithm implementation (1 month)
    - Algorithm evaluation and data analysis (4 months)
    - Writing (2 months)
- Mutation simulator (6 months beginning April 2018)
    - Algorithm implementation (2 months)
    - Algorithm evaluation and data generation (2 months)
    - Writing (2 months)
- Mutation spectrum evolution (8 months beginning October 2018)
    - Algorithm implementation (2 months)
    - Algorithm evaluation and data analysis (4 months)
    - Writing (2 months)

Development of the simulator coincides with the real data phase of the variant caller project and testing occurs while generating simulations to validate the variant caller under different scenarios.
I expect the variant caller and simulator to be published together, although a small application note for the simulator might arise.
During the writing phase of the first two projects I will begin developing the mutation spectrum evolution project.

#References
<!-- 1. Simultaneous development and characterization of the variant calling algorithm and simulator - 4 months. -->
<!-- 2. Development and characterization of the method for identifying evolutionary trajectories of mutation signatures - 4 months. -->

<!-- Determining whether my variant calling method reduces the false positive rate at high depths and low will require substantial simulation work. -->
<!-- In addition we will want simulations for both methods that allow us to determine the balances associated with choosing whole genome or whole exome sequencing in terms of the strength of the mutation signature signals generated by both. -->
<!-- Given a set of true mutations, there are sophisticated methods available for generating NGS sequencing reads (including sequencing errors) for whole genome or exome that include these mutations at user-specified allele frequencies[@Stephens2016]. -->
<!-- Current methods for simulating tumor/normal genome pairs select mutations to simulate by two general methods. -->
<!-- One method is to simply select a random base in the genome, mutate it to a random alternate base, and select an allele frequency for that mutation from $U(0,1)$. -->
<!-- This method does not tend to generate very realistic tumors so an alternate method was devised. -->
<!-- Large catalogs of mutations that have been found in various tumor types are available, and it is simple to sample at random from those mutations to generate a new set of mutations to synthesize reads for. -->
<!-- By using only mutations found in a specific type of tumor, for instance melanoma, this method can be used to generate synthetic tumors with **typical** (average) mutation spectra and allele frequency distributions. -->
<!-- Neither of these methods is suitable for the purposes here because both methods I want to test rely on the idea that the tumor is "generated" by some specific and unique combination of mutational processes. -->
<!-- Thus, I will need to devise a simulation scheme that selects mutations as if they arose as sequential draws from some categorical distribution. -->
<!-- I would also like to be able to simulate these genomes under different models of tumor evolution. -->
<!-- I will develop an R package that can take as input a weight matrix describing the linear combination of mutation signatures the user would like to simulate, along with an evolutionary model, and generates a list of mutations with contexts and allele frequencies generated by that model. -->

### 2019-04-22 workout_probabilities.Rmd

---
title: "workout_probabilities"
author: "Brian Mannakee"
date: "4/22/2019"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

# Intro

Variant callers are statistical classifiers.
Usefulness of any classifier is a combination of the ranking performance of the classifier, and the quality of decisions based on the classifier in the real world.
Here we describe an empirical Bayes method which uses very high confidence variant calls from MuTect to create a biologically inspired prior probability of mutation based in the mutation spectrum of the tumor.
We use realistic tumor simulations to show that this method is superior to MuTect as a classifier based on AUROC, while equal or inferior to MuTect based on AUPRC except in situations where the distribution of allele frequencies is extreme and unrealistic.
We then implement a method of false discovery rate control that recovers the superiority of the classifier in the decision context.

# Emprical Bayes estimate of the posterior probability of mutation

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
$$
\begin{array}{l}
\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
\end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a hypothesis is $\textrm{f}_{\nu,m}(b_i,q_i)$ is defined as

$$
  \textrm{f}_{\nu,m}(b_i,q_i) = \left\{
    \begin{array}{cr}
      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
      \frac{10^{-q_i/10}}{3} & otherwise.
    \end{array}
    \right.
$$

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $log(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
By fixing the threshold posterior odds at two, and the prior probability of mutation a constant $\mu = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3.
Here we examine the effect of the assumption of a constant prior probability of mutation.

### 2019-04-24 workout_probabilities.Rmd
god this is tough to write!

@@
 
-# Emprical Bayes estimate of the posterior probability of mutation
+# Somatic variant calling base statistical model
 
@@
 the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
-For a given read with base $b_i$ and q-score $q_i$, the density function under a hypothesis is $\textrm{f}_{\nu,m}(b_i,q_i)$ is defined as
+For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as
 
@@
 MuTect reports the log likelihood ratio $log(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
-By fixing the threshold posterior odds at two, and the prior probability of mutation a constant $\mu = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3.
+By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
 Here we examine the effect of the assumption of a constant prior probability of mutation.
+
+# Site-specific prior probability of mutation
+
+While variant calling algorithms typically assume a constant probability of mutation at every site in the genome, work by Alexandrov and others show that the random mutation generating process actually varies from site to site in a nucleotide context specific manner.
+We develop a model of the prior probability of mutation to allele $m$ conditional on the observed genomic context $\mathrm{p}(m,M \mid C)$, and demonstrate an empirical Bayes method for computing this probability from MuTect output.
+The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
+$$
+\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
+$$
+since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
+Here $\mathrm{p}(M) = \mu$ as above, and the empirical distribution of contexts $\mathrm{p}(C)$ is the fraction of the genome made up of each context.
+We model $\mathrm{p}(C \mid M)$ as a multinomial distribution with parameter $\boldsymbol{\pi} = \{\pi_i\}, i = 1\dots96$.
+Mutations are drawn from this multinomial distribution such that $\mathrm{p}(C = i \mid M) = \pi_i$.
+The final quantity $\mathrm{p}(m \mid C)$, the probability of mutation to $m$ given a particular three letter context, is a function of $\boldsymbol{\pi}$. 
+We are left to estimate only the vector of probabilities $\boldsymbol{\pi}$.
+
+\[
+\begin{aligned}
+C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
+\boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
+\end{aligned}
+\]
+
+The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.

### 2019-04-24 workout_probabilities.Rmd
still going

@@
 The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
-MuTect reports the log likelihood ratio $log(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
+MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
 By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
@@
 The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.
+We compute an empirical bayes estimate of $\boldsymbol{\pi}$ by choosing $\mathbf{C}$ as the set of mutations assigned a TLOD by MuTect above some threshold, which we choose as 10.
+We show through extensive simulation that our estimate of $\boldsymbol{\pi}$ converges quickly its true simulated value.
+
+Returning to the model above, we can calculate the log posterior odds in favor of $\mathbf{H_1}$ as
+
+\[
+  \textrm{log}_{10} \left(\frac{(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})\mathrm{p}(m,M \mid C)}{(\mathcal{L}_{\nu=0,m}(\mathbf{x})(1-\mathrm{p}(m,M \mid C))} \right) = \textrm{TLOD} + \textrm{log prior odds},
+\]
+
+and the posterior odds ratio in favor of $\mathbf{H_1}$ as
+\[
+10^{(\textrm{TLOD} + \textrm{log prior odds})}.
+\]
+
+We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.

### 2019-04-25 workout_probabilities.Rmd
typed a little

@@
 We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
+Unfortunately, this increased classification performance comes at a cost in terms of decision-making.
+As measured by AUPRC, performance is a complex function of sequencing depth, mutation signature concentration, and the allele frequency distribution.
+In our simulations there is always a point on the PRC curve where MuTect outperforms our method.
+This is a result of the fact that for a given there is an allele frequency at which the ratio of expect true positives to expected false positives from being greater than 1 to less than 1.
+There is also an asymmetry in that around any particular threshold, the gradient of the ratio is fairly steep (good lord, this is awful).

### 2019-04-29 workout_probabilities.Rmd
starting to work on fdr

@@
 There is also an asymmetry in that around any particular threshold, the gradient of the ratio is fairly steep (good lord, this is awful).
+
+# False positive rate control.
+
+We develop a method, following Efron(2008), for controlling the false positive rate.
+Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
+
+$$
+\begin{array}{ll}
+p_0 = \textrm{P}\{\textrm{null}\} \quad & \textrm{f}_{0}(\mathbf{x}) \quad \textrm{density if null}\\
+p_1 = \textrm{P}\{\textrm{nonnull}\} \quad & \textrm{f}_{1}(\mathbf{x}) \quad \textrm{density if nonnull} .
+\end{array}
+$$
+
+In this application we can a compute an estimate of $p_0$ as the ratio of the expected number of false positive sites and the number of expected true positive sites with variant allele frequency $\nu = \hat{f}$.
+At a given site, with read depth $D$, the probability of seeing $n$ reads in error is binomial
+$$
+p_e \sim \textrm{Bin}(n,D,p=\bar{q})
+$$
+where $\bar{q}$ is the average quality score of the alternate reads.
+(This is an estimate, they are have different q-scores)
+The expected number of false positives with probability $p_e$ is then $N*p_e$, where $N$ is the number of sites sequenced.
+In order to compute the expected nubmer of true positives with a given vaf, we need to make an assumption about the allele frequency distribution. 
+In a neutrally evolving tumor, the count of variants with a given allele frequency is(bozic et al)
+$$
+N(f) = \frac{N\mu}{f}.
+$$

### 2019-05-17 workout_probabilities.Rmd

--- 2019-04-25 workout_probabilities.Rmd
@@
 There is also an asymmetry in that around any particular threshold, the gradient of the ratio is fairly steep (good lord, this is awful).
+
+# False positive rate control.
+
+We develop a method, following Efron(2008), for controlling the false positive rate.
+Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
+
+$$
+\begin{array}{ll}
+p_0 = \textrm{P}\{\textrm{null}\} \quad & \textrm{f}_{0}(\mathbf{x}) \quad \textrm{density if null}\\
+p_1 = \textrm{P}\{\textrm{nonnull}\} \quad & \textrm{f}_{1}(\mathbf{x}) \quad \textrm{density if nonnull} .
+\end{array}
+$$
+
+The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
+In a neutrally evolving tumor, the count of variants with a given allele frequency is(bozic et al)
+$$
+N(f) = \frac{N\mu}{f},
+$$
+Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
+This results in the estimated local nonnull probability
+$$
+\hat{p}_1 = \frac{N(f)}{N} = \frac{\mu}{f}
+$$
+and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
+the rest just copies Efron. Time to implement and try to find out how it performs.

### 2019-06-28 NAR_DRAFT.tex
add latex stuff

\documentclass[a4,center,fleqn]{NAR}

% Enter dates of publication
\copyrightyear{2008}
\pubdate{31 July 2009}
\pubyear{2009}
\jvolume{37}
\jissue{12}

\articlesubtype{Genomics and Bioinformatics}

\begin{document}

\title{Article title}

\author{%
Brian K. Mannakee\,$^{1,*}$,
Ryan N. Gutenkunst\,$^{2}$
and Second Co-Author\,$^2$%
\footnote{To whom correspondence should be addressed.
Email: rgutenk@email.arizona.edu}}

\address{%
$^{1}$University of Arizona Mel and Enid Zuckerman College of Public Health
and
$^{2}$University of Arizona Department of Molecular and Cellular Biology}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

\history{%
Received January 1, 2009;
Revised February 1, 2009;
Accepted March 1, 2009}

\maketitle

\begin{abstract}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text.
\end{abstract}

\section{Introduction}

Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency [@Williams2016;@Bozic2016;@Williams2018].
Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease [@Ding2012;@Mardis2012;@Chen2013;@Borad2014;@Findlay2016].
Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this problem increases as sequencing depth increases.

Methods for identifying true somatic mutations - i.e. variant calling -  from NGS data are an active area of research in bioinformatics.
The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity [@Koboldt2012;@Cibulskis2013].
Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers [@Garrison2012],
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\cite{@Larson2012,Roth2012a,@Christoforides2013,@Jones2016,@Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)[@Saunders2012;@Wilm2012;Shiraishi2013b;Gerstung2012;@Carrot-Zhang2017;@Fan2016], and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
All of these methods have varying levels of complexity, and some are focused on specific types of data.
The one thing they all have in common is that they either implicitly or explicitly assume that the probability of a mutation occuring at a give site is proportional to the overall mutation rate, and the same at every site in the genome.

Single nucleotide substitions, i.e. simple mutations, arise in tumors at a rate and at genomic locations driven by two main processes. 
The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context [@Nik-Zainal2012a;@Alexandrov2015;@Lee-Six2018]. 
The second, and far more complex, process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors. 
Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor [@Alexandrov2013a;@Helleday2014a;@Nik-Zainal2016;@Kandoth2013;@Alexandrov2016].

Here we present an empirical bayes method for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
We then extend the method with an application of the local false discovery rate by computing the probability that a site is non-null under an assumption of clonal expansion with either early or small selective differences between clones.
We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.

\section{MATERIALS AND METHODS}

\subsection{Materials subsection one}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.

\subsubsection{Materials subsubsection one.}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text:
\begin{align}
\mathrm{LD}^r = \frac{\mathrm{LD}}{A_\mathrm{iso}}
= 1.5 S \left( 3 \cos^2 \alpha_i - 1 \right)
\end{align}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.

\subsection{Materials subsection two}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text
(see Figure \ref{NAR-fig1}).

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text.
\begin{equation*}
\mathrm{LD} \left( t \right) =
\sum\limits_i
a_i \exp \left( \frac{-t}{\tau_i} \right)
\end{equation*}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text.

\begin{figure}[t]
\begin{center}
\includegraphics{NAR-fig1.eps}
\end{center}
\caption{Caption for figure within column.}
\label{NAR-fig1}
\end{figure}

\section{RESULTS}

\subsection{Results subsection one}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text.

\begin{table}[b]
\tableparts{%
\caption{This is a table caption}
\label{table:01}%
}{%
\begin{tabular*}{\columnwidth}{@{}lllll@{}}
\toprule
Col. head 1 & Col. head 2 & Col. head 3 & Col. head 4 & Col. head 5
\\
& (\%) & (s$^{-1}$) & (\%) & (s$^{-1}$)
\\
\colrule
Row 1 & Row 1 & Row 1 & -- & --
\\
Row 2 & Row 2 & Row 2 & Row 2 & Row 2
\\
\botrule
\end{tabular*}%
}
{This is a table footnote}
\end{table}

\subsection{Results subsection two}

Text.  Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text (see Table \ref{table:01}).

Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text (see Figure \ref{NAR-fig2}a).

Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.

\begin{figure*}[t]
\begin{center}
\includegraphics{NAR-fig2.eps}
\end{center}
\caption{Caption for wide figure over two columns.
\textbf{(a)} Left figure.
\textbf{(b)} Right figure (see (a)).
}
\label{NAR-fig2}
\end{figure*}

\subsection{Results subsection three}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text.

\section{DISCUSSION}

\subsection{Discussion subsection one}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text.

\subsection{Discussion subsection two}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text.

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.

\subsection{Discussion subsection three}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text.

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text.

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text.

\section{CONCLUSION}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text.

\section{ACKNOWLEDGEMENTS}

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text.

\subsubsection{Conflict of interest statement.} None declared.
\newpage

\bibliographystyle{NAR-natbib}
\bibliography{caller_paper}

%\end{bibliography}

\end{document}

### 2019-06-28 workout_probabilities.Rmd
add latex stuff

@@
 ---
-title: "workout_probabilities"
+title: "Probability Model"
 author: "Brian Mannakee"
-date: "4/22/2019"
+date: "6/25/2019"
 output:
-  pdf_document: default
-  word_document: default
-  html_document: default
+    pdf_document:
+        fig_caption: yes
+
+bibliography: caller_paper.bib
 ---
 
-# Intro
+# Introduction
 
@@
 Usefulness of any classifier is a combination of the ranking performance of the classifier, and the quality of decisions based on the classifier in the real world.
-Here we describe an empirical Bayes method which uses very high confidence variant calls from MuTect to create a biologically inspired prior probability of mutation based in the mutation spectrum of the tumor.
-We use realistic tumor simulations to show that this method is superior to MuTect as a classifier based on AUROC, while equal or inferior to MuTect based on AUPRC except in situations where the distribution of allele frequencies is extreme and unrealistic.
-We then implement a method of false discovery rate control that recovers the superiority of the classifier in the decision context.
+Here we describe an empirical Bayes method which uses very high confidence variant calls from MuTect to create a biologically inspired prior probability of mutation based on the mutation spectrum of the tumor.
+We use realistic tumor simulations to show that this method is superior to MuTect as a classifier based on AUROC, while equal or inferior to MuTect based on AUPRC.
+The model above is poorly calibrated, so we use estimates of the mutation rate and the distribution of allele frequencies under the null hypothesis.
+These estimates rely only the well-supported assumption that after initiation the tumor is evolving essentially neutrally.
+We show that the model is well-calibrated with this full estimate of the posterior distribution.
 
@@
 We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
-Unfortunately, this increased classification performance comes at a cost in terms of decision-making.
-As measured by AUPRC, performance is a complex function of sequencing depth, mutation signature concentration, and the allele frequency distribution.
-In our simulations there is always a point on the PRC curve where MuTect outperforms our method.
-This is a result of the fact that for a given there is an allele frequency at which the ratio of expect true positives to expected false positives from being greater than 1 to less than 1.
-There is also an asymmetry in that around any particular threshold, the gradient of the ratio is fairly steep (good lord, this is awful).
+Unfortunately, this increased classification performance comes at a cost in terms of calibration, The probabilities from this model are substantially worse than those from Mutect, such that threshold selection is essentially impossible and precision/recall is substantially worse for this model.
+There are several distributional assumptions in this model that effect model calibration, and we address all of them below.
 
-# False positive rate control.
+# Estimating the mutation rate from the data
 
-We develop a method, following Efron(2008), for controlling the false positive rate.
+As discussed above, MuTect fixes the site-specific mutation probability at $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$.
+All variant callers we are aware of either fix this parameter $\mu$ or allow the user to input the value, but there is no way to really know this value until the variant allele frequencies have been observed.
+As with estimation of the mutation signature, we can use high confidence mutations and a model of tumor evolution to compute the tumor-specific mutation rate.
+@Bozic2016 show that for any variant allele frequency $\alpha$, the total number of mutations with frequency greater than $\alpha$ and less than $0.25$ is
+
+$$
+N(\alpha) = N\mu \left( \frac{1}{\alpha} - \frac{1}{0.25} \right)
+$$
+
+Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
+By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\mu$ and recompute the odds in favor of $\mathbf{H_1}$.
+This improves the calibration of the model, but it is still worse than MuTect.
+The final assumption embedded in the model is that all variants, regardless of allele frequency, have the same prior probability of being true variants.
+Below we address this assumption.
+
+# False positive rate control
+
+We develop a method, following @Efron2008, for controlling the false positive rate.
 Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
@@
 The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
-In a neutrally evolving tumor, the count of variants with a given allele frequency is(bozic et al)
+In a neutrally evolving tumor, the number of cells is growing exponentially, and the count of variants with an allele frequency greater than a given allele frequency $f$ is [@Bozic2016,@Williams2016a]
 $$
@@
 Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
-This results in the estimated local nonnull probability
+The estimated fraction of all of the sites in the genome that will have a mutation with frequency $f$ is
 $$
-\hat{p}_1 = \frac{N(f)}{N} = \frac{\mu}{f}
+\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .1f} -  \frac{\mu}{f + .1f}
 $$
 and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
-the rest just copies Efron. Time to implement and try to find out how it performs.
+@Williams2016a provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
+
+# Results
+
+Figure 1 shows that the calibration is better than MuTect or the model without false positive rate control.
+In addition estimation of mu from the data results in better calibration than assuming a mis-specified rate.
+
+# Figures
+
+![Model performance 500X whole exome for 3 classes of mutation spectrum](figures/wes_left_plot.pdf)
+
+![Allele frequency profile 500X whole exome for 3 classes of mutation spectrum. Frequencies at a threshold p(H1) = 0.8](figures/wes_right_plot.pdf)
+
+![Model performance 100X whole genome for 3 classes of mutation spectrum](figures/wgs_left_plot.pdf)
+
+![Allele frequency profile 100X whole genome for 3 classes of mutation spectrum. Frequencies at a threshold p(H1) = 0.8](figures/wgs_right_plot.pdf)
+
+![Convergence of empirical prior](figures/kl_figure.png)
+
+![Referene input signatures](figures/input_signatures.png)
+
+# References

### 2019-07-01 NAR_DRAFT.tex
add changes

@@
 \documentclass[a4,center,fleqn]{NAR}
+
+\usepackage{NAR-natbib}
+\bibliographystyle{unsrtnat}
 
@@
 \author{%
-Brian K. Mannakee\,$^{1,*}$,
-Ryan N. Gutenkunst\,$^{2}$
-and Second Co-Author\,$^2$%
+Brian K. Mannakee\,$^{1}$ and
+Ryan N. Gutenkunst\,$^{2}$%
 \footnote{To whom correspondence should be addressed.
@@
 Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
 Text. Text.
@@
 Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
-Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency [@Williams2016;@Bozic2016;@Williams2018].
-Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease [@Ding2012;@Mardis2012;@Chen2013;@Borad2014;@Findlay2016].
+Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018}.
+Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \cite{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
 Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this problem increases as sequencing depth increases.
@@
 Methods for identifying true somatic mutations - i.e. variant calling -  from NGS data are an active area of research in bioinformatics.
-The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity [@Koboldt2012;@Cibulskis2013].
-Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers [@Garrison2012],
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\cite{@Larson2012,Roth2012a,@Christoforides2013,@Jones2016,@Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)[@Saunders2012;@Wilm2012;Shiraishi2013b;Gerstung2012;@Carrot-Zhang2017;@Fan2016], and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
+The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity \cite{Koboldt2012,Cibulskis2013}.
+Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers \cite{Garrison2012},
+joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)\cite{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
 All of these methods have varying levels of complexity, and some are focused on specific types of data.
@@
 Single nucleotide substitions, i.e. simple mutations, arise in tumors at a rate and at genomic locations driven by two main processes. 
-The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context [@Nik-Zainal2012a;@Alexandrov2015;@Lee-Six2018].
+The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
 The second, and far more complex, process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
-Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor [@Alexandrov2013a;@Helleday2014a;@Nik-Zainal2016;@Kandoth2013;@Alexandrov2016].
+Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
 
-Here we present an empirical bayes method for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
+Here we present an algorithm f estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
 We then extend the method with an application of the local false discovery rate by computing the probability that a site is non-null under an assumption of clonal expansion with either early or small selective differences between clones.
@@
 
-\bibliographystyle{NAR-natbib}
 \bibliography{caller_paper}
 
-%\end{bibliography}
-
 \end{document}

### 2019-07-01 NAR_DRAFT.tex
citep

@@
 \usepackage{NAR-natbib}
-\bibliographystyle{unsrtnat}
+\bibliographystyle{plainnat}
 
@@
 Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
-Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018}.
-Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \cite{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
+Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \citep{Williams2016,Bozic2016,Williams2018}.
+Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
 Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this problem increases as sequencing depth increases.
@@
 The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity \cite{Koboldt2012,Cibulskis2013}.
-Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers \cite{Garrison2012},
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)\cite{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
+Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers \citep{Garrison2012},
+joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\citep{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)\citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
 All of these methods have varying levels of complexity, and some are focused on specific types of data.
-The one thing they all have in common is that they either implicitly or explicitly assume that the probability of a mutation occuring at a give site is proportional to the overall mutation rate, and the same at every site in the genome.
+The one thing they all have in common is that they either implicitly or explicitly assume that the probability of a mutation occuring at a particular site is proportional to the overall mutation rate, and the same at every site in the genome.
 
 Single nucleotide substitions, i.e. simple mutations, arise in tumors at a rate and at genomic locations driven by two main processes. 
-The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
 The second, and far more complex, process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
-Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
+Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
 

### 2019-07-08 NAR_DRAFT.tex
adding text

@@
 \usepackage{NAR-natbib}
-\bibliographystyle{plainnat}
+\bibliographystyle{unsrtnat}
 
@@
 Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
-Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \citep{Williams2016,Bozic2016,Williams2018}.
+Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018}.
 Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
@@
 Methods for identifying true somatic mutations - i.e. variant calling -  from NGS data are an active area of research in bioinformatics.
-The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity \cite{Koboldt2012,Cibulskis2013}.
+The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
 Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers \citep{Garrison2012},
@@
 
-\subsection{Materials subsection one}
+\subsection{Base probability model}
 
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
+At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
+The goal is to use $\mathbf{x}$ to select between competing hypotheses;
+\begin{equation*}
+\begin{array}{l}
+\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
+\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
+\end{array}
+\end{equation*}
 
-\subsubsection{Materials subsubsection one.}
+where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e.
+the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
+For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as
+
+\begin{equation*}
+  \textrm{f}_{\nu,m}(b_i,q_i) = \left\{
+    \begin{array}{cr}
+      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
+      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
+      \frac{10^{-q_i/10}}{3} & otherwise.
+    \end{array}
+    \right.
+\end{equation*}
+
+The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
+MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or \textrm{t\_lod\_fstar} depending on the version.
+By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
+
+\subsubsection{Estimation of the mutation signature.}
+
+While variant calling algorithms typically assume a constant probability of mutation at every site in the genome, work by Alexandrov and others show that the random mutation generating process actually varies from site to site in a nucleotide context specific manner.
+We develop a model of the prior probability of mutation to allele $m$ conditional on the observed genomic context $\mathrm{p}(m,M \mid C)$, and demonstrate an empirical Bayes method for computing this probability from MuTect output.
+The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
+$$
+\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
+$$
+since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
+Here $\mathrm{p}(M) = \mu$ as above, and the empirical distribution of contexts $\mathrm{p}(C)$ is the fraction of the genome made up of each context.
+We model $\mathrm{p}(C \mid M)$ as a multinomial distribution with parameter $\boldsymbol{\pi} = \{\pi_i\}, i = 1\dots96$.
+Mutations are drawn from this multinomial distribution such that $\mathrm{p}(C = i \mid M) = \pi_i$.
+The final quantity $\mathrm{p}(m \mid C)$, the probability of mutation to $m$ given a particular three letter context, is a function of $\boldsymbol{\pi}$. 
+We are left to estimate only the vector of probabilities $\boldsymbol{\pi}$.
+
+$$
+\begin{aligned}
+C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
+\boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
+\end{aligned}
+$$
+
+The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.
+We compute an empirical bayes estimate of $\boldsymbol{\pi}$ by choosing $\mathbf{C}$ as the set of mutations assigned a TLOD by MuTect above some threshold, which we choose as 10.
+We show through extensive simulation that our estimate of $\boldsymbol{\pi}$ converges quickly its true simulated value.
+
+Returning to the model above, we can calculate the log posterior odds in favor of $\mathbf{H_1}$ as
+
+$$
+  \textrm{log}_{10} \left(\frac{(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})\mathrm{p}(m,M \mid C)}{(\mathcal{L}_{\nu=0,m}(\mathbf{x})(1-\mathrm{p}(m,M \mid C))} \right) = \textrm{TLOD} + \textrm{log prior odds},
+$$
+
+and the posterior odds ratio in favor of $\mathbf{H_1}$ as
+$$
+10^{(\textrm{TLOD} + \textrm{log prior odds})}.
+$$
+
+We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
+Unfortunately, this increased classification performance comes at a cost in terms of calibration, The probabilities from this model are substantially worse than those from Mutect, such that threshold selection is essentially impossible and precision/recall is substantially worse for this model.
+There are several distributional assumptions in this model that effect model calibration, and we address all of them below.
+
+\subsubsection{Estimation of the mutation rate.}
+
+As discussed above, MuTect fixes the site-specific mutation probability at $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$.
+All variant callers we are aware of either fix this parameter $\mu$ or allow the user to input the value, but there is no way to really know this value until the variant allele frequencies have been observed.
+As with estimation of the mutation signature, we can use high confidence mutations and a model of tumor evolution to compute the tumor-specific mutation rate.
+\citet{Bozic2016} show that for any variant allele frequency $\alpha$, the total number of mutations with frequency greater than $\alpha$ and less than $0.25$ is
+
+$$
+N(\alpha) = N\mu \left( \frac{1}{\alpha} - \frac{1}{0.25} \right)
+$$
+
+Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
+By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\mu$ and recompute the odds in favor of $\mathbf{H_1}$.
+This improves the calibration of the model, but it is still worse than MuTect.
+The final assumption embedded in the model is that all variants, regardless of allele frequency, have the same prior probability of being true variants.
+Below we address this assumption.
+
+\subsubsection{False positive rate control.}
+
+We develop a method, following \citet{Efron2008}, for controlling the false positive rate.
+Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
+
+$$
+\begin{array}{ll}
+p_0 = \textrm{P}\{\textrm{null}\} \quad & \textrm{f}_{0}(\mathbf{x}) \quad \textrm{density if null}\\
+p_1 = \textrm{P}\{\textrm{nonnull}\} \quad & \textrm{f}_{1}(\mathbf{x}) \quad \textrm{density if nonnull} .
+\end{array}
+$$
+
+The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
+In a neutrally evolving tumor, the number of cells is growing exponentially, and the count of variants with an allele frequency greater than a given allele frequency $f$ is \cite{Bozic2016,Williams2016a}.
+$$
+N(f) = \frac{N\mu}{f},
+$$
+Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
+The estimated fraction of all of the sites in the genome that will have a mutation with frequency $f$ is
+$$
+\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .1f} -  \frac{\mu}{f + .1f}
+$$
+and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
+citet{Williams2016a} provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
+
+\subsubsection{False positive rate control.}
+
+Tumor variants via a branching process to simulate neutral evolution with no death.
+Variants selected from PCAWG and TCGA lists(with download dates).
+We simulated 100X whole genome and 500X exome normal reads from the GRCH38 reference genome with VarSim \cite{Mu2015}, and aligned them to GRch38 with BWA \cite{Li2009a}, both with default parameters.
+Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
+and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:
+
+\begin{verbatim}
+java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect --reference_sequence $ref_path \
+        --dbsnp $db_snp \
+        --enable_extended_output \
+        --fraction_contamination 0.00 \
+        --tumor_f_pretest 0.00 \
+        --initial_tumor_lod -10.00 \
+        --required_maximum_alt_allele_mapping_quality_score 1 \
+        --input_file:normal $tmp_normal \
+        --input_file:tumor $tmp_tumor \
+        --out $out_path/$chr.txt \
+        --coverage_file $out_path/$chr.cov
+\end{verbatim}
+
+Variants identified by MuTect are labelled as to whether they pass all MuTect filters, pass all filters *other* than the evidence threshold \textrm{tlod\_f\_star}, or fail to pass any filter other than \textrm{tlod\_f\_star}.
+Variants that pass all filters or fail only \textrm{tlod\_f\_star} are then passed to {method} for prior estimation and rescoring.
 
@@
 \end{align}
+
 Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
@@
 \subsubsection{Conflict of interest statement.} None declared.
-\newpage
 

### 2019-07-08 NAR_DRAFT.tex
make tiny for now

@@
 We develop a method, following \citet{Efron2008}, for controlling the false positive rate.
-Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
+Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either \textit{null} (non-variant with $\nu = 0$) or \textit{nonnull} (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
 
@@
 and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
-citet{Williams2016a} provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
+\citet{Williams2016a} provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
 
-\subsubsection{False positive rate control.}
+\subsubsection{Tumor simulations.}
 
@@
 
+\begin{tiny}
 \begin{verbatim}
+
   java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect --reference_sequence $ref_path \
@@
         --coverage_file $out_path/$chr.cov
+
 \end{verbatim}
-
+\end{tiny}
 Variants identified by MuTect are labelled as to whether they pass all MuTect filters, pass all filters *other* than the evidence threshold \textrm{tlod\_f\_star}, or fail to pass any filter other than \textrm{tlod\_f\_star}.

### 2019-07-08 NAR_DRAFT.tex
check in

@@
 
-\subsubsection{Estimation of the mutation signature.}
+\subsection{Estimation of the mutation signature.}
 
@@
 
-\subsubsection{Estimation of the mutation rate.}
+\subsection{Estimation of the mutation rate.}
 
@@
 
-\subsubsection{False positive rate control.}
+\subsection{False positive rate control.}
 
@@
 $$
-\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .1f} -  \frac{\mu}{f + .1f}
+\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .01} -  \frac{\mu}{f + .01}
 $$
@@
 
-\subsubsection{Tumor simulations.}
+\subsection{Tumor simulations.}
 
-Tumor variants via a branching process to simulate neutral evolution with no death.
-Variants selected from PCAWG and TCGA lists(with download dates).
-We simulated 100X whole genome and 500X exome normal reads from the GRCH38 reference genome with VarSim \cite{Mu2015}, and aligned them to GRch38 with BWA \cite{Li2009a}, both with default parameters.
+We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
+Variants were selected from TCGA and PCAWG variant files(dates).
+Whole genome (100X depth), and whole exome (500X depth) reads from the GRCH38 reference genome with VarSim \cite{Mu2015}, and aligned them to GRch38 with BWA \cite{Li2009a}, both with default parameters.
 Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
@@
 Variants that pass all filters or fail only \textrm{tlod\_f\_star} are then passed to {method} for prior estimation and rescoring.
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text:
-\begin{align}
-\mathrm{LD}^r = \frac{\mathrm{LD}}{A_\mathrm{iso}}
-= 1.5 S \left( 3 \cos^2 \alpha_i - 1 \right)
-\end{align}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-
-\subsection{Materials subsection two}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text
-(see Figure \ref{NAR-fig1}).
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text.
-\begin{equation*}
-\mathrm{LD} \left( t \right) =
-\sum\limits_i
-a_i \exp \left( \frac{-t}{\tau_i} \right)
-\end{equation*}
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text.
 

### 2019-07-09 NAR_DRAFT.tex
check in

@@
 \bibliographystyle{unsrtnat}
-
+\usepackage{color}
+\newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
+\newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
 % Enter dates of publication
@@
 By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\mu$ and recompute the odds in favor of $\mathbf{H_1}$.
-This improves the calibration of the model, but it is still worse than MuTect.
-The final assumption embedded in the model is that all variants, regardless of allele frequency, have the same prior probability of being true variants.
-Below we address this assumption.
 
-\subsection{False positive rate control.}
+%`\subsection{False positive rate control.}
 
-We develop a method, following \citet{Efron2008}, for controlling the false positive rate.
-Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either \textit{null} (non-variant with $\nu = 0$) or \textit{nonnull} (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
+%We develop a method, following \citet{Efron2008}, for controlling the false positive rate.
+%Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either \textit{null} (non-variant with $\nu = 0$) or \textit{nonnull} (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
 
-$$
-\begin{array}{ll}
-p_0 = \textrm{P}\{\textrm{null}\} \quad & \textrm{f}_{0}(\mathbf{x}) \quad \textrm{density if null}\\
-p_1 = \textrm{P}\{\textrm{nonnull}\} \quad & \textrm{f}_{1}(\mathbf{x}) \quad \textrm{density if nonnull} .
-\end{array}
-$$
+%$$
+%\begin{array}{ll}
+%p_0 = \textrm{P}\{\textrm{null}\} \quad & \textrm{f}_{0}(\mathbf{x}) \quad \textrm{density if null}\\
+%p_1 = \textrm{P}\{\textrm{nonnull}\} \quad & \textrm{f}_{1}(\mathbf{x}) \quad \textrm{density if nonnull} .
+%\end{array}
+%$$
 
-The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
-In a neutrally evolving tumor, the number of cells is growing exponentially, and the count of variants with an allele frequency greater than a given allele frequency $f$ is \cite{Bozic2016,Williams2016a}.
-$$
-N(f) = \frac{N\mu}{f},
-$$
-Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
-The estimated fraction of all of the sites in the genome that will have a mutation with frequency $f$ is
-$$
-\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .01} -  \frac{\mu}{f + .01}
-$$
-and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
-\citet{Williams2016a} provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
+%The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
+%In a neutrally evolving tumor, the number of cells is growing exponentially, and the count of variants with an allele frequency greater than a given allele frequency $f$ is \cite{Bozic2016,Williams2016a}.
+%$$
+%N(f) = \frac{N\mu}{f},
+%$$
+%Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
+%The estimated fraction of all of the sites in the genome that will have a mutation with frequency $f$ is
+%$$
+%\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .01} -  \frac{\mu}{f + .01}
+%$$
+
+%\bkmcomment{This is wrong. This is the cdf. I need the integral of the derivative which is the density.}
+%and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
+%\citet{Williams2016a} provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
 
@@
 \begin{center}
-\includegraphics{NAR-fig1.eps}
+\includegraphics{figures/wes_left_plot.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{NAR-fig2.eps}
+\includegraphics{figures/kl_figure.png}
 \end{center}

### 2019-07-09 NAR_DRAFT.tex
add aml31 stub

@@
 
-Here we present an algorithm f estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
+Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
 We then extend the method with an application of the local false discovery rate by computing the probability that a site is non-null under an assumption of clonal expansion with either early or small selective differences between clones.
@@
 Variants that pass all filters or fail only \textrm{tlod\_f\_star} are then passed to {method} for prior estimation and rescoring.
+
+\subsection{Real tumor data.}
+\subsubsection{Acute Myeloid Leukemia}
+We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date}.
+We merged the gold and platinum lists, and define is \textit{not present} any variant for which deep sequencing was performed and zero alternate reads were observed.
+This is a very conservative metric.
 

### 2019-07-09 NAR_DRAFT.tex
add aml31 citation

@@
 \subsubsection{Acute Myeloid Leukemia}
-We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date}.
+We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date} \cite{Griffith2015}.
 We merged the gold and platinum lists, and define is \textit{not present} any variant for which deep sequencing was performed and zero alternate reads were observed.
 This is a very conservative metric.
+\subsubsection{What I call the cell paper} Not sure if this will get used.
 

### 2019-07-10 NAR_DRAFT.tex
check in figure 1
add figuref 1 caption

@@
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics{figures/wes_left_plot.pdf}
-\end{center}
-\caption{Caption for figure within column.}
-\label{NAR-fig1}
-\end{figure}
+%\begin{figure}[t]
+%\begin{center}
+%\includegraphics{figures/wes_left_plot.pdf}
+%\end{center}
+%\caption{Caption for figure within column.}
+%\label{NAR-fig1}
+%\end{figure}
 
@@
 \begin{center}
-\includegraphics{figures/kl_figure.png}
+\includegraphics{figures/fig1.pdf}
 \end{center}
-\caption{Caption for wide figure over two columns.
-\textbf{(a)} Left figure.
-\textbf{(b)} Right figure (see (a)).
+\caption{Model performance on 500X simulated whole exome.
+\textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
+\textbf{(b)} Convergence of the estimated mutation signature to the simulated signature with mutations in descending order of confidence.
+\textbf{(d)} \bkmcomment{signatures or one single calibration curve.
+This figure could also lose current c and d and have all three calibration curves to the right.
+KL and signatures in another figure?}
 }

### 2019-07-11 NAR_DRAFT.tex
check in aml figure
add minor caption
getting much closer

--- 2019-07-09 NAR_DRAFT.tex
@@
 \begin{center}
-\includegraphics{figures/wes_left_plot.pdf}
+\includegraphics{figures/aml_plot.pdf}
 \end{center}
-\caption{Caption for figure within column.}
+\caption{Performance on aml 31.}
 \label{NAR-fig1}
@@
 \begin{center}
-\includegraphics{figures/kl_figure.png}
+\includegraphics{figures/fig1.pdf}
 \end{center}
-\caption{Caption for wide figure over two columns.
-\textbf{(a)} Left figure.
-\textbf{(b)} Right figure (see (a)).
+\caption{Model performance on 500X simulated whole exome.
+\textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
+Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
+A perfectly calibrated model would generate the diagonal line.
 }

### 2019-07-12 NAR_DRAFT.tex
check in figure 2
check in figure 1
check in signature words
check in

@@
 We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date} \cite{Griffith2015}.
-We merged the gold and platinum lists, and define is \textit{not present} any variant for which deep sequencing was performed and zero alternate reads were observed.
+We merged the gold and platinum lists, and define as \textit{not present} any variant for which deep sequencing was performed and zero alternate reads were observed.
 This is a very conservative metric.
@@
 \begin{center}
-\includegraphics{figures/aml_plot.pdf}
+\includegraphics{figures/fig1.pdf}
 \end{center}
-\caption{Performance on aml 31.}
+\caption{Signature words.}
 \label{NAR-fig1}
@@
 \begin{center}
-\includegraphics{figures/fig1.pdf}
+\includegraphics{figures/fig2.pdf}
 \end{center}

### 2019-07-16 NAR_DRAFT.tex
text

@@
 
-\subsection{Results subsection one}
+\subsection{Convergence of the mutation prior to the true mutation profile}
 
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text.
+The dirichlet prior probability of mutation converges quickly to the true mutation signature.
+The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-fig1}\textbf{e,f}).
+\subsection{Performance on real data}
+
+\begin{itemize}
+  \item Similar performance in aml31
+  \item Similar calibration
+  \item caveats about the fact that we don't have true false positive/false negative data
+\end{itemize}
+
+\subsection{Simulated data}
+
+\subsubsection{classification}
+We simulated both whole genome and whole exome tumors with three different mutation signatures.
+Measures of classification performance are positively correlated with signature concentration.
+For all simulations the calibrated model produces a better variant classifier than MuTect (Figure \ref{NAR-fig2} \textbf{a} and \textbf{c}).
+
+\subsubsection{calibration}
+We computed the Integrated Calibration Index \cite{Austin2019} for both simulated (Figure \ref{NAR-fig2} \textbf{b,d}) and real tumors \bkmcomment{aml31 when the plot is ready.}.
+In all cases we find that the Calibrated model is better calibrated than MuTect.\bkmcomment{not perfect though}
 
@@
 
-\subsection{Results subsection two}
-
-Text.  Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text (see Table \ref{table:01}).
-
-Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text (see Figure \ref{NAR-fig2}a).
-
-Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-
 \begin{figure*}[t]
@@
 \end{figure*}
-
-\subsection{Results subsection three}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text.
 

### 2019-07-16 NAR_DRAFT.tex
calibration and mutation rate table

@@
 
+\subsection{Calibration}
+We use the Integrated Calibration Index to quantify the difference between MuTect and the Calibrated model \cite{Austin2019}.
+Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-fig1} \textbf{e}).
+Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the total area between the result and a diagonal line.
+\bkmcomment{Honestly, just putting the R code here exactly as Austin does is probably the best explanation?}
+
 \begin{figure}[t]
@@
 
-\begin{table}[b]
+\begin{table*}[b]
 \tableparts{%
-\caption{This is a table caption}
+\caption{This is a table with auroc, auprc, ICI, and maybe fraction called?}
 \label{table:01}%
 }{%
-\begin{tabular*}{\columnwidth}{@{}lllll@{}}
+\begin{tabular*}{\columnwidth}{@{}llllllll@{}}
 \toprule
-Col. head 1 & Col. head 2 & Col. head 3 & Col. head 4 & Col. head 5
+Experiment & $\mu$ & AUROC &  & AUPRC & & ICI & 
 \\
-& (\%) & (s$^{-1}$) & (\%) & (s$^{-1}$)
+& & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
 \\
 \colrule
-Row 1 & Row 1 & Row 1 & -- & --
+Row 1 & Row 1 & Row 1 & -- & -- &&&
 \\
-Row 2 & Row 2 & Row 2 & Row 2 & Row 2
+Row 2 & Row 2 & Row 2 & Row 2 & Row 2 &&&
 \\
@@
 }
-{This is a table footnote}
-\end{table}
+{AUROC/AUPRC = area under ROC/PRC curve. ICI = Integrated calibration index}
+\end{table*}
 

### 2019-07-16 NAR_DRAFT.tex
working on table
small table changes

@@
 \subsubsection{Acute Myeloid Leukemia}
-We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date} \cite{Griffith2015}.
+We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date, and primary/relapse distinction} \cite{Griffith2015}.
 We merged the gold and platinum lists, and define as \textit{not present} any variant for which deep sequencing was performed and zero alternate reads were observed.
@@
 We use the Integrated Calibration Index to quantify the difference between MuTect and the Calibrated model \cite{Austin2019}.
-Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-fig1} \textbf{e}).
+Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-fig2}\textbf{b,d}).
 Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the total area between the result and a diagonal line.
-\bkmcomment{Honestly, just putting the R code here exactly as Austin does is probably the best explanation?}
 
@@
 \section{RESULTS}
+\bkmcomment{I can't find a measured mutation rate for aml31. I get 3e-8 which isn't out of line for a leukemia primary.}
 
@@
 }{%
-\begin{tabular*}{\columnwidth}{@{}llllllll@{}}
+\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
 \toprule
-Experiment & $\mu$ & AUROC &  & AUPRC & & ICI & 
+Depth & Signature & $\mu$ & AUROC &  & AUPRC & & ICI & 
 \\
-& & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
+& & & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
 \\
 \colrule
-Row 1 & Row 1 & Row 1 & -- & -- &&&
+100X WGS & Concentrated & Row 1 & Row 1 & -- & -- &&
 \\
-Row 2 & Row 2 & Row 2 & Row 2 & Row 2 &&&
+100X WGS & Intermediate & Row 1 & Row 1 & -- & -- &&
+\\
+100X WGS & Diffuse & Row 1 & Row 1 & -- & -- &&
+\\
+500X WES & Concentrated & Row 1 & Row 1 & -- & -- &&
+\\
+500X WES & Intermediate & Row 1 & Row 1 & -- & -- &&
+\\
+500X WES & Diffuse & Row 1 & Row 1 & -- & -- &&
+\\
+AML 31 & Actual & Row 1 & Row 1 & -- & -- &&
 \\
@@
 }
-{AUROC/AUPRC = area under ROC/PRC curve. ICI = Integrated calibration index}
+{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
 \end{table*}

### 2019-07-17 NAR_DRAFT.tex
make a start fixing the methods/probability model section

@@
 We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
+\newpage
+\section{MATERIALS AND METHODS}
+\subsection{Somatic variant calling base probability model}
 
-\section{MATERIALS AND METHODS}
-
-\subsection{Base probability model}
-
+\bkmcomment{I feel like a discussion of the mutation signature should go first. Referencing the figure? The issue is that I have this m,M notation but the context C rolls in the m.}
 At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
 The goal is to use $\mathbf{x}$ to select between competing hypotheses;
-\begin{equation*}
+$$
   \begin{array}{l}
@@
   \end{array}
-\end{equation*}
-
+$$
 where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e.
-the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
+the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
 For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as
 
-\begin{equation*}
-  \textrm{f}_{\nu,m}(b_i,q_i) = \left\{
+Making the common assumption that NGS reads are independent, the posterior probability of a given hypothesis is 
+$$
+  \begin{array}{rl}
+    P(m,\nu) = &p(m,\nu) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}) \\
+             = &p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
+  \end{array}
+$$
+where the density $\textrm{f}_{m,\nu}(x_i)$ is
+
+$$
+  \textrm{f}_{m,\nu}({b_i,q_i}) = \left\{
     \begin{array}{cr}
@@
     \right.
-\end{equation*}
+$$
 
-The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
-MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or \textrm{t\_lod\_fstar} depending on the version.
-By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
+Assuming that the identity of the alternate allele and the estimated allele frequency are independent of each other, and $\nu$ is uniformly distributed results in the final model
+$$
+  P(m,\nu) = p(m) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}).
+$$
 
-\subsection{Estimation of the mutation signature.}
+\subsection{Site-specific prior probability of mutation}
 
-While variant calling algorithms typically assume a constant probability of mutation at every site in the genome, work by Alexandrov and others show that the random mutation generating process actually varies from site to site in a nucleotide context specific manner.
-We develop a model of the prior probability of mutation to allele $m$ conditional on the observed genomic context $\mathrm{p}(m,M \mid C)$, and demonstrate an empirical Bayes method for computing this probability from MuTect output.
+The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
+However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted $p(m,M | C)$(citations).
 The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
 $$
-\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
+\mathrm{p}(m,M \mid C) = \mathrm{p}(C \mid M,m) \frac{p(m,M)}{p(C)}
 $$
-since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
-Here $\mathrm{p}(M) = \mu$ as above, and the empirical distribution of contexts $\mathrm{p}(C)$ is the fraction of the genome made up of each context.
-We model $\mathrm{p}(C \mid M)$ as a multinomial distribution with parameter $\boldsymbol{\pi} = \{\pi_i\}, i = 1\dots96$.
-Mutations are drawn from this multinomial distribution such that $\mathrm{p}(C = i \mid M) = \pi_i$.
-The final quantity $\mathrm{p}(m \mid C)$, the probability of mutation to $m$ given a particular three letter context, is a function of $\boldsymbol{\pi}$. 
-We are left to estimate only the vector of probabilities $\boldsymbol{\pi}$.
-
-$$
-\begin{aligned}
-C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
-\boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
-\end{aligned}
-$$
-
-The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.
-We compute an empirical bayes estimate of $\boldsymbol{\pi}$ by choosing $\mathbf{C}$ as the set of mutations assigned a TLOD by MuTect above some threshold, which we choose as 10.
-We show through extensive simulation that our estimate of $\boldsymbol{\pi}$ converges quickly its true simulated value.
-
-Returning to the model above, we can calculate the log posterior odds in favor of $\mathbf{H_1}$ as
-
-$$
-  \textrm{log}_{10} \left(\frac{(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})\mathrm{p}(m,M \mid C)}{(\mathcal{L}_{\nu=0,m}(\mathbf{x})(1-\mathrm{p}(m,M \mid C))} \right) = \textrm{TLOD} + \textrm{log prior odds},
-$$
-
-and the posterior odds ratio in favor of $\mathbf{H_1}$ as
-$$
-10^{(\textrm{TLOD} + \textrm{log prior odds})}.
-$$
-
-We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
-Unfortunately, this increased classification performance comes at a cost in terms of calibration, The probabilities from this model are substantially worse than those from Mutect, such that threshold selection is essentially impossible and precision/recall is substantially worse for this model.
-There are several distributional assumptions in this model that effect model calibration, and we address all of them below.
+\bkmcomment{This is a problem. I am having trouble with the fact that C includes m. Maybe $p(C | M)$}
 
@@
 \\
-& & & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
+& & (estimated) & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
 \\
@@
 \\
-AML 31 & Actual & Row 1 & Row 1 & -- & -- &&
+AML 31 & Actual & $3.6e-8$ & Row 1 & -- & -- &&
 \\

### 2019-07-17 workout_probabilities.Rmd
make a start fixing the methods/probability model section

@@
 
-# Somatic variant calling base statistical model
+# Somatic variant calling base probability model
 
@@
 where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e.
-the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
+the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
 For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as
 
+Making the common assumption that NGS reads are independent, the posterior probability of a given hypothesis, $P(m,\nu)$ is 
 $$
-  \textrm{f}_{\nu,m}(b_i,q_i) = \left\{
+  \begin{array}{cr}
+    P(m,\nu) = &p(m,\nu) \cdot $\mathcal{L}_{m,\nu}(\mathbf{x}). \\
+             = &p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
+  \end{array}
+$$
+where the density \textrm{f}_{m,\nu}(x_i) is
+$$
+  \textrm{f}_{m,\nu}({b_i,q_i}) = \left\{
     \begin{array}{cr}
@@
 
-The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
-MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
-By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
-Here we examine the effect of the assumption of a constant prior probability of mutation.
+Assuming that the identity of the alternate allele and the estimated allele frequency are independent of each other, and $\nu$ is uniformly distributed results in the final model
+$$
+  P(m,\nu) = p(m) \times $\mathcal{L}_{m,\nu}(\mathbf{x}).
+$$
 
 # Site-specific prior probability of mutation
+
+The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
+However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted $p(m,M | C)$(citations).
+The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
+$$
+\mathrm{p}(m,M \mid C) = \mathrm{p}(C \mid M,m) \frac{p(m,M)}{p(C)}
+$$
+\bkmcomment{This is a problem. I am having trouble with the fact that C includes m. Maybe $p(C | M)$}
 

### 2019-07-19 NAR_DRAFT.tex
check in new figure for signatures
check in aml plot
nicer equations
fill out wes in table

@@
 
-\bkmcomment{I feel like a discussion of the mutation signature should go first. Referencing the figure? The issue is that I have this m,M notation but the context C rolls in the m.}
 At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
-The goal is to use $\mathbf{x}$ to select between competing hypotheses;
+We want to use $\mathbf{x}$ to select between competing hypotheses;
 $$
@@
   \begin{array}{rl}
-    P(m,\nu) = &p(m,\nu) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}) \\
+    P(m,\nu) = &p(m,\nu) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}) \\ \\
              = &p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
@@
     \begin{array}{cr}
-      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
-      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
+      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\ \\
+      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\ \\
       \frac{10^{-q_i/10}}{3} & otherwise.
@@
 
-\subsection{Site-specific prior probability of mutation}
+\subsection{Estimation of the mutation signature}
+The 3-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
+There are four possible bases 3` and 5` of the site, leading to 16 possible flanking base combinations.
+Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site, and three possible alternate alleles for each.
+Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles (see Figure \ref{NAR-sigfig}).
+The mutation context $C$ is indexed by $i=\{1 \dots 16\}$ flanking base combinations, $j = \{1,2\}$ reference bases, and $j = \{1,2,3\}$ alternate alleles.
+We model $ p(C | M) $ as a multinomial distribution with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j,k}\} $ and pseudo-count hyper-parameter $\boldsymbol{\alpha}$
 
-The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
-However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted $p(m,M | C)$(citations).
-The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
 $$
-\mathrm{p}(m,M \mid C) = \mathrm{p}(C \mid M,m) \frac{p(m,M)}{p(C)}
+\begin{array}{rl}
+  C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
+  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
+\end{array}
 $$
-\bkmcomment{This is a problem. I am having trouble with the fact that C includes m. Maybe $p(C | M)$}
+We use high confidence variant calls -- odds in favor of the mutation greater than 10:1 -- to estimate $\boldsymbol{\pi}$.
 
 \subsection{Estimation of the mutation rate.}
-
-As discussed above, MuTect fixes the site-specific mutation probability at $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$.
-All variant callers we are aware of either fix this parameter $\mu$ or allow the user to input the value, but there is no way to really know this value until the variant allele frequencies have been observed.
 As with estimation of the mutation signature, we can use high confidence mutations and a model of tumor evolution to compute the tumor-specific mutation rate.
@@
 Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
-By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\mu$ and recompute the odds in favor of $\mathbf{H_1}$.
+By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\hat{\mu}$.
+In all results here we select $\alpha = 0.05$.
+
+\subsection{Site-specific prior probability of mutation}
+The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
+However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted here $p(m,M | C)$ \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
+$$
+\begin{array}{rl}
+  \mathrm{p}(m,M \mid C) = &\mathrm{p}(m \mid C) \mathrm{p}(M \mid C) \\ \\ 
+  = &\mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
+\end{array}
+$$
+since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
+The estimated mutation signature and mutation rate, allow us to compute $p(m,M | C)$ as
+
+$$
+\begin{array}{rl}
+  \mathrm{p}(m \mid C) = & \pi_{i,j,k} \\ \\
+  \mathrm{p}(C \mid M) = & \sum_{k}\pi_{i,j} \\ \\
+  \mathrm{p}(M) = & \hat{\mu},
+\end{array}
+$$
+with $\mathrm{p}(C)$ the fraction of all sites in the genome with context $C_{i,j}$ divided by 3.
 
@@
 Variants were selected from TCGA and PCAWG variant files(dates).
-Whole genome (100X depth), and whole exome (500X depth) reads from the GRCH38 reference genome with VarSim \cite{Mu2015}, and aligned them to GRch38 with BWA \cite{Li2009a}, both with default parameters.
+Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
 Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
@@
 Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-fig2}\textbf{b,d}).
-Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the total area between the result and a diagonal line.
+Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the result and a diagonal line.
 
@@
 \begin{center}
-\includegraphics{figures/fig1.pdf}
+\includegraphics{figures/signature_figure.pdf}
 \end{center}
-\caption{Signature words.}
-\label{NAR-fig1}
+\caption{Mutation signatures. 
+\textbf{(a)} The observed mutation signature of the AML31 primary tumor. 
+\textbf{(b)} A mutation signature used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+\textbf{(c)} Equal proportions of signatures 1, 4, \& 5.
+\textbf{(d)} Equal proportions of signatures 1, 3, \& 5.
+KL divergence between the estimated signature and the simulated mutation signature for \textbf{e)} Whole exomes and \textbf{f)} whole genomes.
+The estimated mutation signature converges quickly to the target as the number of mutations evaluated rises.
+}
+\label{NAR-sigfig}
 \end{figure}
@@
 \section{RESULTS}
-\bkmcomment{I can't find a measured mutation rate for aml31. I get 3e-8 which isn't out of line for a leukemia primary.}
+\begin{figure*}[t]
+  \begin{center}
+  \includegraphics{figures/aml_plot.pdf}
+  \end{center}
+  \caption{Model performance on 300X Whole leukemia genome.
+  \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
+  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
+  A perfectly calibrated model would generate the diagonal line.}
+  \label{NAR-aml}
+\end{figure*}
 
@@
 The dirichlet prior probability of mutation converges quickly to the true mutation signature.
-The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-fig1}\textbf{e,f}).
+The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-sigfig}\textbf{e,f}).
 \subsection{Performance on real data}
 
-\begin{itemize}
-  \item Similar performance in aml31
-  \item Similar calibration
-  \item caveats about the fact that we don't have true false positive/false negative data
-\end{itemize}
+The Calibrated model has better AUROC and AUPRC than the MuTect model in this real tumor (Figure \ref{NAR-aml}\textbf{a,c}, Table \ref{table:01}).
+The Calibrated model has slightly worse calibration as measure by ICI, but both models substantially underestimate the true positive probability throughout the range of possible values (Figure \ref{NAR-aml} \textbf{b,d}).
+One caveat to consider when considering these results is that while AML31 was subject to extensive variant calling validation, their list of true positive calls likely does not include every true variant at low allele frequencies, and true negative calls were not evaluated.
 
@@
 \subsubsection{calibration}
-We computed the Integrated Calibration Index \cite{Austin2019} for both simulated (Figure \ref{NAR-fig2} \textbf{b,d}) and real tumors \bkmcomment{aml31 when the plot is ready.}.
-In all cases we find that the Calibrated model is better calibrated than MuTect.\bkmcomment{not perfect though}
+The Calibrated model tends to score true negative variants lower and true positive variants higher than Mutect (Figure \ref{NAR-fig2} \textbf{b,d}).
+The Integrated calibration index for the Calibrated model is better than that of the MuTect model for all tumors simulated.
 
@@
 \tableparts{%
-\caption{This is a table with auroc, auprc, ICI, and maybe fraction called?}
+\caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
 \label{table:01}%
@@
 \colrule
-100X WGS & Concentrated & Row 1 & Row 1 & -- & -- &&
+100X whole genome & Concentrated & Row 1 & Row 1 & -- & -- & -- & -- & --
 \\
-100X WGS & Intermediate & Row 1 & Row 1 & -- & -- &&
+100X whole genome & Intermediate & Row 1 & Row 1 & -- & -- & -- & -- & --
 \\
-100X WGS & Diffuse & Row 1 & Row 1 & -- & -- &&
+100X whole genome & Diffuse & Row 1 & Row 1 & -- & -- & -- & -- & --
 \\
-500X WES & Concentrated & Row 1 & Row 1 & -- & -- &&
+500X whole exome & Concentrated & Row 1 & .848 & .928 & .674 & .757 & .0585 & .0465
 \\
-500X WES & Intermediate & Row 1 & Row 1 & -- & -- &&
+500X whole exome & Intermediate & Row 1 & .847 & .842 & .677 & .707 & .0587 & .0566 
 \\
-500X WES & Diffuse & Row 1 & Row 1 & -- & -- &&
+500X whole exome & Diffuse & Row 1 & .850 & .873 & .676 & .698 & .0565 & .0575
 \\
-AML 31 & Actual & $3.6e-8$ & Row 1 & -- & -- &&
+AML 31 & Actual & $3.6e-8$ & .935 & .949 & .995 & .996 & 0.121 & 0.155
 \\

### 2019-07-29 NAR_DRAFT.tex
some work

@@
 \usepackage{color}
+\usepackage{xfrac}
 \newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
 \newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
+\newcommand{\batcave}{\textbf{BATCAVE }}
 % Enter dates of publication
@@
 
-\title{Article title}
+\title{\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation}
 
@@
 \address{%
-$^{1}$University of Arizona Mel and Enid Zuckerman College of Public Health
+$^{1}$Mel and Enid Zuckerman College of Public Health, University of Arizona, Tucson AZ
 and
-$^{2}$University of Arizona Department of Molecular and Cellular Biology}
+$^{2}$Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ}
 % Affiliation must include:
@@
 We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
-\newpage
+\bkmcomment{batcave,batcaver,batcaveR}
+
 \section{MATERIALS AND METHODS}
@@
 the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
-For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as
+The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
+Assuming that reads are independent, we have
 
-Making the common assumption that NGS reads are independent, the posterior probability of a given hypothesis is 
+\begin{equation}
+  \label{eqn:1}
 $$
-  \begin{array}{rl}
-    P(m,\nu) = &p(m,\nu) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}) \\ \\
-             = &p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
-  \end{array}
+  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
 $$
-where the density $\textrm{f}_{m,\nu}(x_i)$ is
+\end{equation}
+where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.
+
+Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e.
+$p(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
+\begin{equation}
+  \label{eqn:2}
+$$
+  P(m,\nu) = p(m) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i).
+$$
+\end{equation}
+Here we show how \batcave can be used to provide a tumor-specific estimate of the mutation probability $p(m)$.
+
+\subsection{Site-specific prior probability of mutation}
+The probability which we have denoted $p(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
+However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $p(m,M | C)$ \bkmcomment{citation for genomic context in general} .
+Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as
+
+\begin{equation}
+  \label{eqn:full_prior}
+  $$
+  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}.
+  $$
+\end{equation}
+We next show how to estimate the quantities in equation \ref{eqn:full_prior}.
+
+\subsection{Estimation of the mutation profile}
+There are many aspects of local genomic architecture, generally described as the genomic \textit{context}, which can affect the local somatic mutation rate at a particular site in the genome.
+Here we focus on the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+The tri-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
+There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
+Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, and three possible alternate alleles {\{A,(C/T),G\}} for each.
+Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles.
+
+\begin{figure}[t]
+  \begin{center}
+  \includegraphics{figures/signature_figure.pdf}
+  \end{center}
+  \caption{Mutation profiles. 
+  \textbf{(a)} The observed mutation profile of the AML31 primary tumor. 
+  \textbf{(b)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{(c)} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{(d)} Equal proportions of signatures 1, 3, \& 5.
+  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{e)} Whole exomes and \textbf{f)} whole genomes.
+  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
+  }
+  \label{NAR-sigfig}
+  \end{figure}
+
+Figure \ref{NAR-sigfig} shows several examples of mutation profiles from both real and simulated tumors.
+The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi} = \{\pi_1,\dots,\pi_{96}\}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all mutations present in the tumor that occur at each of the 96 possible contexts.
+We define a mutation context $C$, indexed by $i=\{1 \dots 16\}$ flanking base combinations, $j = \{1,2\}$ reference bases, and $k = \{1,2,3\}$ alternate alleles. 
+The distribution of observed contexts $ p(C \mid M) $, where $M$ is the condition that a mutation has occurred, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j,k}\} $.
+In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
+As a result we need to model $p(C \mid M)$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,
+
+$$
+\begin{aligned}
+  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
+  C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
+\end{aligned}
+$$
+In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a-priori a mutation is equally likely in any context.
+
+Let $\mathrm{D}$ be the set of high confidence variant calls, and $c \in \mathrm{D}$ be the context of each mutation in $\mathrm{D}$.
+The posterior distribution of $\boldsymbol{\pi}$ is then $p(\boldsymbol{\pi} \mid D) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
+  $$
+    \alpha^{\prime}_{i,j,k} = \alpha_{i,j,k} + \sum\limits_{c \in \mathrm{D}} \mathrm{I}\{c = c_{i,j,k}\}.
+  $$
+Returning to equation \ref{eqn:full_prior}, we can compute the posterior predictive probability $p(C_{i,j} \mid M,\mathrm{D})$ as
+\begin{equation}
+  \label{eqn:post_pred}
+  $$
+  p(C = C_{i,j} \mid M,\mathrm{D}) = \frac{\sum_{k}\alpha^{\prime}_{i,j,k}}{\sum_{i,j,k}\alpha^{\prime}_{i,j,k}}.
+  $$
+\end{equation}
+The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i,j}$, $p(m \mid C_{i,j},\mathrm{D})$ is then
+\begin{equation}
+  \label{eqn:to_allele}
+  $$
+   p(m = k \mid C_{i,j},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j,k}}{\sum_{k} \alpha^{\prime}_{i,j,k}}.
+  $$
+\end{equation}
+The remaining context dependent component in equation \ref{eqn:full_prior} is $p(C)$, the prior probability of seeing a mutation with a particular tri-nucleotide context.
+We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i,j}$ divided by three.
+\bkmcomment{I have a pre-computed table in batcaver for wes and wgs.
+Should I add that bit here?}
+
+\subsection{Estimation of the mutation rate.}
+The final piece of equation \ref{eqn:full_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
+In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth and death rate.
+\citet{Williams2016} showed that if 
+$$
+  \beta = \frac{b-d}{b},
+$$
+then the number of new mutations per cell division is
+$$
+  \frac{N \mu}{\beta}.
+$$
+In a neutrally or nearly neutrally evolving tumor, they further show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
+\begin{equation}
+  \label{eqn:mut_rate}
+$$
+  \mathrm{M}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
+$$
+\end{equation}
+\bkmcomment{Gets weaker here in terms of how this needs to be said and how much of it needs to be said. Need to decide where to add the factor of two.}
+
+The number of bases $N$ is fixed at $3\cdot10^9$ for a whole genome sequence and $3\cdot10^7$ for a whole exome.
+Returning to the set of high confidence mutations $\mathrm{D}$, we can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, $f_{min}$ as the lowest, and count the number of mutations in $D$ falling between the two.
+In practice it is usually best to select a conservative value for $f_{min}$, which is strongly dependent on sequencing depth, and we include this as a parameter in the R implementation of \batcave.
+In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
+A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}. 
+Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu/\beta$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
+We now return to equation \ref{eqn:2} and describe the probability density function and method implemented in \batcave.
+
+\subsection{The mutect likelihood function}
+\bkmcomment{Assuming that the above full specification of the method is on track, this is where the mutect likelihood model goes along with a description of how \batcave folds it in and uses it.
+I could use some help with thinking about how far to go here.}
 
@@
 $$
-
-Assuming that the identity of the alternate allele and the estimated allele frequency are independent of each other, and $\nu$ is uniformly distributed results in the final model
-$$
-  P(m,\nu) = p(m) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}).
-$$
-
-\subsection{Estimation of the mutation signature}
-The 3-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
-There are four possible bases 3` and 5` of the site, leading to 16 possible flanking base combinations.
-Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site, and three possible alternate alleles for each.
-Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles (see Figure \ref{NAR-sigfig}).
-The mutation context $C$ is indexed by $i=\{1 \dots 16\}$ flanking base combinations, $j = \{1,2\}$ reference bases, and $j = \{1,2,3\}$ alternate alleles.
-We model $ p(C | M) $ as a multinomial distribution with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j,k}\} $ and pseudo-count hyper-parameter $\boldsymbol{\alpha}$
-
-$$
-\begin{array}{rl}
-  C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
-  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
-\end{array}
-$$
-We use high confidence variant calls -- odds in favor of the mutation greater than 10:1 -- to estimate $\boldsymbol{\pi}$.
-
-\subsection{Estimation of the mutation rate.}
-As with estimation of the mutation signature, we can use high confidence mutations and a model of tumor evolution to compute the tumor-specific mutation rate.
-\citet{Bozic2016} show that for any variant allele frequency $\alpha$, the total number of mutations with frequency greater than $\alpha$ and less than $0.25$ is
-
-$$
-N(\alpha) = N\mu \left( \frac{1}{\alpha} - \frac{1}{0.25} \right)
-$$
-
-Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
-By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\hat{\mu}$.
-In all results here we select $\alpha = 0.05$.
-
-\subsection{Site-specific prior probability of mutation}
-The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
-However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted here $p(m,M | C)$ \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
-$$
-\begin{array}{rl}
-  \mathrm{p}(m,M \mid C) = &\mathrm{p}(m \mid C) \mathrm{p}(M \mid C) \\ \\ 
-  = &\mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
-\end{array}
-$$
-since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
-The estimated mutation signature and mutation rate, allow us to compute $p(m,M | C)$ as
-
-$$
-\begin{array}{rl}
-  \mathrm{p}(m \mid C) = & \pi_{i,j,k} \\ \\
-  \mathrm{p}(C \mid M) = & \sum_{k}\pi_{i,j} \\ \\
-  \mathrm{p}(M) = & \hat{\mu},
-\end{array}
-$$
-with $\mathrm{p}(C)$ the fraction of all sites in the genome with context $C_{i,j}$ divided by 3.
-
 %`\subsection{False positive rate control.}
@@
 
-\subsection{Real tumor data.}
-\subsubsection{Acute Myeloid Leukemia}
-We downloaded the whole genome sequence for aml31 \bkmcomment{citation and download date, and primary/relapse distinction} \cite{Griffith2015}.
-We merged the gold and platinum lists, and define as \textit{not present} any variant for which deep sequencing was performed and zero alternate reads were observed.
-This is a very conservative metric.
-\subsubsection{What I call the cell paper} Not sure if this will get used.
-
-\subsection{Calibration}
-We use the Integrated Calibration Index to quantify the difference between MuTect and the Calibrated model \cite{Austin2019}.
-Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-fig2}\textbf{b,d}).
-Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the result and a diagonal line.
-
-\begin{figure}[t]
+\section{RESULTS}
+% \begin{figure*}[t]
+%   \begin{center}
+%   \includegraphics{figures/aml_plot.pdf}
+%   \end{center}
+%   \caption{Model performance on 300X Whole leukemia genome.
+%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
+%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
+%   A perfectly calibrated model would generate the diagonal line.}
+%   \label{NAR-aml}
+% \end{figure*}
+\begin{figure*}
   \begin{center}
-\includegraphics{figures/signature_figure.pdf}
+  \includegraphics{figures/fig2.pdf}
   \end{center}
-\caption{Mutation signatures. 
-\textbf{(a)} The observed mutation signature of the AML31 primary tumor. 
-\textbf{(b)} A mutation signature used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-\textbf{(c)} Equal proportions of signatures 1, 4, \& 5.
-\textbf{(d)} Equal proportions of signatures 1, 3, \& 5.
-KL divergence between the estimated signature and the simulated mutation signature for \textbf{e)} Whole exomes and \textbf{f)} whole genomes.
-The estimated mutation signature converges quickly to the target as the number of mutations evaluated rises.
-}
-\label{NAR-sigfig}
-\end{figure}
-
-\section{RESULTS}
-\begin{figure*}[t]
-  \begin{center}
-  \includegraphics{figures/aml_plot.pdf}
-  \end{center}
-  \caption{Model performance on 300X Whole leukemia genome.
-  \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
+  \caption{Model performance on 500X simulated whole exome.
+  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
+  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
   A perfectly calibrated model would generate the diagonal line.}
-  \label{NAR-aml}
+\label{NAR-wes_fig}
 \end{figure*}
 
-\subsection{Convergence of the mutation prior to the true mutation profile}
+\subsection{Simulated data}
+We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
+For each of the two types we simulated tumors with variants drawn from three different mutation profiles (Concentrated,Intermediate, and Diffuse) as described in Materials \& Methods.
+These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
+\bkmcomment{Should there be a longer discussion of what sorts of tumors the profiles represent, or leave that for the methods only?}
 
-The dirichlet prior probability of mutation converges quickly to the true mutation signature.
-The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-sigfig}\textbf{e,f}).
-\subsection{Performance on real data}
+We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are deeply unbalanced.
+By both metrics \batcave outperforms MuTect (Figure \ref{NAR-wes_fig} \textbf{a)} and \textbf{c)}, and Table \ref{table:01}).
+The extent to which \batcave outperforms is dependent on both the sequencing depth and the level of concentration of the mutation signature.
+Deeper sequencing and more concentrated mutation signatures increase the performance advantage of \batcave.
 
-The Calibrated model has better AUROC and AUPRC than the MuTect model in this real tumor (Figure \ref{NAR-aml}\textbf{a,c}, Table \ref{table:01}).
-The Calibrated model has slightly worse calibration as measure by ICI, but both models substantially underestimate the true positive probability throughout the range of possible values (Figure \ref{NAR-aml} \textbf{b,d}).
-One caveat to consider when considering these results is that while AML31 was subject to extensive variant calling validation, their list of true positive calls likely does not include every true variant at low allele frequencies, and true negative calls were not evaluated.
+\subsection{Real tumor data.}
+\bkmcomment{need to rework this figure. the template allows only one full width figure.}
 
-\subsection{Simulated data}
-
-\subsubsection{classification}
-We simulated both whole genome and whole exome tumors with three different mutation signatures.
-Measures of classification performance are positively correlated with signature concentration.
-For all simulations the calibrated model produces a better variant classifier than MuTect (Figure \ref{NAR-fig2} \textbf{a} and \textbf{c}).
-
-\subsubsection{calibration}
-The Calibrated model tends to score true negative variants lower and true positive variants higher than Mutect (Figure \ref{NAR-fig2} \textbf{b,d}).
-The Integrated calibration index for the Calibrated model is better than that of the MuTect model for all tumors simulated.
+\bkmcomment{add TPR/FPR at 1\% and 3\%?}
 
@@
 
-\begin{figure*}[t]
-\begin{center}
-\includegraphics{figures/fig2.pdf}
-\end{center}
-\caption{Model performance on 500X simulated whole exome.
-\textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
-A perfectly calibrated model would generate the diagonal line.
-}
-\label{NAR-fig2}
-\end{figure*}
+\subsection{Convergence of the mutation prior to the true mutation profile}
+
+The dirichlet prior probability of mutation converges quickly to the true mutation signature.
+The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-sigfig}\textbf{e,f}).
 
@@
 \subsection{Discussion subsection one}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text.
-
-\subsection{Discussion subsection two}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text.
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-
-\subsection{Discussion subsection three}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text.
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text.
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text.
 

### 2019-07-29 NAR_DRAFT.tex
check in

@@
 
+\bkmcomment{Introduce D earlier as its own section describing how we get it? Or expound here instead.}
+
 Let $\mathrm{D}$ be the set of high confidence variant calls, and $c \in \mathrm{D}$ be the context of each mutation in $\mathrm{D}$.
@@
 We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
+The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
 Variants were selected from TCGA and PCAWG variant files(dates).
@@
 We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
-For each of the two types we simulated tumors with variants drawn from three different mutation profiles (Concentrated,Intermediate, and Diffuse) as described in Materials \& Methods.
+For each of the two types we simulated tumors with variants drawn from three different mutation profiles (Concentrated,Intermediate, and Diffuse) as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
 These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
@@
 Deeper sequencing and more concentrated mutation signatures increase the performance advantage of \batcave.
+
+The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 \bkmcomment{filling in today} for all simulated tumors.
+We note that the effective mutation rate is lower than the simulated rate of 3e-7.
+In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.
+\bkmcomment{possible to quantify that this is very close to the actual insertion rate. Do it}.
+
+We measured calibration performance using the Integrated Calibration Index described in Methods \citep{Austin2019}.
 

### 2019-07-29 NAR_DRAFT.tex
comments

@@
 \newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
-\newcommand{\batcave}{\textbf{BATCAVE }}
+\newcommand{\batcave}{BATCAVE }
 % Enter dates of publication
@@
   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
-  A perfectly calibrated model would generate the diagonal line.}
+  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
 \label{NAR-wes_fig}
@@
 We measured calibration performance using the Integrated Calibration Index described in Methods \citep{Austin2019}.
+For all simulated tumors the calibration of \batcave is better across the full spectrum of posterior probabilities.
+As we can see in Figure \ref{NAR-wes_fig} \textbf{b)} and \textbf{d)}, \batcave tends to increase posterior probabilities of low probability but true variants (top density curves), and decrease the probability of low probability but false variants (bottom density curves.).
+As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.
 
@@
 \toprule
-Depth & Signature & $\mu$ & AUROC &  & AUPRC & & ICI & 
+Depth & Signature & $\mu/\beta$ & AUROC &  & AUPRC & & ICI & 
 \\
-& & (estimated) & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
+& & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
 \\
 \colrule
-100X whole genome & Concentrated & Row 1 & Row 1 & -- & -- & -- & -- & --
+100X whole genome & Concentrated & 3.6e-7 & Row 1 & -- & -- & -- & -- & --
 \\
@@
 \\
-AML 31 & Actual & $3.6e-8$ & .935 & .949 & .995 & .996 & 0.121 & 0.155
+AML 31 & Actual & 3.6e-8 & .935 & .949 & .995 & .996 & 0.121 & 0.155
 \\

### 2019-07-29 NAR_DRAFT.tex
simulation results

@@
 
+The heart of \batcave, the context-dependent prior probability of mutation, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
+We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
+In Figure \ref{NAR-sigfig} \textbf{e)} and \textbf{f)} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.
+
 \subsection{Real tumor data.}
@@
 \\
-100X whole genome & Intermediate & Row 1 & Row 1 & -- & -- & -- & -- & --
+100X whole genome & Intermediate & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
 \\

### 2019-07-29 NAR_DRAFT.tex
found a citation for different scales of genomic context

@@
 \subsection{Estimation of the mutation profile}
-There are many aspects of local genomic architecture, generally described as the genomic \textit{context}, which can affect the local somatic mutation rate at a particular site in the genome.
-Here we focus on the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \citet{Buisson2019}.
+Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
 The tri-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
@@
 The final piece of equation \ref{eqn:full_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
-In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth and death rate.
+In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth $b$ and death $d$ rates.
 \citet{Williams2016} showed that if 
@@
 \subsection{Tumor simulations.}
-
+\bkmcomment{if $\pi$ works above, describe profile as $\pi$, and $\pi$ as linear combination of 3 signatures.}
 We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
@@
 
-The heart of \batcave, the context-dependent prior probability of mutation, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
+The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
 We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
@@
 \toprule
-Depth & Signature & $\mu/\beta$ & AUROC &  & AUPRC & & ICI & 
+Depth & Mutation profile & $\mu/\beta$ & AUROC & & AUPRC & & ICI & 
 \\
@@
 \\
-100X whole genome & Diffuse & Row 1 & Row 1 & -- & -- & -- & -- & --
+100X whole genome & Diffuse & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
 \\
@@
 
-\subsection{Convergence of the mutation prior to the true mutation profile}
+\section{DISCUSSION}
 
-The dirichlet prior probability of mutation converges quickly to the true mutation signature.
-The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-sigfig}\textbf{e,f}).
-
-\section{DISCUSSION}
+Simulations suggest that if the mutation profile is the data generating distribution, \batcave can estimate effective mutation rate and data generating distribution, and doing so improves both classification and calibration.
 

### 2019-07-29 NAR_DRAFT.tex
mu estimates

@@
 \subsection{Estimation of the mutation profile}
-There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \citet{Buisson2019}.
+There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
 Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
@@
 
-Figure \ref{NAR-sigfig} shows several examples of mutation profiles from both real and simulated tumors.
+Figure \ref{NAR-sigfig} \textbf{a-d} shows several examples of mutation profiles from both real and simulated tumors.
 The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi} = \{\pi_1,\dots,\pi_{96}\}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all mutations present in the tumor that occur at each of the 96 possible contexts.
@@
 We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are deeply unbalanced.
-By both metrics \batcave outperforms MuTect (Figure \ref{NAR-wes_fig} \textbf{a)} and \textbf{c)}, and Table \ref{table:01}).
+By both metrics \batcave outperforms MuTect (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{c}, and Table \ref{table:01}).
 The extent to which \batcave outperforms is dependent on both the sequencing depth and the level of concentration of the mutation signature.
@@
 
-The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 \bkmcomment{filling in today} for all simulated tumors.
-We note that the effective mutation rate is lower than the simulated rate of 3e-7.
+The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 for all simulated tumors.
+We note that the effective mutation rate is lower than the simulated rate of 3e-6.
 In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.
@@
 For all simulated tumors the calibration of \batcave is better across the full spectrum of posterior probabilities.
-As we can see in Figure \ref{NAR-wes_fig} \textbf{b)} and \textbf{d)}, \batcave tends to increase posterior probabilities of low probability but true variants (top density curves), and decrease the probability of low probability but false variants (bottom density curves.).
+As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (top density curves), and decrease the probability of low probability but false variants (bottom density curves.).
 As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.
@@
 We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-In Figure \ref{NAR-sigfig} \textbf{e)} and \textbf{f)} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
 For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.
@@
 \\
-500X whole exome & Concentrated & Row 1 & .848 & .928 & .674 & .757 & .0585 & .0465
+500X whole exome & Concentrated & 3.6e-7 & .848 & .928 & .674 & .757 & .0585 & .0465
 \\
-500X whole exome & Intermediate & Row 1 & .847 & .842 & .677 & .707 & .0587 & .0566 
+500X whole exome & Intermediate & 3.6e-7 & .847 & .842 & .677 & .707 & .0587 & .0566 
 \\
-500X whole exome & Diffuse & Row 1 & .850 & .873 & .676 & .698 & .0565 & .0575
+500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .0565 & .0575
 \\

### 2019-07-29 NAR_DRAFT.tex

@@
   \begin{center}
-  \includegraphics{figures/fig2.pdf}
+  \includegraphics{figures/fig_wes.pdf}
   \end{center}
@@
 \colrule
-100X whole genome & Concentrated & 3.6e-7 & Row 1 & -- & -- & -- & -- & --
+100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
 \\
-100X whole genome & Intermediate & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
+100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
 \\
-100X whole genome & Diffuse & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
+100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
 \\
-500X whole exome & Concentrated & 3.6e-7 & .848 & .928 & .674 & .757 & .0585 & .0465
+500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .0585 & .0462
 \\
-500X whole exome & Intermediate & 3.6e-7 & .847 & .842 & .677 & .707 & .0587 & .0566 
+500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .0587 & .0568 
 \\
-500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .0565 & .0575
+500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .0565 & .0576
 \\
@@
 
-Simulations suggest that if the mutation profile is the data generating distribution, \batcave can estimate effective mutation rate and data generating distribution, and doing so improves both classification and calibration.
+-- Simulations suggest that if the mutation profile is the data generating distribution, \batcave can estimate effective mutation rate and data generating distribution, and doing so improves both classification and calibration.
 
+-- One hope is to enable higher confidence in low frequency alleles as sequencing depth increases.
+Depth seems to have a very large effect on performance i.e.
+results are more compelling for 500X wes than for 100x wgs, although if you do 100 wgs you will still benefit.
+In other words sequence depth matters more performance wise than sequencing breadth if the only desire is to see low frequency alleles.
+No advantage is gained (after a certain depth) of adding variants(sites) by wgs sequencing.
 \subsection{Discussion subsection one}
@@
 
+\section{Supplementary Figures}
+
+\begin{figure*}
+  \begin{center}
+  \includegraphics{figures/fig_wes.pdf}
+  \end{center}
+  \caption{Model performance on 100X simulated whole genome.
+  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
+  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
+  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
+\label{NAR-wgs_fig}
+\end{figure*}
+
 \end{document}

### 2019-07-29 NAR_DRAFT.tex
fixes

@@
 
-\begin{figure*}
+\begin{figure*}[t]
   \begin{center}
-  \includegraphics{figures/fig_wes.pdf}
+  \includegraphics{figures/aml_plot.pdf}
   \end{center}
-  \caption{Model performance on 100X simulated whole genome.
-  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
-  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
-\label{NAR-wgs_fig}
+  \caption{Model performance on 300X Whole leukemia genome.
+  \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
+  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
+  A perfectly calibrated model would generate the diagonal line.}
+  \label{NAR-aml}
 \end{figure*}

### 2019-07-31 NAR_DRAFT.tex
new figures

@@
 The probability which we have denoted $p(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
-However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $p(m,M | C)$ \bkmcomment{citation for genomic context in general} .
+However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $p(m,M | C)$ \cite{Buisson2019}.
 Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as
@@
 
+\subsection{Evaluation methods}
+ROC
+PRC
+ICI. We used loess smoothing "loess" function in R with span = 0.5.
 \begin{tiny}
@@
 
+\begin{figure*}
+  \begin{center}
+  \includegraphics{figures/fig_wes.pdf}
+  \end{center}
+  \caption{Model performance on 100X simulated whole genome.
+  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
+  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
+  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
+\label{NAR-wgs_fig}
+\end{figure*}
 \end{document}

### 2019-08-01 NAR_DRAFT.tex
add substitution type

@@
 
-At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
+At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
 We want to use $\mathbf{x}$ to select between competing hypotheses;
@@
 $$
-  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
+  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
 $$
@@
 Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The tri-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
+The tri-nucleotide context at a particular genomic site is made up of the identity of the reference base and the identity of the 3` and 5` flanking bases.
 There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
-Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, and three possible alternate alleles {\{A,(C/T),G\}} for each.
-Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles.
+Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
+We define a mutation context $C$ indexed by $i=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
+For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
+We can thus define $\mathrm{S}$, the substitution type, indexed by $i=\{1 \dots 32\}$ contexts and $j = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
+Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as
+
+\begin{equation}
+  \label{eqn:detailed_prior}
+  $$
+  \begin{array}{rcc}
+  \mathrm{p}(\mathrm{S}) &=&  \mathrm{p}(m =j, M | C = i) \\
+                            &=& \mathrm{p}(m = j \mid C = i) \mathrm{p}(C = i \mid M)\frac{p(M)}{p(C = i)},
+  \end{array}
+  $$
+\end{equation}
+where $\mathrm{p}(\mathrm{S})$ is the probability that a substitution of type $S$ has occurred.
 
@@
   \caption{Mutation profiles. 
-  \textbf{(a)} The observed mutation profile of the AML31 primary tumor. 
-  \textbf{(b)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(c)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(d)} Equal proportions of signatures 1, 3, \& 5.
-  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{e)} Whole exomes and \textbf{f)} whole genomes.
+  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
+  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
+  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
+  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
   The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
@@
 
-Figure \ref{NAR-sigfig} \textbf{a-d} shows several examples of mutation profiles from both real and simulated tumors.
-The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi} = \{\pi_1,\dots,\pi_{96}\}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all mutations present in the tumor that occur at each of the 96 possible contexts.
-We define a mutation context $C$, indexed by $i=\{1 \dots 16\}$ flanking base combinations, $j = \{1,2\}$ reference bases, and $k = \{1,2,3\}$ alternate alleles. 
-The distribution of observed contexts $ p(C \mid M) $, where $M$ is the condition that a mutation has occurred, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j,k}\} $.
+Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
+The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
+The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j}\} $.
 In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
-As a result we need to model $p(C \mid M)$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,
+As a result we need to model $p(S)$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,
 
@@
   \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
-  C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
+  S \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
 \end{aligned}
@@
 Let $\mathrm{D}$ be the set of high confidence variant calls, and $c \in \mathrm{D}$ be the context of each mutation in $\mathrm{D}$.
-The posterior distribution of $\boldsymbol{\pi}$ is then $p(\boldsymbol{\pi} \mid D) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
+The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
   $$
-    \alpha^{\prime}_{i,j,k} = \alpha_{i,j,k} + \sum\limits_{c \in \mathrm{D}} \mathrm{I}\{c = c_{i,j,k}\}.
+    \alpha^{\prime}_{i,j} = \alpha_{i,j} + \sum\limits_{c \in \mathrm{D}} \mathrm{I}\{c = c_{i,}\}.
   $$
-Returning to equation \ref{eqn:full_prior}, we can compute the posterior predictive probability $p(C_{i,j} \mid M,\mathrm{D})$ as
+Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C_{i} \mid M,\mathrm{D})$ as
 \begin{equation}
@@
   $$
-  p(C = C_{i,j} \mid M,\mathrm{D}) = \frac{\sum_{k}\alpha^{\prime}_{i,j,k}}{\sum_{i,j,k}\alpha^{\prime}_{i,j,k}}.
+  p(C = C_{i} \mid M,\mathrm{D}) = \frac{\sum_{j}\alpha^{\prime}_{i,j}}{\sum_{i,j}\alpha^{\prime}_{i,j}}.
   $$
 \end{equation}
-The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i,j}$, $p(m \mid C_{i,j},\mathrm{D})$ is then
+The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i}$, $p(m = j \mid C_{i},\mathrm{D})$ is then
 \begin{equation}
@@
   $$
-   p(m = k \mid C_{i,j},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j,k}}{\sum_{k} \alpha^{\prime}_{i,j,k}}.
+   p(m = j \mid C_{i},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j}}{\sum_{j} \alpha^{\prime}_{i,j}}.
   $$
 \end{equation}
-The remaining context dependent component in equation \ref{eqn:full_prior} is $p(C)$, the prior probability of seeing a mutation with a particular tri-nucleotide context.
-We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i,j}$ divided by three.
-\bkmcomment{I have a pre-computed table in batcaver for wes and wgs.
-Should I add that bit here?}
+The remaining context dependent component in equation \ref{eqn:full_prior} is $p(C = C_{i})$, the prior probability of seeing a particular tri-nucleotide context.
+We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i}$ present in the genome.
+\bkmcomment{I have a pre-computed table in batcaver for wes and wgs. Should I add that bit here?}
 

### 2019-08-02 NAR_DRAFT.tex
small fix
check in methods changes

@@
 In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
-As a result we need to model $p(S)$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,
+As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,
 
@@
   \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
-  S \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
+  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
 \end{aligned}
@@
 
-\bkmcomment{Introduce D earlier as its own section describing how we get it? Or expound here instead.}
-
-Let $\mathrm{D}$ be the set of high confidence variant calls, and $c \in \mathrm{D}$ be the context of each mutation in $\mathrm{D}$.
+In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
+These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant.
+Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
 The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
   $$
-    \alpha^{\prime}_{i,j} = \alpha_{i,j} + \sum\limits_{c \in \mathrm{D}} \mathrm{I}\{c = c_{i,}\}.
+    \alpha^{\prime}_{i,j} = \alpha_{i,j} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{i,j}\}.
   $$
-Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C_{i} \mid M,\mathrm{D})$ as
+Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = C_{i} \mid M,\mathrm{D})$ as
 \begin{equation}
@@
 \end{equation}
-The remaining context dependent component in equation \ref{eqn:full_prior} is $p(C = C_{i})$, the prior probability of seeing a particular tri-nucleotide context.
+The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = C_{i})$, the prior probability of seeing a particular tri-nucleotide context.
 We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i}$ present in the genome.
-\bkmcomment{I have a pre-computed table in batcaver for wes and wgs. Should I add that bit here?}
 
 \subsection{Estimation of the mutation rate.}
-The final piece of equation \ref{eqn:full_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
+The final piece of equation \ref{eqn:detailed_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
 In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth $b$ and death $d$ rates.
@@
 $$
+We make the simplifying assumption that $d = 0$, such that $\beta = 1$
 In a neutrally or nearly neutrally evolving tumor, they further show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
@@
 \end{equation}
-\bkmcomment{Gets weaker here in terms of how this needs to be said and how much of it needs to be said. Need to decide where to add the factor of two.}
 
@@
 A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
-Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu/\beta$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
+We make the simplifying assumption that $d = 0$, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
+Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
 We now return to equation \ref{eqn:2} and describe the probability density function and method implemented in \batcave.
@@
 \subsection{The mutect likelihood function}
-\bkmcomment{Assuming that the above full specification of the method is on track, this is where the mutect likelihood model goes along with a description of how \batcave folds it in and uses it.
-I could use some help with thinking about how far to go here.}
-
-$$
-  \textrm{f}_{m,\nu}({b_i,q_i}) = \left\{
-    \begin{array}{cr}
-      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\ \\
-      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\ \\
-      \frac{10^{-q_i/10}}{3} & otherwise.
-    \end{array}
-    \right.
-$$
-%`\subsection{False positive rate control.}
-
-%We develop a method, following \citet{Efron2008}, for controlling the false positive rate.
-%Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either \textit{null} (non-variant with $\nu = 0$) or \textit{nonnull} (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,
-
-%$$
-%\begin{array}{ll}
-%p_0 = \textrm{P}\{\textrm{null}\} \quad & \textrm{f}_{0}(\mathbf{x}) \quad \textrm{density if null}\\
-%p_1 = \textrm{P}\{\textrm{nonnull}\} \quad & \textrm{f}_{1}(\mathbf{x}) \quad \textrm{density if nonnull} .
-%\end{array}
-%$$
-
-%The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
-%In a neutrally evolving tumor, the number of cells is growing exponentially, and the count of variants with an allele frequency greater than a given allele frequency $f$ is \cite{Bozic2016,Williams2016a}.
-%$$
-%N(f) = \frac{N\mu}{f},
-%$$
-%Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
-%The estimated fraction of all of the sites in the genome that will have a mutation with frequency $f$ is
-%$$
-%\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .01} -  \frac{\mu}{f + .01}
-%$$
-
-%\bkmcomment{This is wrong. This is the cdf. I need the integral of the derivative which is the density.}
-%and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
-%\citet{Williams2016a} provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.
+\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or > 2.0)\cite{Cibulskis2013}.
+\bkmcomment{need a little more time with this.}
 

### 2019-08-02 NAR_DRAFT.tex
changes

@@
 $$
-  P(m,\nu) = p(m) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i).
+  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
 $$
@@
   $$
-  \begin{array}{rcc}
-  \mathrm{p}(\mathrm{S}) &=&  \mathrm{p}(m =j, M | C = i) \\
+  \begin{array}{rcl}
+  \mathrm{p}(\mathrm{S_{i,j}}) &=&  \mathrm{p}(m =j, M | C = i) \\
                             &=& \mathrm{p}(m = j \mid C = i) \mathrm{p}(C = i \mid M)\frac{p(M)}{p(C = i)},
@@
 \end{equation}
-where $\mathrm{p}(\mathrm{S})$ is the probability that a substitution of type $S$ has occurred.
+where $\mathrm{p}(\mathrm{S_{i,j}})$ is the probability that a substitution of type $\mathrm{S_{i,j}}$ will occur.
 
@@
 We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
-For each of the two types we simulated tumors with variants drawn from three different mutation profiles (Concentrated,Intermediate, and Diffuse) as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
+For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a breast cancer (intermediate), and a lung cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
 These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
-\bkmcomment{Should there be a longer discussion of what sorts of tumors the profiles represent, or leave that for the methods only?}
 
@@
 \toprule
-Depth & Mutation profile & $\mu/\beta$ & AUROC & & AUPRC & & ICI & 
+Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
 \\
@@
 \\
-AML 31 & Actual & 3.6e-8 & .935 & .949 & .995 & .996 & 0.121 & 0.155
+Griffith et al. 2015 & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+\\
+Shi et al. 2018 & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
 \\

### 2019-08-02 NAR_DRAFT.tex
changes

@@
 $$
-We make the simplifying assumption that $d = 0$, such that $\beta = 1$
 In a neutrally or nearly neutrally evolving tumor, they further show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is

### 2019-08-05 NAR_DRAFT.tex
morning
finish methods

@@
 We want to use $\mathbf{x}$ to select between competing hypotheses;
+\begin{equation}
+  \label{eqn:hypothesis}
 $$
@@
 $$
+\end{equation}
 where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e.
@@
 $$
-  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
+  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
 $$
@@
 
-\subsection{The mutect likelihood function}
-\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or > 2.0)\cite{Cibulskis2013}.
-\bkmcomment{need a little more time with this.}
+\subsection{The likelihood function}
+\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or 2.0) \cite{Cibulskis2013}.
+MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
+In addition, MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
+We have used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
+
+\begin{equation}
+  \label{eqn:tlod}
+    $$
+    \mathrm{TLOD} = \frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}.
+    $$
+\end{equation}
+
+The log posterior odds is the log likelihood ratio (\texttt{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as
+
+\begin{equation}
+  \label{eqn:computed_posterior}
+  $$
+  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{log}_{10}(\mathrm{logit}(\mathrm{p}(\mathrm{S_{i,j}}))},
+  $$
+\end{equation}
+where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{i,j}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
+When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
+They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
+We compute the prior odds for Mutect as
+\begin{equation}
+  \label{eqn:mutect_posterior}
+  $$
+  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
+  $$
+\end{equation}
+since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$.
 
 \subsection{Tumor simulations.}
-\bkmcomment{if $\pi$ works above, describe profile as $\pi$, and $\pi$ as linear combination of 3 signatures.}
 We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
+Tumors were simulated with three different mutation profiles, all of which include a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
+The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
+The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is common in lung cancers.
+The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
 The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
-Variants were selected from TCGA and PCAWG variant files(dates).
+Variants were selected from a combination of Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
 Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
@@
 
-\subsection{Evaluation methods}
-ROC
-PRC
-ICI. We used loess smoothing "loess" function in R with span = 0.5.
 \begin{tiny}
@@
 We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
-For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a breast cancer (intermediate), and a lung cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
+For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
 These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
@@
 In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.
-\bkmcomment{possible to quantify that this is very close to the actual insertion rate. Do it}.
 
@@
 \\
-Griffith et al. 2015 & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+\citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
 \\
-Shi et al. 2018 & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
+\citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
 \\

### 2019-08-05 NAR_DRAFT.tex
typo
starting real data

@@
 They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
-We compute the prior odds for Mutect as
+We compute the posterior odds for Mutect as
 \begin{equation}
@@
 \end{figure*}
+\subsection{Convergence of the site-specific prior to the data generating distribution}
+
+The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
+We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
+In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.
 
@@
 
-The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 for all simulated tumors.
-We note that the effective mutation rate is lower than the simulated rate of 3e-6.
+The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table \ref{table:01}).
+We note that the effective mutation rate is lower than the simulated rate of $3\cdot10^{-6}$.
 In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.
@@
 
-The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
-We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.
+\subsection{Real tumor data.}
+We utilized two real data sets where deep sequencing and variant validation were performed, with the specific purpose of evaluating tumor variant calling pipelines.
+\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
+We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
+We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
+The area under the precision recall curve is essentially identical for MuTect and \batcave (RD figure coming and Table \ref{table:01}).
 
-\subsection{Real tumor data.}
-\bkmcomment{need to rework this figure. the template allows only one full width figure.}
-
-\bkmcomment{add TPR/FPR at 1\% and 3\%?}
+\
 

### 2019-08-08 NAR_DRAFT.tex
local changes

@@
 Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
-Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018}.
+Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
 Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
@@
 The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The second, and far more complex, process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
+The second -- and far more complex -- process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
 Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
@@
 \end{equation}
-Here we show how \batcave can be used to provide a tumor-specific estimate of the mutation probability $p(m)$.
+Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $p(m)$.
 
@@
 Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
-We now return to equation \ref{eqn:2} and describe the probability density function and method implemented in \batcave.
 
 \subsection{The likelihood function}
-\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or 2.0) \cite{Cibulskis2013}.
+We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
 MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
@@
 We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision recall curve is essentially identical for MuTect and \batcave (RD figure coming and Table \ref{table:01}).
+The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (RD figure coming and Table \ref{table:01}).
 
-\
+\citet{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
+We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
+We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave.
+
+These two data sets are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
+We report only precision-recall comparisons here because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
+There is thus no way to compute true and false negative fractions for these data sets.
+As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
 

### 2019-08-08 NAR_DRAFT.tex
discussion outline

@@
 
--- Simulations suggest that if the mutation profile is the data generating distribution, \batcave can estimate effective mutation rate and data generating distribution, and doing so improves both classification and calibration.
-
--- One hope is to enable higher confidence in low frequency alleles as sequencing depth increases.
-Depth seems to have a very large effect on performance i.e.
-results are more compelling for 500X wes than for 100x wgs, although if you do 100 wgs you will still benefit.
-In other words sequence depth matters more performance wise than sequencing breadth if the only desire is to see low frequency alleles.
-No advantage is gained (after a certain depth) of adding variants(sites) by wgs sequencing.
-\subsection{Discussion subsection one}
+\begin{itemize}
+  \item Every tumor is different. Even tumors of the same type have meaningfully different mutation profiles
+  \item The solution must be specific to each tumor we examine
+  \item The fact that we can think of tumor variant calls as two sets of data, with different measurement error, derived from the same data generating process, makes this possible
+  \item The mutation profile is estimated from the high-confidence (low measurement error) variants, and used as a prior for the low-confidence (high measurement error) variants
+  \item We demonstrate that the mutation generating process (mutation profile) can be reliably retrieved from simulated NGS data.
+  \item We also demonstrate that on the limited real data available, which consists of validated high confidence mutations, the results are equivalent
+  \item What is needed is validation experiments that also allow for the estimation of true and false negative rates, in order to fully illuminate the utility of \batcave
+  \item We have devoloped this algorithm as a post-calling evaluation tool to be used at the end of a variant calling pipeline
+  \item The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
+  \item The benefit from the algorithm varies with both the concentration of the signature and the depth of sequencing
+  \item Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors
+  \item We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
+  \item One size does not fit all with respect to the evidence threshold.
+Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
+  \item We believe the improved calibration achieved with this algorithm will encourage more thoughtful consideration of thresholds in various variant calling contexts
+\end{itemize}
 

### 2019-08-12 NAR_DRAFT.tex
discussion outline more
address Ryan comments

@@
 Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
+The specific processes driving mutations in a particular tumor generate a tumor's unique mutation profile.
+Because profiles vary widely both within and between tumor types.
 
 Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
-We then extend the method with an application of the local false discovery rate by computing the probability that a site is non-null under an assumption of clonal expansion with either early or small selective differences between clones.
 We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
@@
 \subsection{The likelihood function}
-We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
-In addition, MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
-We have used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
+We designed \batcave to be used as a post-calling evaluation tool with MuTect
+MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
+We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
 
@@
 
+\subsection{Real Data}
+\begin{itemize}
+  \item aml31 \cite{Griffith2015}
+  \item item used only the platinum calls for evaluation
+  \item Shi2018 \cite{Shi2018}
+  \item used the gold standard calls, and combined all six biological replicates 
+\end{itemize}
+
+\subsection{Calibration metric}
+
+	We use the Integrated Calibration Index to quantify the difference between MuTect and the Calibrated model \cite{Austin2019}.
+	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-wes_fig} \textbf{e}).
+	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the mean area between the result and a diagonal line.
+
 \section{RESULTS}
-% \begin{figure*}[t]
-%   \begin{center}
-%   \includegraphics{figures/aml_plot.pdf}
-%   \end{center}
-%   \caption{Model performance on 300X Whole leukemia genome.
-%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
-%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
-%   A perfectly calibrated model would generate the diagonal line.}
-%   \label{NAR-aml}
-% \end{figure*}
+We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
+MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
+\batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
+The mutation profile is computed from the high probability sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
+
 \begin{figure*}
@@
   \caption{Model performance on 500X simulated whole exome.
-  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
+  \textbf{(a)} Precision recall curves and \textbf{(c)} Receiver operating characteristic curves for 3 mutation signatures.
   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
-  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
+  A perfectly calibrated model would generate the diagonal line.}
 \label{NAR-wes_fig}
@@
 
-We measured calibration performance using the Integrated Calibration Index described in Methods \citep{Austin2019}.
+We measured calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
+The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
+This is useful in our setting because the density of predicted probabilities tends to be strongly bi-modal.
 For all simulated tumors the calibration of \batcave is better across the full spectrum of posterior probabilities.

### 2019-08-12 NAR_DRAFT.tex
new bib

@@
 The final piece of equation \ref{eqn:detailed_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
-In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth $b$ and death $d$ rates.
-\citet{Williams2016} showed that if 
-$$
-  \beta = \frac{b-d}{b},
-$$
-then the number of new mutations per cell division is
-$$
-  \frac{N \mu}{\beta}.
-$$
-In a neutrally or nearly neutrally evolving tumor, they further show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
+In an exponentially growing and neutrally or nearly neutrally evolving tumor, \citet{Williams2018} show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
 \begin{equation}
@@
 A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
-We make the simplifying assumption that $d = 0$, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
+The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
+We make the simplifying assumption that there is no cell death, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
 Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
@@
 \end{equation}
-since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$.
+since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.
 
@@
 The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
-The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is common in lung cancers.
+The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
 The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
@@
 \batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
-The mutation profile is computed from the high probability sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
-
+The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
 \begin{figure*}
@@
 This is useful in our setting because the density of predicted probabilities tends to be strongly bi-modal.
-For all simulated tumors the calibration of \batcave is better across the full spectrum of posterior probabilities.
 As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (top density curves), and decrease the probability of low probability but false variants (bottom density curves.).
+For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse.
+This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.
 As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.
@@
 \subsection{Real tumor data.}
-We utilized two real data sets where deep sequencing and variant validation were performed, with the specific purpose of evaluating tumor variant calling pipelines.
+We utilized two real data sets.
+In each, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
+These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
 \citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
+We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
 The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (RD figure coming and Table \ref{table:01}).
@@
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave.
-
-These two data sets are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
-We report only precision-recall comparisons here because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
-There is thus no way to compute true and false negative fractions for these data sets.
-As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
+We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave.
 
@@
 
+\batcave is a post-call variant evalutation tool that 
+We demonstrated here that this method improves classification accuracy in both simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
+
+We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
+These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
+At the same time, there is no way to compute true and false negative fractions for these data sets.
+As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
+Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+
 \begin{itemize}
-  \item Every tumor is different. Even tumors of the same type have meaningfully different mutation profiles
+  \item Every tumor is different. Differences within and between tumor types are pronounced \cite{@Stephens2005, @Burrell2005, @Nakamura2015, @Witkeiwicz2015, @Kumar2016}
   \item The solution must be specific to each tumor we examine

### 2019-08-12 NAR_DRAFT.tex
add discussion citations

@@
 \begin{itemize}
-  \item Every tumor is different. Differences within and between tumor types are pronounced \cite{@Stephens2005, @Burrell2005, @Nakamura2015, @Witkeiwicz2015, @Kumar2016}
+  \item Every tumor is different. Differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}
   \item The solution must be specific to each tumor we examine
@@
   \item Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors
+  \item Deep sequencing is useful \cite{Zehir2017}
   \item We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
@@
 Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
+  \item Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
+  \item One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
   \item We believe the improved calibration achieved with this algorithm will encourage more thoughtful consideration of thresholds in various variant calling contexts
+  \item The field believes that little information can be gained by wes rather than panels, but they are assuming the only information is mutations \cite{Hyman2015}
 \end{itemize}

### 2019-08-12 NAR_DRAFT.tex
check in changes

@@
 Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
-We define a mutation context $C$ indexed by $i=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
+We define a mutation context $C$ indexed by $c=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
 For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
-We can thus define $\mathrm{S}$, the substitution type, indexed by $i=\{1 \dots 32\}$ contexts and $j = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
+We can thus define $\mathrm{S}$, the substitution type, indexed by $c=\{1 \dots 32\}$ contexts and $m = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
 Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as
@@
   \begin{array}{rcl}
-  \mathrm{p}(\mathrm{S_{i,j}}) &=&  \mathrm{p}(m =j, M | C = i) \\
-                            &=& \mathrm{p}(m = j \mid C = i) \mathrm{p}(C = i \mid M)\frac{p(M)}{p(C = i)},
+  \mathrm{p}(\mathrm{S_{m,c}}) &=&  \mathrm{p}(m, M | C = c) \\
+                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{p(M)}{p(C = c)},
   \end{array}
@@
 \end{equation}
-where $\mathrm{p}(\mathrm{S_{i,j}})$ is the probability that a substitution of type $\mathrm{S_{i,j}}$ will occur.
+where $\mathrm{p}(\mathrm{S_{m,c}})$ is the probability that a substitution of type $\mathrm{S_{m,c}}$ will occur.
 
@@
 The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
-The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j}\} $.
+The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
 In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
@@
   $$
-    \alpha^{\prime}_{i,j} = \alpha_{i,j} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{i,j}\}.
+    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
   $$
-Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = C_{i} \mid M,\mathrm{D})$ as
+Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
 \begin{equation}
@@
   $$
-  p(C = C_{i} \mid M,\mathrm{D}) = \frac{\sum_{j}\alpha^{\prime}_{i,j}}{\sum_{i,j}\alpha^{\prime}_{i,j}}.
+  p(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
   $$
 \end{equation}
-The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i}$, $p(m = j \mid C_{i},\mathrm{D})$ is then
+The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
 \begin{equation}
@@
   $$
-   p(m = j \mid C_{i},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j}}{\sum_{j} \alpha^{\prime}_{i,j}}.
+   p(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
   $$
 \end{equation}
-The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = C_{i})$, the prior probability of seeing a particular tri-nucleotide context.
-We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i}$ present in the genome.
+The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
+We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the genome.
 

### 2019-08-12 NAR_DRAFT.tex
check in change

@@
 The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
-We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the genome.
+We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
 

### 2019-08-12 NAR_DRAFT.tex
fix logit

@@
   $$
-  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{log}_{10}(\mathrm{logit}(\mathrm{p}(\mathrm{S_{i,j}}))},
+  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S_{m,c}}))},
   $$
 \end{equation}
-where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{i,j}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
+where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{m,c}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
 When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
@@
 \end{equation}
-since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.
+since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.
 

### 2019-08-13 NAR_DRAFT.tex
working on discussion

@@
 Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
-The specific processes driving mutations in a particular tumor generate a tumor's unique mutation profile.
-Because profiles vary widely both within and between tumor types.
+The specific processes driving mutations in a particular tumor generate a tumor's unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}
 
@@
 \batcave is a post-call variant evalutation tool that 
-We demonstrated here that this method improves classification accuracy in both simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
-
+We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
 We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
@@
 Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+
+We have chosen to implement \batcave as a post-calling algorithm for use with MuTect because it presents a simple model and a well-tested set of heuristic filters.
+However, in principle there is no barrier to using the method with other variant callers.
+For instance Strelka, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability of that the tumor carries a somatic mutation is set to a constant.
+Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
+In general we are advancing the idea that there is substantial prior information about the site-specific mutation probability encoded in NGS data, and we should be using it.
 

### 2019-08-13 NAR_DRAFT.tex
another paragraph

@@
 However, in principle there is no barrier to using the method with other variant callers.
-For instance Strelka, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability of that the tumor carries a somatic mutation is set to a constant.
+For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
 Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
-In general we are advancing the idea that there is substantial prior information about the site-specific mutation probability encoded in NGS data, and we should be using it.
+In general we are advancing the idea that there is substantial prior information about the tumor and site-specific mutation probability encoded in NGS data, and we should be using it.
 

### 2019-08-13 NAR_DRAFT.tex
commit to merge

--- 2019-08-12 NAR_DRAFT.tex
@@
 Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
-The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
+The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{m,c}$.
 The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
@@
 \end{equation}
-where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{m,c}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
+where $\mathrm{p}(\mathrm{S_{m,c}})$ is the prior probability of a substitution of type $\mathrm{S_{m,c}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
 When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.

### 2019-08-13 NAR_DRAFT.tex
fix TLOD
fix a bunch of p

@@
 the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
-The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
+The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
 Assuming that reads are independent, we have
@@
 Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e.
-$p(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
+$\mathrm{p}(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
 \begin{equation}
@@
 \subsection{Site-specific prior probability of mutation}
-The probability which we have denoted $p(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
-However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $p(m,M | C)$ \cite{Buisson2019}.
+The probability which we have denoted $\mathrm{p}(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $\mathrm{p}(m,M)$.
+However, $\mathrm{p}(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
 Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as
@@
   $$
-  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}.
+  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
   $$
@@
   $$
-  p(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
+  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
   $$
@@
   $$
-   p(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
+   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
   $$
@@
     $$
-    \mathrm{TLOD} = \frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}.
+    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
     $$

### 2019-08-14 NAR_DRAFT.tex
morning

@@
 
-\batcave is a post-call variant evalutation tool that 
+\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
 We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
@@
 
-We have chosen to implement \batcave as a post-calling algorithm for use with MuTect because it presents a simple model and a well-tested set of heuristic filters.
+The \batcave algorithm focuses on the tri-nucleotide context and substitution type, but there are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
+These other types of context could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
+
+We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
 However, in principle there is no barrier to using the method with other variant callers.

### 2019-08-14 NAR_DRAFT.tex
work

--- 2019-08-13 NAR_DRAFT.tex
@@
 The specific processes driving mutations in a particular tumor generate a tumor's unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}
+Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
 
@@
 
-Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
 The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{m,c}$.
-The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
+The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is modeled as multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
 In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
@@
 These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant.
-Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
+Let $\mathrm{D}$ be the set of high confidence variant calls -- which we define as having posterior odds greater than 10 to 1 -- and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
 The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
@@
 	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-wes_fig} \textbf{e}).
-	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the mean area between the result and a diagonal line.
+	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average area between the loess fit and a diagonal line.
 
@@
 We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
+MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
 \batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
@@
 \subsection{Convergence of the site-specific prior to the data generating distribution}
-
 The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
-We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
+We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
 In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.
+For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for our real datasets are \bkmcomment{X} and for our simulated data sets are 1500 to 5000.
 
@@
 
-We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are deeply unbalanced.
+We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
 By both metrics \batcave outperforms MuTect (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{c}, and Table \ref{table:01}).
@@
 \\
-500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .0585 & .0462
+500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
 \\
-500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .0587 & .0568 
+500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
 \\
-500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .0565 & .0576
+500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
 \\
@@
 
-\begin{figure*}[t]
-  \begin{center}
-  \includegraphics{figures/aml_plot.pdf}
-  \end{center}
-  \caption{Model performance on 300X Whole leukemia genome.
-  \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
-  A perfectly calibrated model would generate the diagonal line.}
-  \label{NAR-aml}
-\end{figure*}
+% \begin{figure*}[t]
+%   \begin{center}
+%   \includegraphics{figures/aml_plot.pdf}
+%   \end{center}
+%   \caption{Model performance on 300X Whole leukemia genome.
+%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
+%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
+%   A perfectly calibrated model would generate the diagonal line.}
+%   \label{NAR-aml}
+% \end{figure*}
 

### 2019-08-15 NAR_DRAFT.tex
check in

@@
 \subsection{Real Data}
+We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
+We downloaded the primary and tumor bam files generated for the whole genome sequences in \cite{Griffith2015}, from dbGaP accession number phs000159.v8.p4.
+\citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
+The whole exome sequencing bam files generated by \cite{Shi2018} were downloaded from NCBI SRA no. SRP070662.
+\citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
+We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.
+
 \begin{itemize}

### 2019-08-15 NAR_DRAFT.tex
small fixes

@@
 We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
-We downloaded the primary and tumor bam files generated for the whole genome sequences in \cite{Griffith2015}, from dbGaP accession number phs000159.v8.p4.
+We downloaded the normal and primary tumor bam files generated for the whole genome sequences in \citet{Griffith2015} from dbGaP accession number phs000159.v8.p4.
 \citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
-The whole exome sequencing bam files generated by \cite{Shi2018} were downloaded from NCBI SRA no. SRP070662.
+The whole exome sequencing bam files generated by \citet{Shi2018} were downloaded from NCBI SRA no. SRP070662.
 \citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
 We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.
-
-\begin{itemize}
-  \item aml31 \cite{Griffith2015}
-  \item item used only the platinum calls for evaluation
-  \item Shi2018 \cite{Shi2018}
-  \item used the gold standard calls, and combined all six biological replicates 
-\end{itemize}
 
@@
 	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-wes_fig} \textbf{e}).
-	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average area between the loess fit and a diagonal line.
+	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.
 
@@
 
-\begin{table*}[b]
-\tableparts{%
-\caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
-\label{table:01}%
-}{%
-\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
-\toprule
-Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
-\\
-& & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
-\\
-\colrule
-100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
-\\
-100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
-\\
-100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
-\\
-500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
-\\
-500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
-\\
-500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
-\\
-\citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
-\\
-\citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
-\\
-\botrule
-\end{tabular*}%
-}
-{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
-\end{table*}
-
 \section{DISCUSSION}
@@
 
-The \batcave algorithm focuses on the tri-nucleotide context and substitution type, but there are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
-These other types of context could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
+The \batcave algorithm focuses on the tri-nucleotide context and substitution type. 
+There are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
+These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
 
@@
 \end{figure*}
+
+\begin{table*}[b]
+  \tableparts{%
+  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
+  \label{table:01}%
+  }{%
+  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
+  \toprule
+  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
+  \\
+  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
+  \\
+  \colrule
+  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
+  \\
+  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
+  \\
+  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
+  \\
+  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
+  \\
+  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
+  \\
+  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
+  \\
+  \citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+  \\
+  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
+  \\
+  \botrule
+  \end{tabular*}%
+  }
+  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
+  \end{table*}
 \end{document}

### 2019-08-15 NAR_DRAFT.tex
small changes

@@
 
+\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
+The first is the set of mutations that pass all heuristics filters, and a very large posterior probability of being present given the observed read data.
+We think of these high-probability variants as data generated from a distribution, the mutation profile, with very low measurement error.
+This low measurement error data allows us to reliably estimate the distribution of substitution types.
+Variants with low probability given the read data can be thought of as noisy measurements of samples from the same data generating distribution.
+We then use the estimated distribution of substitution types to sharpen the prior probability of mutation for these lower probability variants.
+
+The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
+\bkmcomment{coming}
+
+The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
+Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors.
+Deep sequencing is useful \cite{Zehir2017}.
+We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
+
+One size does not fit all with respect to the evidence threshold.
+Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
+Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
+One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
+
 \begin{itemize}
-  \item Every tumor is different. Differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}
-  \item The solution must be specific to each tumor we examine
   \item The fact that we can think of tumor variant calls as two sets of data, with different measurement error, derived from the same data generating process, makes this possible
@@
   \item We have devoloped this algorithm as a post-calling evaluation tool to be used at the end of a variant calling pipeline
-  \item The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
-  \item The benefit from the algorithm varies with both the concentration of the signature and the depth of sequencing
-  \item Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors
-  \item Deep sequencing is useful \cite{Zehir2017}
-  \item We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
-  \item One size does not fit all with respect to the evidence threshold.
-Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
-  \item Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
-  \item One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
+  \
   \item We believe the improved calibration achieved with this algorithm will encourage more thoughtful consideration of thresholds in various variant calling contexts

### 2019-08-16 NAR_DRAFT.tex
add calibration plots
add wgs figure
add figures

@@
 \subsection{The likelihood function}
-We designed \batcave to be used as a post-calling evaluation tool with MuTect
-MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
+We designed \batcave to be used as a post-calling evaluation tool with MuTect.
+MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
 We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
@@
 We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
+The algorithm is effecient, processing 22,000 variants per second on a typical iMac desktop computer, which corresponds to 100 seconds to process a 500X exome and 2000 seconds for a 100X whole genome.
 MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
@@
 \end{figure*}
+
 \subsection{Convergence of the site-specific prior to the data generating distribution}
@@
 In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for our real datasets are \bkmcomment{X} and for our simulated data sets are 1500 to 5000.
+For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.
 
@@
 These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/ppv_wes.pdf}
+  \end{center}
+  \caption{Model performance on 500X simulated whole exome. 
+  Positive predictive values vs. Recall for three different probability cutoffs. 
+  Note the restricted axes to improve readability of points.}
+\label{NAR-ppv_fig}
+\end{figure}
 
@@
 
-We measured calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
+We measured overall calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
 The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
-This is useful in our setting because the density of predicted probabilities tends to be strongly bi-modal.
-As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (top density curves), and decrease the probability of low probability but false variants (bottom density curves.).
+This is useful in our setting because the density of posterior probabilities tends to be strongly bi-modal.
+A large fraction of true negative variants have posterior probabilities of less than $10{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
+As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
 For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse.
@@
   \begin{center}
-  \includegraphics{figures/fig_wes.pdf}
+  \includegraphics{figures/fig_wgs.pdf}
   \end{center}
@@
   \end{table*}
+
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/ppv_wgs.pdf}
+  \end{center}
+  \caption{Model performance on 500X simulated whole exome. 
+  Positive predictive values vs. Recall for three different probability cutoffs. 
+  Note the restricted axes to improve readability of points.}
+\label{NAR-ppv_wgs_fig}
+\end{figure}
+
 \end{document}

### 2019-08-17 NAR_DRAFT.tex
check in change

--- 2019-08-15 NAR_DRAFT.tex
@@
 \subsection{Real Data}
-We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
-We downloaded the normal and primary tumor bam files generated for the whole genome sequences in \citet{Griffith2015} from dbGaP accession number phs000159.v8.p4.
-\citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
-The whole exome sequencing bam files generated by \citet{Shi2018} were downloaded from NCBI SRA no. SRP070662.
-\citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
-We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.
+\begin{itemize}
+  \item aml31 \cite{Griffith2015}
+  \item item used only the platinum calls for evaluation
+  \item Shi2018 \cite{Shi2018}
+  \item used the gold standard calls, and combined all six biological replicates 
+\end{itemize}
 
@@
 	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-wes_fig} \textbf{e}).
-	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.
+	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average area between the loess fit and a diagonal line.
 
@@
 
-The \batcave algorithm focuses on the tri-nucleotide context and substitution type. 
-There are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
-These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
+The \batcave algorithm focuses on the tri-nucleotide context and substitution type, but there are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
+These other types of context could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
 
@@
 
-\bibliography{caller_paper}
-
 \section{Supplementary Figures}
@@
 \end{figure*}
+
+\section{Supplememtary Table}
 
@@
   \end{table*}
+
+  \bibliography{caller_paper}
 \end{document}

### 2019-08-17 NAR_DRAFT.tex
trying to add supplement. not easy

--- 2019-08-16 NAR_DRAFT.tex
@@
 
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/signature_figure.pdf}
+  \end{center}
+  \caption{Mutation profiles. 
+  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
+  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
+  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
+  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
+  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
+  }
+  \label{NAR-sigfig}
+  \end{figure}
+
 \section{MATERIALS AND METHODS}
@@
 
-\begin{figure}[t]
-  \begin{center}
-  \includegraphics{figures/signature_figure.pdf}
-  \end{center}
-  \caption{Mutation profiles. 
-  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
-  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
-  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
-  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
-  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
-  }
-  \label{NAR-sigfig}
-  \end{figure}
-
 The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{m,c}$.
@@
 We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
+
+add to change
 
@@
 
-\section{RESULTS}
-We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-The algorithm is effecient, processing 22,000 variants per second on a typical iMac desktop computer, which corresponds to 100 seconds to process a 500X exome and 2000 seconds for a 100X whole genome.
-MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
-\batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
-The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
   \begin{figure*}
@@
 
+\section{RESULTS}
+We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
+The algorithm is effecient, processing 22,000 variants per second on a typical iMac desktop computer, which corresponds to 100 seconds to process a 500X exome and 2000 seconds for a 100X whole genome.
+MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
+\batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
+The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
+
 \subsection{Convergence of the site-specific prior to the data generating distribution}
@@
 These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
-
-\begin{figure}
-  \begin{center}
-  \includegraphics{figures/ppv_wes.pdf}
-  \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
-  Note the restricted axes to improve readability of points.}
-\label{NAR-ppv_fig}
-\end{figure}
 
@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/ppv_wes.pdf}
+  \end{center}
+  \caption{Model performance on 500X simulated whole exome. 
+  Positive predictive values vs. Recall for three different probability cutoffs. 
+  Note the restricted axes to improve readability of points.\bkmcomment{how do called allele frequency distributions differ?}}
+\label{NAR-ppv_fig}
+\end{figure}
 
@@
 
-\section{Supplementary Figures}
-
-% \begin{figure*}[t]
-%   \begin{center}
-%   \includegraphics{figures/aml_plot.pdf}
-%   \end{center}
-%   \caption{Model performance on 300X Whole leukemia genome.
-%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
-%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
-%   A perfectly calibrated model would generate the diagonal line.}
-%   \label{NAR-aml}
-% \end{figure*}
-
-\begin{figure*}
-  \begin{center}
-  \includegraphics{figures/fig_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 100X simulated whole genome.
-  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
-  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
-\label{NAR-wgs_fig}
-\end{figure*}
-
-\begin{table*}[b]
-  \tableparts{%
-  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
-  \label{table:01}%
-  }{%
-  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
-  \toprule
-  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
-  \\
-  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
-  \\
-  \colrule
-  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
-  \\
-  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
-  \\
-  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
-  \\
-  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
-  \\
-  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
-  \\
-  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
-  \\
-  \citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
-  \\
-  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
-  \\
-  \botrule
-  \end{tabular*}%
-  }
-  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
-  \end{table*}
-
-\begin{figure}
-  \begin{center}
-  \includegraphics{figures/ppv_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
-  Note the restricted axes to improve readability of points.}
-\label{NAR-ppv_wgs_fig}
-\end{figure}
+\include{supplement}
 

### 2019-08-17 supplement.tex
trying to add supplement. not easy

\section{Supplementary Figures}

% \begin{figure*}[t]
%   \begin{center}
%   \includegraphics{figures/aml_plot.pdf}
%   \end{center}
%   \caption{Model performance on 300X Whole leukemia genome.
%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
%   A perfectly calibrated model would generate the diagonal line.}
%   \label{NAR-aml}
% \end{figure*}

\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wgs.pdf}
  \end{center}
  \caption{Model performance on 100X simulated whole genome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the BATCAVE model.
  A perfectly calibrated model would generate the diagonal line.}
\label{NAR-wgs_fig}
\end{figure*}

\begin{figure}
  \begin{center}
  \includegraphics{figures/ppv_wgs.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points.}
\label{NAR-ppv_wgs_fig}
\end{figure}

\section{Supplememtary Table}

\begin{table*}[b]
  \tableparts{%
  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
  \label{table:01}%
  }{%
  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
  \toprule
  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
  \\
  & & (estimated) & MuTect & BATCAVE & MuTect & BATCAVE & MuTect & BATCAVE
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  Williams & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  Shi & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular*}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
  \end{table*}

### 2019-08-18 NAR_DRAFT.tex
wrangled a supplement

--- 2019-08-16 NAR_DRAFT.tex
@@
 \bibliography{caller_paper}
+\clearpage
+\onecolumn
+\section{Supplememtary Figures}
 
-\section{Supplementary Figures}
-
-% \begin{figure*}[t]
-%   \begin{center}
-%   \includegraphics{figures/aml_plot.pdf}
-%   \end{center}
-%   \caption{Model performance on 300X Whole leukemia genome.
-%   \textbf{(a)} Precision/recall curves and \textbf{(c)} Reciever operating characteristic curves.
-%   Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the Calibrated model.
-%   A perfectly calibrated model would generate the diagonal line.}
-%   \label{NAR-aml}
-% \end{figure*}
-
-\begin{figure*}
+\begin{figure}[b]
   \begin{center}
@@
   \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
-  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
+  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the BATCAVE model.
+  A perfectly calibrated model would generate the diagonal line.}
 \label{NAR-wgs_fig}
-\end{figure*}
+\end{figure}
 
-\begin{table*}[b]
+\clearpage
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/ppv_wgs.pdf}
+  \end{center}
+  \caption{Model performance on 500X simulated whole exome. 
+  Positive predictive values vs. Recall for three different probability cutoffs. 
+  Note the restricted axes to improve readability of points.}
+\label{NAR-ppv_wgs_fig}
+\end{figure}
+
+\section{Supplememtary Table}
+
+\begin{table}
   \tableparts{%
@@
   \\
-  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
+  & & (estimated) & MuTect & BATCAVE & MuTect & BATCAVE & MuTect & BATCAVE
   \\
@@
   \\
-  \citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+  Williams & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
   \\
-  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
+  Shi & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
   \\
@@
   {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
-  \end{table*}
-
-\begin{figure}
-  \begin{center}
-  \includegraphics{figures/ppv_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
-  Note the restricted axes to improve readability of points.}
-\label{NAR-ppv_wgs_fig}
-\end{figure}
+ \end{table}
 

### 2019-08-18 supplement.tex
wrangled a supplement

@@
 % \end{figure*}
-
-\begin{figure*}
-  \begin{center}
-  \includegraphics{figures/fig_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 100X simulated whole genome.
-  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the BATCAVE model.
-  A perfectly calibrated model would generate the diagonal line.}
-\label{NAR-wgs_fig}
-\end{figure*}
-
-\begin{figure}
-  \begin{center}
-  \includegraphics{figures/ppv_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
-  Note the restricted axes to improve readability of points.}
-\label{NAR-ppv_wgs_fig}
-\end{figure}
-
-\section{Supplememtary Table}
-
-\begin{table*}[b]
-  \tableparts{%
-  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
-  \label{table:01}%
-  }{%
-  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
-  \toprule
-  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
-  \\
-  & & (estimated) & MuTect & BATCAVE & MuTect & BATCAVE & MuTect & BATCAVE
-  \\
-  \colrule
-  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
-  \\
-  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
-  \\
-  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
-  \\
-  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
-  \\
-  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
-  \\
-  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
-  \\
-  Williams & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
-  \\
-  Shi & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
-  \\
-  \botrule
-  \end{tabular*}%
-  }
-  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
-  \end{table*}

### 2019-08-19 NAR_DRAFT.tex
small fixes

@@
 \newcommand{\batcave}{BATCAVE }
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
 % Enter dates of publication
@@
 \end{equation}
-Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $p(m)$.
+Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.
 
@@
 
-	We use the Integrated Calibration Index to quantify the difference between MuTect and the Calibrated model \cite{Austin2019}.
+	We use the Integrated Calibration Index to quantify the difference between MuTect and \batcave \cite{Austin2019}.
 	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-wes_fig} \textbf{e}).
@@
 \onecolumn
+\beginsupplement
 \section{Supplememtary Figures}

### 2019-08-19 NAR_DRAFT.tex
fix a bunch of mathrm mistakes and some text

@@
 
-Single nucleotide substitions, i.e. simple mutations, arise in tumors at a rate and at genomic locations driven by two main processes. 
+Single nucleotide variants (SNV) arise in tumors at a rate and at genomic locations driven by two main processes. 
 The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
@@
   \begin{array}{rcl}
-  \mathrm{p}(\mathrm{S_{m,c}}) &=&  \mathrm{p}(m, M | C = c) \\
+  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                             &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{p(M)}{p(C = c)},
@@
 \end{equation}
-where $\mathrm{p}(\mathrm{S_{m,c}})$ is the probability that a substitution of type $\mathrm{S_{m,c}}$ will occur.
+where $\mathrm{p}(\mathrm{S}_{m,c})$ is the probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
 
@@
 We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
-
-add to change
+The R implementation for \batcave ships with pre-computed tables for both whole exomes and whole genomes.
 
 \subsection{Estimation of the mutation rate.}
-The final piece of equation \ref{eqn:detailed_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
+The final piece of equation \ref{eqn:detailed_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which we specify as the per-base mutation rate $\mu$.
 In an exponentially growing and neutrally or nearly neutrally evolving tumor, \citet{Williams2018} show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
@@
 We designed \batcave to be used as a post-calling evaluation tool with MuTect.
-MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
+MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in equation \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
 We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
@@
   $$
-  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S_{m,c}}))},
+  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))},
   $$
 \end{equation}
-where $\mathrm{p}(\mathrm{S_{m,c}})$ is the prior probability of a substitution of type $\mathrm{S_{m,c}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
+where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
 When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
@@
 
+\bkmcomment{Discuss assumption that vaf in uniform. opportunity for further work.}
+
 One size does not fit all with respect to the evidence threshold.

### 2019-08-19 NAR_DRAFT.tex
small fixes

@@
 The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
-Variants were selected from a combination of Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
+Variants were selected from a combination of the Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
 Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (RD figure coming and Table \ref{table:01}).
+The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a)} and Table \ref{table:01}).
 
@@
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave.
+We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{b)} and Table \ref{table:01}).
 

### 2019-08-19 NAR_DRAFT.tex
finish table S1
better cosmic language

@@
 We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
-Tumors were simulated with three different mutation profiles, all of which include a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
+Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (\bkmcomment{website or citation}).
+Each simulate profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
 The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
@@
   \colrule
-  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
+  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
   \\
-  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
+  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & 214
   \\
-  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
+  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
   \\

### 2019-08-19 NAR_DRAFT.tex
fixing little things

@@
 	We use the Integrated Calibration Index to quantify the difference between MuTect and \batcave \cite{Austin2019}.
-	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and the Calibrated model (Figure \ref{NAR-wes_fig} \textbf{e}).
+	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and \batcave (Figure \ref{NAR-wgs_fig} \textbf{e} and \textbf{f} and Table \ref{table:01}).
 	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.
@@
 As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
-For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse.
+For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
 This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.

### 2019-08-19 NAR_DRAFT.tex
minor changes
add figure 3 to results description

@@
 \newcommand{\beginsupplement}{%
+        \clearpage
+        \onecolumn
         \setcounter{table}{0}
@@
 
+Figure \ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration in the portion of the posterior probability distribution relevant to clinicians and researchers.
+For any probability between .6 and .95 \batcave is better calibrated than Mutect.
+Essentially, for any posterior probability above .7 MuTect has a false positive rate around 8\%, which \batcave has a false positive rate that increases as posterior probability decreases.
+The cost of this compression is recall.
+At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
+
 \subsection{Real tumor data.}
@@
 The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
-\bkmcomment{coming}
 
@@
 \bibliography{caller_paper}
-\clearpage
-\onecolumn
+
 \beginsupplement

### 2019-08-19 NAR_DRAFT.tex
new figures without blue, and new results text

@@
 Essentially, for any posterior probability above .7 MuTect has a false positive rate around 8\%, which \batcave has a false positive rate that increases as posterior probability decreases.
-The cost of this compression is recall.
+The cost of this compression is recall, which is 3-5\% better than MuTect at 95\% posterior probability, and 3-7\% better at 90\%
 At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a)} and Table \ref{table:01}).
+The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a} and Table \ref{table:01}).
 
@@
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{b)} and Table \ref{table:01}).
+We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{b} and Table \ref{table:01}).
 
@@
   Positive predictive values vs. Recall for three different probability cutoffs. 
-  Note the restricted axes to improve readability of points.\bkmcomment{how do called allele frequency distributions differ?}}
+  Note the restricted axes to improve readability of points}
 \label{NAR-ppv_fig}

### 2019-08-19 NAR_DRAFT.tex
better

@@
 Figure \ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration in the portion of the posterior probability distribution relevant to clinicians and researchers.
-For any probability between .6 and .95 \batcave is better calibrated than Mutect.
-Essentially, for any posterior probability above .7 MuTect has a false positive rate around 8\%, which \batcave has a false positive rate that increases as posterior probability decreases.
-The cost of this compression is recall, which is 3-5\% better than MuTect at 95\% posterior probability, and 3-7\% better at 90\%
+For any probability between 60 and 95\% \batcave is better calibrated than Mutect.
+Essentially, for any posterior probability above 70\% MuTect has a false positive rate around 8\%, while \batcave has a false positive rate that increases as posterior probability decreases.
+The cost of this compression is recall, which is 3-5\% better than MuTect at 95\% posterior probability, and 3-7\% better at 90\%, depending on muation profile concentration.
 At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.

### 2019-08-19 NAR_DRAFT.tex
fix captions for supplement figures

@@
   \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the BATCAVE model.
+  Distribution of estimated true positive probabilities for true positive  variants for \textbf{b)} the MuTect model and \textbf{d)} the BATCAVE model.
   A perfectly calibrated model would generate the diagonal line.}
@@
   \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
+  \caption{Model performance on 100X simulated whole genome. 
   Positive predictive values vs. Recall for three different probability cutoffs. 

### 2019-08-19 NAR_DRAFT.tex
fix the table again

@@
   \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive  variants for \textbf{b)} the MuTect model and \textbf{d)} the BATCAVE model.
+  Distribution of estimated true positive probabilities for true positive  variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
   A perfectly calibrated model would generate the diagonal line.}
@@
   \tableparts{%
-  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
+  \caption{Performance characteristics as measured for all data sets.}
   \label{table:01}%
@@
   \\
-  & & (estimated) & MuTect & BATCAVE & MuTect & BATCAVE & MuTect & BATCAVE
+  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
   \\
@@
   \\
-  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & 214
+  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
   \\
@@
   \\
-  Williams & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+  \citet{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
   \\
-  Shi & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
+  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
   \\

### 2019-08-19 NAR_DRAFT.tex
minor text changes

@@
 \batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
-The first is the set of mutations that pass all heuristics filters, and a very large posterior probability of being present given the observed read data.
+The first is the set of mutations that pass all heuristics filters, and a having a very large posterior probability of being present given the observed read data.
 We think of these high-probability variants as data generated from a distribution, the mutation profile, with very low measurement error.

### 2019-08-30 NAR_DRAFT.tex
check in

@@
 
-At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
+At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector \bkmcomment{push equation onto separate line} $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
 We want to use $\mathbf{x}$ to select between competing hypotheses;
@@
 $$
-  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
+  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
 $$
@@
   \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
-                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{p(M)}{p(C = c)},
+                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)},
   \end{array}
@@
 We make the simplifying assumption that there is no cell death, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
-Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
+Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
 
@@
 
-\begin{table}
+\begin{table}[b]
   \tableparts{%

### 2019-08-30 NAR_DRAFT.tex
fix simulate

@@
 Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (\bkmcomment{website or citation}).
-Each simulate profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
+Each simulated profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
 The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.

### 2019-09-03 NAR_DRAFT.tex
slow going

@@
 We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
-However, in principle there is no barrier to using the method with other variant callers.
+However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
 For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
 Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
-In general we are advancing the idea that there is substantial prior information about the tumor and site-specific mutation probability encoded in NGS data, and we should be using it.
+The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
+Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.
 
@@
 
-The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
-
 The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
@@
 We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
-
-\bkmcomment{Discuss assumption that vaf in uniform. opportunity for further work.}
 

### 2019-09-03 NAR_DRAFT.tex
commit to merge

--- 2019-08-30 NAR_DRAFT.tex
@@
 
-We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
+We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple statistical model and a well-tested set of heuristic filters.
 However, in principle there is no barrier to using the method with other variant callers.
@@
 Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
-In general we are advancing the idea that there is substantial prior information about the tumor and site-specific mutation probability encoded in NGS data, and we should be using it.
+The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
 
 \batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
-The first is the set of mutations that pass all heuristics filters, and a having a very large posterior probability of being present given the observed read data.
-We think of these high-probability variants as data generated from a distribution, the mutation profile, with very low measurement error.
-This low measurement error data allows us to reliably estimate the distribution of substitution types.
-Variants with low probability given the read data can be thought of as noisy measurements of samples from the same data generating distribution.
-We then use the estimated distribution of substitution types to sharpen the prior probability of mutation for these lower probability variants.
-
-The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
+The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
+These are high-probability variants generated from a distribution, the mutation profile, and measured with very small error.
+This data allows us to reliably estimate the distribution of substitution types.
+Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
+We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.
 

### 2019-09-03 NAR_DRAFT.tex

@@
 
-We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple statistical model and a well-tested set of heuristic filters.
-However, in principle there is no barrier to using the method with other variant callers.
+We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
+However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
 For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
 Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
-The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.
+The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
+Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.
 

### 2019-09-03 NAR_DRAFT.tex
working on discussion

@@
 We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
+
 We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
@@
 
-The \batcave algorithm focuses on the tri-nucleotide context and substitution type. 
-There are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
+The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
+There are, however, many other types of \textit{genomic context} which can effect local mutation rates such as replication timing, expression level, and histone occupancy \cite{Buisson2019} .
 These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
@@
 The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
-These are high-probability variants generated from a distribution, the mutation profile, and measured with very small error.
+These high-probability variants are generated from the mutation profile of the tumor, and measured with very small error.
 This data allows us to reliably estimate the distribution of substitution types.
@@
 One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
-
-\begin{itemize}
-  \item The fact that we can think of tumor variant calls as two sets of data, with different measurement error, derived from the same data generating process, makes this possible
-  \item The mutation profile is estimated from the high-confidence (low measurement error) variants, and used as a prior for the low-confidence (high measurement error) variants
-  \item We demonstrate that the mutation generating process (mutation profile) can be reliably retrieved from simulated NGS data.
-  \item We also demonstrate that on the limited real data available, which consists of validated high confidence mutations, the results are equivalent
-  \item What is needed is validation experiments that also allow for the estimation of true and false negative rates, in order to fully illuminate the utility of \batcave
-  \item We have devoloped this algorithm as a post-calling evaluation tool to be used at the end of a variant calling pipeline
-  \
-  \item We believe the improved calibration achieved with this algorithm will encourage more thoughtful consideration of thresholds in various variant calling contexts
-  \item The field believes that little information can be gained by wes rather than panels, but they are assuming the only information is mutations \cite{Hyman2015}
-\end{itemize}
 

### 2019-09-03 NAR_DRAFT.tex
new papers and citations

@@
 The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
-There are, however, many other types of \textit{genomic context} which can effect local mutation rates such as replication timing, expression level, and histone occupancy \cite{Buisson2019} .
+There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010} .
 These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

### 2019-09-03 NAR_DRAFT.tex
reorganize discussion
move a bit to conclusion

@@
 
-We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
-These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
-At the same time, there is no way to compute true and false negative fractions for these data sets.
-As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
-Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
-
 The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
@@
 Deep sequencing is useful \cite{Zehir2017}.
+
+We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
+These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
+At the same time, there is no way to compute true and false negative fractions for these data sets.
+As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
+Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+
+\section{CONCLUSION}
+
 We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
-
 One size does not fit all with respect to the evidence threshold.
@@
 One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
-
-\section{CONCLUSION}
-
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text.
 

### 2019-09-03 NAR_DRAFT.tex
add more citations

@@
 
-We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
+We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent \cite{Griffith2015,Shi2018}.
 These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.

### 2019-09-04 NAR_DRAFT.tex
lunch changes

@@
 \section{DISCUSSION}
+We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
+One size does not fit all with respect to the evidence threshold.
+Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
+Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
+One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
+The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
+Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors.
+Deep sequencing is useful \cite{Zehir2017}.
+\bkmcomment{the above should be some version of, variant calling is important, deeper is better, we are taking a step in that direction}
 
@@
 We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
+\bkmcomment{deeper summary of results, and some meaning Somithing about calibration dummy}.
 
 The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
-There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010} .
+There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
+\bkmcomment{what to say here? Are we acknowledging a weakness, or just observing there are other types?}
 These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
 
+\bkmcomment{summarize run-time here}
 We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
@@
 
+\bkmcomment{can I work a citation in here? using data sets from the same data generating distribution as a prior?}
 \batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
 The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
-These high-probability variants are generated from the mutation profile of the tumor, and measured with very small error.
-This data allows us to reliably estimate the distribution of substitution types.
+This data allows us to reliably estimate the distribution of substitution types, i.e. the mutation profile.
 Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
 We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.
-
-The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
-Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors.
-Deep sequencing is useful \cite{Zehir2017}.
 
@@
 One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
+
+\section{Software Availability}
+github for source code
+zenodo for figures, simulations
+zenodo for package at time of bioarxiv
 

### 2019-09-04 NAR_DRAFT.tex
flailing

@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a} and Table \ref{table:01}).
+The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-real_data_fig} \textbf{a} and Table \ref{table:01}).
 
@@
 \section{DISCUSSION}
+
 We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
@@
 
+\begin{figure}[b]
+  \begin{center}
+  \includegraphics{figures/aml_plot.pdf}
+  \end{center}
+  \caption{aml and cell. Need no color and drop calibration}
+\label{NAR-real_data_fig}
+\end{figure}
+
 \clearpage

### 2019-09-05 NAR_DRAFT.tex
check in discussion

@@
 
-Figure \ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration in the portion of the posterior probability distribution relevant to clinicians and researchers.
-For any probability between 60 and 95\% \batcave is better calibrated than Mutect.
+Figure \ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration for 500X whole exomes in the portion of the posterior probability distribution relevant to clinicians and researchers.
+For any probability between 60 and 90\% \batcave is better calibrated than Mutect.
 Essentially, for any posterior probability above 70\% MuTect has a false positive rate around 8\%, while \batcave has a false positive rate that increases as posterior probability decreases.
@@
 At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
+Figure \ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
+Note that recall never improves because the uncalled true positive variants are below the threshold of detection for any reasonable level of false positive calls.
 
@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-real_data_fig} \textbf{a} and Table \ref{table:01}).
+The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{b} and Table \ref{table:01}).
 
@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{b} and Table \ref{table:01}).
+
+We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent \cite{Griffith2015,Shi2018}.
+These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
+At the same time, there is no way to compute true and false negative fractions for these data sets.
+As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
 
@@
 
-We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
-One size does not fit all with respect to the evidence threshold.
-Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
-Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
-One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
-The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
-Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors.
-Deep sequencing is useful \cite{Zehir2017}.
-\bkmcomment{the above should be some version of, variant calling is important, deeper is better, we are taking a step in that direction}
+Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \cite{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.
+As \citet{Griffith2015} show, whole genome sequencing at approx.
+350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing.
+We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
 
 \batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
-We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
-\bkmcomment{deeper summary of results, and some meaning Somithing about calibration dummy}.
+We demonstrated here that \batcave improves classification accuracy in synthetic data, leading to better more accurate variant calls at any posterior probability threshold.
+Importantly, \batcave is better calibrated than MuTect at relevant probability thresholds, allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
+We are unable to study calibration in real data, because complete false positive/true negative data is not available in currently available for real data
+Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+
+\bkmcomment{can I work a citation in here? using data sets from the same data generating distribution as a prior? Struggling to find one, but will keep looking.}
+\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
+The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
+This data allows us to reliably estimate the distribution of substitution types, i.e. the mutation profile.
+Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
+We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.
 
@@
 
-\bkmcomment{summarize run-time here}
 We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
@@
 Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
+The R implementation \texttt{batcaver} adds 1 second per 22,000 variants evaluated to a standard GATK best practices variant calling pipeline. 
 The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
@@
 
-\bkmcomment{can I work a citation in here? using data sets from the same data generating distribution as a prior?}
-\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
-The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
-This data allows us to reliably estimate the distribution of substitution types, i.e. the mutation profile.
-Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
-We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.
-
-We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent \cite{Griffith2015,Shi2018}.
-These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
-At the same time, there is no way to compute true and false negative fractions for these data sets.
-As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
-Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications.
+One size does not fit all with respect to the evidence threshold. 
+Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of precision and recall.
+\batcave is designed to present the user with a more accurate sense of these trade-offs.
 
 \section{CONCLUSION}
-
-We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
-One size does not fit all with respect to the evidence threshold.
-Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
-Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
-One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
+\bkmcomment{I looked at a bunch of NAR methods papers, and none include a conclusions section, so I moved it to the last paragraph of discussion.}
 
 \section{Software Availability}
+\bkmcomment{working on this today}
 github for source code
@@
 \label{NAR-wgs_fig}
-\end{figure}
-
-\begin{figure}[b]
-  \begin{center}
-  \includegraphics{figures/aml_plot.pdf}
-  \end{center}
-  \caption{aml and cell. Need no color and drop calibration}
-\label{NAR-real_data_fig}
 \end{figure}

### 2019-09-05 NAR_DRAFT.tex
minor change
grammar

@@
 Importantly, \batcave is better calibrated than MuTect at relevant probability thresholds, allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
-We are unable to study calibration in real data, because complete false positive/true negative data is not available in currently available for real data
+We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
 Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
@@
 Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of precision and recall.
-\batcave is designed to present the user with a more accurate sense of these trade-offs.
+We present \batcave, and the R package \texttt{batcaver} to provide the user with a more accurate sense of these trade-offs.
 

### 2019-09-06 NAR_DRAFT.tex
paragraph
figure changes

@@
   \begin{center}
-  \includegraphics{figures/signature_figure.pdf}
+  \includegraphics{figures/signatures_only.pdf}
   \end{center}
@@
   \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
-  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
-  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
   }
@@
 where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
-When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
-They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
+When comparing our posterior odds to those of MuTect in this paper, we use \citet{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
 We compute the posterior odds for Mutect as
@@
 \end{equation}
-since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.
+since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$.
 
@@
 The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/kl_only.pdf}
+  \end{center}
+  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
+  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
+\label{NAR-ppv_fig}
+\end{figure}
 
@@
 Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
+
 The R implementation \texttt{batcaver} adds 1 second per 22,000 variants evaluated to a standard GATK best practices variant calling pipeline. 

### 2019-09-06 NAR_DRAFT.tex
mutect stuff

@@
 where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
-When comparing our posterior odds to those of MuTect in this paper, we use \citet{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
+When comparing our posterior odds to those of MuTect in this paper, we following \citet{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
 We compute the posterior odds for Mutect as

### 2019-09-06 NAR_DRAFT.tex
get rid of useless color

@@
   The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
-\label{NAR-ppv_fig}
+\label{NAR-kl_fig}
 \end{figure}
@@
 We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+In Figure \ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
 For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.

### 2019-09-09 NAR_DRAFT.tex
add note comment
numbered citations
changes to discussion

@@
 \usepackage{NAR-natbib}
-\bibliographystyle{unsrtnat}
+\bibliographystyle{NAR-natbib}
 \usepackage{color}
@@
 
-\bkmcomment{can I work a citation in here? using data sets from the same data generating distribution as a prior? Struggling to find one, but will keep looking.}
-\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
-The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
-This data allows us to reliably estimate the distribution of substitution types, i.e. the mutation profile.
-Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
-We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.
+\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
+In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
+Early mutations are classified with low error, while later mutations are classified with less certainty.
+We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
+Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific that sharpens the classification and calibration of later mutations.
+Thus, \batcave is best described as an algorithm trained on early mutations.
+
+\bkmcomment{this needs more work}
+We simulate tumors with three mutation profiles, and evaluate two real tumor data sets.
+Among these performance is best with the most concentrated simulatin profile.
+The intermediate and diffuse mutation profiles are designed to represent baseline or minimal profiles of Lung and Breast tumors respectively.
+In fact, lung and breast tumors have heterogeneous mutation profiles with a wide range of levels of concentration.
+The Acute Myeloid Leukemia we evaluated has a particularly diffuse profile, which is common for these tumors \cite{Alexandrov2019}.
 
@@
 
-\section{CONCLUSION}
-\bkmcomment{I looked at a bunch of NAR methods papers, and none include a conclusions section, so I moved it to the last paragraph of discussion.}
+% \section{CONCLUSION}
+% \bkmcomment{I looked at a bunch of NAR methods papers, and none include a conclusions section, so I moved it to the last paragraph of discussion.}
 
@@
 \bkmcomment{working on this today}
+
 github for source code
+
 zenodo for figures, simulations
+
 zenodo for package at time of bioarxiv
@@
 
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text.
+This work was supported by the National Science Foundation via Graduate Research Fellowship DGE-1143953 to BKM and \bkmcomment{RNG NIH}.
+The authors would like to thank Professor Edward J. Bedrick for fruitful discussions about the statistical model.
 

### 2019-09-10 NAR_DRAFT.tex
edits
figure 2 caption

@@
 \citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
-The whole exome sequencing bam files generated by \citet{Shi2018} were downloaded from NCBI SRA no. SRP070662.
+The whole exome sequencing bam files generated by \citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
 \citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
@@
     \end{center}
-    \caption{Model performance on 500X simulated whole exome.
-    \textbf{(a)} Precision recall curves and \textbf{(c)} Receiver operating characteristic curves for 3 mutation signatures.
-    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
-    A perfectly calibrated model would generate the diagonal line.}
+    \caption{Model performance.
+    \textbf{(a)} Precision-recall curves and \textbf{(b)} Receiver operating characteristic curves for 3 mutation signatures.
+    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(c)} the MuTect model and \textbf{(d))} the \batcave model.
+    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(e)} and \textbf{(f)}}
   \label{NAR-wes_fig}

### 2019-09-10 NAR_DRAFT.tex
note to fix figure 2
remove figure note
incredible struggle to get numeric and inline. back to author-year for now. looks to be OK for submission

@@
 \documentclass[a4,center,fleqn]{NAR}
-
-\usepackage{NAR-natbib}
-\bibliographystyle{NAR-natbib}
+\bibliographystyle{plainnat}
 \usepackage{color}
 \usepackage{xfrac}
+
 \newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
@@
 Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers \citep{Garrison2012},
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\citep{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)\citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
+joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)\citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
 All of these methods have varying levels of complexity, and some are focused on specific types of data.
@@
     Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(c)} the MuTect model and \textbf{(d))} the \batcave model.
-    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(e)} and \textbf{(f)}}
+    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(e)} and \textbf{(f)}.}
   \label{NAR-wes_fig}

### 2019-09-10 NAR_DRAFT.tex
add r package name and remove comment

@@
 Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
-We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
-\bkmcomment{batcave,batcaver,batcaveR}
+We provide a simple implementation in R, \texttt{batcaver} that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
 
@@
 \citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
-The whole exome sequencing bam files generated by \citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
+The whole exome sequencing bam files generated by \citet{Shi2018} were downloaded from NCBI SRA no. SRP070662.
 \citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.

### 2019-09-12 NAR_DRAFT.tex
starting on abstract
abstract

@@
 \begin{abstract}
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
-Text. Text.
+The accumulation of somatic mutations is an inevitable consequence of tumor development.
+Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
+Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in resistance, and the value of tumor heterogeneity as a biomarker.
+The identification of low frequency mutations from Next Generation Sequencing data remains challenging.
+In general the ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
+This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
+Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
+\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
+We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
+We provide an R package implementing the \batcave algorithm for use with MuTect output.
 \end{abstract}
@@
 
-At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector \bkmcomment{push equation onto separate line} $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
+At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
 We want to use $\mathbf{x}$ to select between competing hypotheses;

### 2019-09-12 NAR_DRAFT.tex
minor additions to abstract

@@
 Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
-Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in resistance, and the value of tumor heterogeneity as a biomarker.
+Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
 The identification of low frequency mutations from Next Generation Sequencing data remains challenging.

### 2019-09-12 NAR_DRAFT.tex
Initial comments from Ryan

@@
 In general the ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
+\rngcomment{A little too much intro in abstract.}
 This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
@@
 We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
+\rngcomment{Need a sentence about what new analysis this enables. How does \batcave (at least help) solve the problem you raised earlier?}
 We provide an R package implementing the \batcave algorithm for use with MuTect output.
@@
 
-Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
-Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
-Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
-Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this problem increases as sequencing depth increases.
+Cancer develops through the accumulation of somatic mutations and clonal selection of cells with mutations that confer an advantage.
+Understanding the evolutionary history of a tumor, including the mutations that drive its growth, the genetic diversity within it, and the accumulation of new mutations, requires accurate variant identification, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
+Accurate variant calling is also critical for optimizing the treatment of individual patients' disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
+Low frequency mutations challenge current variant calling methods, because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this challenge increases with sequencing depth.
 
-Methods for identifying true somatic mutations - i.e. variant calling -  from NGS data are an active area of research in bioinformatics.
-The earliest widely used somatic variant callers aimed specifically at tumors, Mutect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants, setting a threshold for that score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
-Subsequent research gave rise to a number of alternate variant calling strategies including haplotype based callers \citep{Garrison2012},
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone)\cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE)\citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and a mixture of ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC).
-All of these methods have varying levels of complexity, and some are focused on specific types of data.
-The one thing they all have in common is that they either implicitly or explicitly assume that the probability of a mutation occuring at a particular site is proportional to the overall mutation rate, and the same at every site in the genome.
+Many methods have been developed for calling somatic mutations from NGS data.
+The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
+Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
+joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
+These methods vary in their complexity and specific focus.
+But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.
 
-Single nucleotide variants (SNV) arise in tumors at a rate and at genomic locations driven by two main processes. 
-The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The second -- and far more complex -- process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
+The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
+%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
+Somatic mutations accumulate spontaneously in all dividing tissues, and the local rate of mutation depends sensitively on the nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
+Tumor cells additionally acquire mutations through exposure to mutagens or degradation, via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
 Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
@@
 
-This work was supported by the National Science Foundation via Graduate Research Fellowship DGE-1143953 to BKM and \bkmcomment{RNG NIH}.
-The authors would like to thank Professor Edward J. Bedrick for fruitful discussions about the statistical model.
+This work was supported by the National Science Foundation via Graduate Research Fellowship award number DGE-1143953 to BKM and by the National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM127348 to RNG.
+We thank Professor Edward J. Bedrick for fruitful discussions about the statistical model.
 

### 2019-09-12 NAR_DRAFT.tex
More pass by Ryan.

@@
 %Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
-Somatic mutations accumulate spontaneously in all dividing tissues, and the local rate of mutation depends sensitively on the nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
-Tumor cells additionally acquire mutations through exposure to mutagens or degradation, via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors.
-Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
-The specific processes driving mutations in a particular tumor generate a tumor's unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}
-Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
+The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
+Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
+The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
+For example, Fig. \ref{NAR-sigfig} \textbf{a-c}, shows mutation profiles from several tumors.
+\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
+%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
 
-Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
-We provide a simple implementation in R, \texttt{batcaver} that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
+Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
+Our algorithm estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
+Our implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
+Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
+Our algorithm is simple, computationally inexpensive, and can be integrated into numerous other variant callers.
+Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.
 
@@
   \includegraphics{figures/signatures_only.pdf}
+
+  \rngcomment{Need more basic labels for each panel. E.g. ``Real breast tumor'', ``Real accumulated myeloid leukemia'', ``Simulated concentrated profile''}
   \end{center}
-  \caption{Mutation profiles. 
-  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
-  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
+  \caption{Tumor mutation profiles.
+  \rngcomment{Add explanation of how to read a profile. ``In each panel, the histograms show...''}
+  \textbf{(a)} The observed mutation profile of a breast tumor \cite{Shi2018}.
+  \textbf{(b)} The observed mutation profile of an acute myeloid leukemia \cite{Griffith2015}.
+  \textbf{(c)} \rngcomment{Add a third (and possibly fourth) real tumor, with a more concentrated profile.
+So we can illustrate that our simulations aren't out of line with reality.
+Ideally two tumors of same type that have noticeably different profiles.}
   \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
@@
 \section{MATERIALS AND METHODS}
-\subsection{Somatic variant calling base probability model}
+\subsection{Somatic variant calling probability model}
 
-At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
-We want to use $\mathbf{x}$ to select between competing hypotheses;
+At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is local read depth.
+Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
 \begin{equation}
   \label{eqn:hypothesis}
-$$
   \begin{array}{l}
     \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
-    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
+    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f}.
   \end{array}
-$$
 \end{equation}
-where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e.
-the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
-The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
-Assuming that reads are independent, we have
+Here $m$ is any of the 3 possible alternate non-reference bases and $\nu$ is the variant allele frequency.
+The maximum likelihood estimate of $\nu$ is simply $\hat{f}$, the number of variant reads divided by the local read depth.
+The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given that hypothesis and the prior probability of that hypothesis. 
+Assuming that reads are independent, this is
+\begin{equation}\label{eqn:1}
+  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
+\end{equation}
+where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads.
 
-\begin{equation}
-  \label{eqn:1}
-$$
-  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
-$$
+Assuming that the identity of the alternate allele and its allele frequency are independent and that $\nu$ is uniformly distributed, Eq.~\ref{eqn:1} becomes
+\begin{equation}  \label{eqn:2}
+  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
 \end{equation}
-where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.
-
-Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e.
-$\mathrm{p}(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
-\begin{equation}
-  \label{eqn:2}
-$$
-  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
-$$
-\end{equation}
-Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.
+The focus of \batcave is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.
 
 \subsection{Site-specific prior probability of mutation}
-The probability which we have denoted $\mathrm{p}(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $\mathrm{p}(m,M)$.
-However, $\mathrm{p}(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
-Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as
-
+The probability that we have denoted $\mathrm{p}(m)$ in Eq.~\ref{eqn:2} is more precisely the joint probability that a mutation has occurred $M$ and that it was to allele $m$, which we denote $\mathrm{p}(m,M)$.
+But $\mathrm{p}(m,M)$ is not uniform across the genome.
+Rather it depends on the local genomic context $C$, so its full form is $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
+Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' theorem to further decompose as
 \begin{equation}
   \label{eqn:full_prior}
-  $$
   \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
-  $$
 \end{equation}
-We next show how to estimate the quantities in equation \ref{eqn:full_prior}.
+We next show how to estimate the quantities in Eq.~\ref{eqn:full_prior}.
 
 \subsection{Estimation of the mutation profile}
-There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
-Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The tri-nucleotide context at a particular genomic site is made up of the identity of the reference base and the identity of the 3` and 5` flanking bases.
-There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
-Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
-We define a mutation context $C$ indexed by $c=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
-For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
-We can thus define $\mathrm{S}$, the substitution type, indexed by $c=\{1 \dots 32\}$ contexts and $m = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
-Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as
-
+Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
+Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
+Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
+At the focal site, a mutation $m$ can be to any of three alternate alleles.
+Indexing by the $c=\{1 \dots 32\}$ contexts and and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
+Eq.~\ref{eqn:full_prior} is then
 \begin{equation}
   \label{eqn:detailed_prior}
-  $$
   \begin{array}{rcl}
   \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
-                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)},
+                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
   \end{array}
-  $$
 \end{equation}
-where $\mathrm{p}(\mathrm{S}_{m,c})$ is the probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
+The first two terms on the right-hand side can be estimated from the mutation profile (Fig.~\ref{NAR-sigfig}).
+%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
 

### 2019-09-13 NAR_DRAFT.tex
Edits by Ryan.

@@
 \end{equation}
-The first two terms on the right-hand side can be estimated from the mutation profile (Fig.~\ref{NAR-sigfig}).
+The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-sigfig}).
 %Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
 
-The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{m,c}$.
-The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is modeled as multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
-In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
-As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,
+\rngcomment{Got here...}
 
+We model the observed mutation profile $S$ (Fig.~\ref{NAR-sigfig}a-c) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
+In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
+But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
+We thus model $\mathrm{p}(\mathrm{S})$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
 $$
@@
 $$
-In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a-priori a mutation is equally likely in any context.
+In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.
 
-In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
-These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant.
-Let $\mathrm{D}$ be the set of high confidence variant calls -- which we define as having posterior odds greater than 10 to 1 -- and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
+To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
+These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
+Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 with the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
 The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where

### 2019-09-15 NAR_DRAFT.tex
publishes with numberic citations
Need to fix figure labels

@@
 \documentclass[a4,center,fleqn]{NAR}
-\bibliographystyle{plainnat}
+\bibliographystyle{NAR-natbib}
 \usepackage{color}
@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Fig. \ref{NAR-sigfig} \textbf{a-c}, shows mutation profiles from several tumors.
+For example, Figure \ref{NAR-sigfig}\textbf{A} through \textbf{C}, shows mutation profiles from several tumors.
 \rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
+\bkmcomment{Fixed in portions of text. need to change all figures to have capital letter panel labels. A and so on.}
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
@@
 \end{equation}
-The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-sigfig}).
+The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
 %Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
@@
 
-We model the observed mutation profile $S$ (Fig.~\ref{NAR-sigfig}a-c) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+We model the observed mutation profile $S$ (Figure~\ref{NAR-sigfig}\textbf{A} throught \textbf{C}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
 Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.

### 2019-09-16 NAR_DRAFT.tex
abstract. probably worse

@@
 \begin{abstract}
-The accumulation of somatic mutations is an inevitable consequence of tumor development.
+%The accumulation of somatic mutations is an inevitable consequence of tumor development.
 Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
 Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
-The identification of low frequency mutations from Next Generation Sequencing data remains challenging.
-In general the ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
-\rngcomment{A little too much intro in abstract.}
-This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
+The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
+% \rngcomment{A little too much intro in abstract.}
+% This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
 Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
@@
 We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
+The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and the identification lower frequency variants without generating excessive false positive calls.
 \rngcomment{Need a sentence about what new analysis this enables. How does \batcave (at least help) solve the problem you raised earlier?}

### 2019-09-16 NAR_DRAFT.tex
new figure 1 and caption

@@
   \includegraphics{figures/signatures_only.pdf}
-
-  \rngcomment{Need more basic labels for each panel. E.g. ``Real breast tumor'', ``Real accumulated myeloid leukemia'', ``Simulated concentrated profile''}
   \end{center}
   \caption{Tumor mutation profiles.
-  \rngcomment{Add explanation of how to read a profile. ``In each panel, the histograms show...''}
-  \textbf{(a)} The observed mutation profile of a breast tumor \cite{Shi2018}.
-  \textbf{(b)} The observed mutation profile of an acute myeloid leukemia \cite{Griffith2015}.
-  \textbf{(c)} \rngcomment{Add a third (and possibly fourth) real tumor, with a more concentrated profile.
-So we can illustrate that our simulations aren't out of line with reality.
-Ideally two tumors of same type that have noticeably different profiles.}
-  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
+  In each panel, the y-axis is the proportion of total mutations in the tumor with each of the 96 substitution types shown on the x-axis.
+  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
+  \textbf{(B)} The observed mutation profile of a breast tumor we used in our real data analysis \cite{Shi2018}.
+  \textbf{(C)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{(D)} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{(E)} Equal proportions of signatures 1, 3, \& 5.
+  \textbf{(F)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al. (2019) \citep{Alexandrov2019}.
+  \textbf{(G)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al. (2019) \citep{Alexandrov2019}.
   }

### 2019-09-16 NAR_DRAFT.tex
fixing rest of figure calls

@@
 The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
-% \rngcomment{A little too much intro in abstract.}
 % This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
@@
 We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
-The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and the identification lower frequency variants without generating excessive false positive calls.
-\rngcomment{Need a sentence about what new analysis this enables. How does \batcave (at least help) solve the problem you raised earlier?}
+The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and identification of lower frequency variants while providing well calibrated false positive rate expectations.
 We provide an R package implementing the \batcave algorithm for use with MuTect output.
@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure \ref{NAR-sigfig}\textbf{A} through \textbf{C}, shows mutation profiles from several tumors.
+For example, Figure \ref{NAR-sigfig}\textbf{A}--\textbf{C}, shows mutation profiles from several tumors.
 \rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
@@
   \textbf{(E)} Equal proportions of signatures 1, 3, \& 5.
-  \textbf{(F)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al. (2019) \citep{Alexandrov2019}.
-  \textbf{(G)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al. (2019) \citep{Alexandrov2019}.
+  \textbf{(F)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
+  \textbf{(G)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
   }
@@
     \caption{Model performance.
-    \textbf{(a)} Precision-recall curves and \textbf{(b)} Receiver operating characteristic curves for 3 mutation signatures.
-    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(c)} the MuTect model and \textbf{(d))} the \batcave model.
-    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(e)} and \textbf{(f)}.}
+    \textbf{(A)} Precision-recall curves and \textbf{(B)} Receiver operating characteristic curves for 3 mutation signatures.
+    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.
+    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(E)} and \textbf{(F)}.}
   \label{NAR-wes_fig}
@@
   \end{center}
-  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
+  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
   The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
@@
 We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-In Figure \ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
 For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.
@@
 We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
-For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
+For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}\textbf{C,D,E}.
 These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
@@
 We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
-By both metrics \batcave outperforms MuTect (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{c}, and Table \ref{table:01}).
+By both metrics \batcave outperforms MuTect (Figure~\ref{NAR-wes_fig}\textbf{A}--\textbf{B}, and Table~\ref{table:01}).
 The extent to which \batcave outperforms is dependent on both the sequencing depth and the level of concentration of the mutation signature.
@@
 
-The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table \ref{table:01}).
+The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table~\ref{table:01}).
 We note that the effective mutation rate is lower than the simulated rate of $3\cdot10^{-6}$.
@@
 This is useful in our setting because the density of posterior probabilities tends to be strongly bi-modal.
-A large fraction of true negative variants have posterior probabilities of less than $10{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
-As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
+A large fraction of true negative variants have posterior probabilities of less than $10^{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
+As we can see from the comparison in Figure~\ref{NAR-wes_fig}\textbf{C}--\textbf{D}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
 For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
@@
 
-Figure \ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration for 500X whole exomes in the portion of the posterior probability distribution relevant to clinicians and researchers.
+Figure~\ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration for 500X whole exomes in the portion of the posterior probability distribution relevant to clinicians and researchers.
 For any probability between 60 and 90\% \batcave is better calibrated than Mutect.
@@
 At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
-Figure \ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
+Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
 Note that recall never improves because the uncalled true positive variants are below the threshold of detection for any reasonable level of false positive calls.
@@
 We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{b} and Table \ref{table:01}).
+The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{E} and Table \ref{table:01}).
 
@@
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{b} and Table \ref{table:01}).
+We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
 

### 2019-09-16 NAR_DRAFT.tex
fixing a bunch of inline citations and figure labels

@@
 We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
-We downloaded the normal and primary tumor bam files generated for the whole genome sequences in \citet{Griffith2015} from dbGaP accession number phs000159.v8.p4.
-\citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
-The whole exome sequencing bam files generated by \citet{Shi2018} were downloaded from NCBI SRA no. SRP070662.
-\citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
+We downloaded the normal and primary tumor bam files generated for the whole genome sequences in Griffith et al., 2015~\citep{Griffith2015} from dbGaP accession number phs000159.v8.p4.
+Griffith et al., 2015~\citep{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
+The whole exome sequencing bam files generated by Shi et al., 2015~\citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
+Shi et al., 2018~\citep{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
 We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.
@@
 	We use the Integrated Calibration Index to quantify the difference between MuTect and \batcave \cite{Austin2019}.
-	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and \batcave (Figure \ref{NAR-wgs_fig} \textbf{e} and \textbf{f} and Table \ref{table:01}).
+	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and \batcave (Figure~\ref{NAR-wgs_fig}\textbf{A}--\textbf{B} and Table~\ref{table:01}).
 	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.
@@
 These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
-\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
+Griffith et al., 2015~\citep{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
@@
 
-\citet{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
+Shi et al., 2018~\citep{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
 We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
@@
 
-Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \cite{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.
-As \citet{Griffith2015} show, whole genome sequencing at approx.
+Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \citep{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.
+As Griffith et al., 2015~\citep{Griffith2015} show, whole genome sequencing at approx.
 350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing.

### 2019-09-16 NAR_DRAFT.tex
more typos and things that don't work corrected

@@
   \textbf{(E)} Equal proportions of signatures 1, 3, \& 5.
-  \textbf{(F)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
-  \textbf{(G)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
+  \textbf{(F)}--\textbf{(G)} The observed mutation profile of two real breast tumors obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
   }
@@
 The final piece of equation \ref{eqn:detailed_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which we specify as the per-base mutation rate $\mu$.
-In an exponentially growing and neutrally or nearly neutrally evolving tumor, \citet{Williams2018} show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
+In an exponentially growing and neutrally or nearly neutrally evolving tumor, Williams et al.
+2018~\citep{Williams2018} show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
 \begin{equation}

### 2019-09-16 NAR_DRAFT.tex
little things

@@
 \rngcomment{Got here...}
-
+\bkmcomment{No currently published paper has numbered equations, and it is not mentioned in the guidelines. I am using their template for numbered and non-numbered equations.}
 We model the observed mutation profile $S$ (Figure~\ref{NAR-sigfig}\textbf{A} throught \textbf{C}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
@@
 In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
-A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
+A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
 The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
@@
 We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
-Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (\bkmcomment{website or citation}).
+Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (version 2), downloaded from the COSMIC website.
 Each simulated profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.

### 2019-09-16 NAR_DRAFT.tex
little fixes

@@
 In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.
+For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
 

### 2019-09-16 NAR_DRAFT.tex
small fixes

@@
 Variants identified by MuTect are labelled as to whether they pass all MuTect filters, pass all filters *other* than the evidence threshold \textrm{tlod\_f\_star}, or fail to pass any filter other than \textrm{tlod\_f\_star}.
-Variants that pass all filters or fail only \textrm{tlod\_f\_star} are then passed to {method} for prior estimation and rescoring.
+Variants that pass all filters or fail only \textrm{tlod\_f\_star} are then passed to \batcave for prior estimation and rescoring.
 
@@
 
+\section{Data Availability}
+COSMIC mutation signatures version 2 is available at \footnotesize{https://cancer.sanger.ac.uk/cosmic/signatures\_v2}.
+
 \section{ACKNOWLEDGEMENTS}

### 2019-09-16 NAR_DRAFT.tex
change texttt to textrm for consistency

@@
 We designed \batcave to be used as a post-calling evaluation tool with MuTect.
-MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in equation \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
+MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in equation \ref{eqn:hypothesis} as either \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
 We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
@@
 
-The log posterior odds is the log likelihood ratio (\texttt{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as
+The log posterior odds is the log likelihood ratio (\textrm{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as
 
@@
 Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
-Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
-and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:
+Variants were spiked to create tumors with Bamsurgeon with default parameters~\citep{Ewing2015a},
+and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
 
@@
 \subsection{Real Data}
-We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
+We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer~\cite{Griffith2015,Shi2018}.
 We downloaded the normal and primary tumor bam files generated for the whole genome sequences in Griffith et al., 2015~\citep{Griffith2015} from dbGaP accession number phs000159.v8.p4.

### 2019-09-16 NAR_DRAFT.tex
add observation about APOBEC signature in 1F

@@
   \end{center}
+  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
   \caption{Tumor mutation profiles.
@@
 We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
-Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific that sharpens the classification and calibration of later mutations.
+Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
 Thus, \batcave is best described as an algorithm trained on early mutations.

### 2019-09-17 NAR_DRAFT.tex
fixing figure legends. need to fix labels in plots

@@
   \caption{Model performance on 100X simulated whole genome.
-  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
-  Distribution of estimated true positive probabilities for true positive  variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
+  \textbf{(A)} Precision recall curves and \textbf{(B)} Reciever operating characteristic curves for 3 mutation signatures.
+  Distribution of estimated true positive probabilities for true positive  variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.
   A perfectly calibrated model would generate the diagonal line.}
@@
   \caption{Model performance on 100X simulated whole genome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
+  Positive predictive values vs. Recall for five different probability cutoffs. 
   Note the restricted axes to improve readability of points.}

### 2019-09-17 NAR_DRAFT.tex
fix supp fig S1 and add smaller file size png version
trying to get figure 3 to fit in the text at a reasonable place
non-breaking spaces

@@
 \end{equation}
-where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
-When comparing our posterior odds to those of MuTect in this paper, we following \citet{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
+where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation~\ref{eqn:detailed_prior} and specified in equations~\ref{eqn:post_pred}--\ref{eqn:mut_rate}.
+When comparing our posterior odds to those of MuTect in this paper, we following Cibulskis et al., 2013~\citep{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
 We compute the posterior odds for Mutect as
@@
 
+\subsection{Convergence of the site-specific prior to the data generating distribution}
+The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
+We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
+In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
+
 \begin{figure}
   \begin{center}
-  \includegraphics{figures/kl_only.pdf}
+  \includegraphics[width=.35\textwidth]{figures/kl_only.pdf}
   \end{center}
@@
 \end{figure}
-
-\subsection{Convergence of the site-specific prior to the data generating distribution}
-The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
-We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
 
@@
   \begin{center}
-  \includegraphics{figures/fig_wgs.pdf}
+  \includegraphics[width=\textwidth]{figures/fig_wgs.png}
   \end{center}

### 2019-09-17 NAR_DRAFT.tex
add text to describe the mutation profile figure in more detail

@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure \ref{NAR-sigfig}\textbf{A}--\textbf{C}, shows mutation profiles from several tumors.
-\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
-\bkmcomment{Fixed in portions of text. need to change all figures to have capital letter panel labels. A and so on.}
+For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
+These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
+The synthetic tumor mutation profiles we generated are designed to capture this diversity, and assess the performance of \batcave across a spectrum of reasonable mutation profiles.
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

### 2019-09-17 NAR_DRAFT.tex
pretty ugly way of wording things, but in a better place

@@
 These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
-The synthetic tumor mutation profiles we generated are designed to capture this diversity, and assess the performance of \batcave across a spectrum of reasonable mutation profiles.
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
@@
 The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
+The synthetic tumor mutation profiles we generated are designed to capture the observed diversity in mutation profiles, and assess the performance of \batcave across a spectrum of reasonable mutation profiles.
 The simulated mutation rate was $\mu = 3\cdot10^{-6}$.

### 2019-09-17 NAR_DRAFT.tex
hate the figure placement hassle
get there eventually
good figures now
number all of the equations for consistency

@@
 We thus model $\mathrm{p}(\mathrm{S})$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
-$$
+\begin{equation}
 \begin{aligned}
@@
 \end{aligned}
-$$
+\end{equation}
 In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.
@@
 The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
-  $$
+\begin{equation}
     \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
-  $$
+\end{equation}
+
 Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
@@
 
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=.35\textwidth]{figures/kl_only.pdf}
+  \end{center}
+  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
+  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
+\label{NAR-kl_fig}
+\end{figure}
+
 \subsection{Real Data}
@@
 For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
-
-\begin{figure}
-  \begin{center}
-  \includegraphics[width=.35\textwidth]{figures/kl_only.pdf}
-  \end{center}
-  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
-  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
-\label{NAR-kl_fig}
-\end{figure}
 
@@
   \begin{center}
-  \includegraphics{figures/ppv_wes.pdf}
+  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
   \end{center}

### 2019-09-17 NAR_DRAFT.tex
making changes to wes figure caption

@@
 \rngcomment{Got here...}
-\bkmcomment{No currently published paper has numbered equations, and it is not mentioned in the guidelines. I am using their template for numbered and non-numbered equations.}
+\bkmcomment{No currently published paper has numbered equations, and it is not mentioned in the guidelines.
+I am using their template for numbered equations and made all equations numbered for consistency.}
 We model the observed mutation profile $S$ (Figure~\ref{NAR-sigfig}\textbf{A} throught \textbf{C}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
@@
     Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.
-    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(E)} and \textbf{(F)}.}
+    A perfectly calibrated model would generate the diagonal line. 
+    \textbf{(E)}--\textbf{(F)} Precision-recall curves for real data.}
   \label{NAR-wes_fig}

### 2019-09-17 NAR_DRAFT.tex
extraneous and
constant not uniform

@@
 These methods vary in their complexity and specific focus.
-But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.
+But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.
 
@@
 At the focal site, a mutation $m$ can be to any of three alternate alleles.
-Indexing by the $c=\{1 \dots 32\}$ contexts and and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
+Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
 Eq.~\ref{eqn:full_prior} is then

### 2019-09-17 NAR_DRAFT.tex
follow not following

@@
 where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation~\ref{eqn:detailed_prior} and specified in equations~\ref{eqn:post_pred}--\ref{eqn:mut_rate}.
-When comparing our posterior odds to those of MuTect in this paper, we following Cibulskis et al., 2013~\citep{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
+When comparing our posterior odds to those of MuTect in this paper, we follow Cibulskis et al., 2013~\citep{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
 We compute the posterior odds for Mutect as

### 2019-09-17 NAR_DRAFT.tex
Comments from Ryan through Results.

--- 2019-09-17 NAR_DRAFT.tex
@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
-These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
+For example, Figure \ref{NAR-sigfig}\textbf{A}--\textbf{C}, shows mutation profiles from several tumors.
+\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
+\bkmcomment{Fixed in portions of text. need to change all figures to have capital letter panel labels. A and so on.}
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
@@
   \caption{Tumor mutation profiles.
-  In each panel, the y-axis is the proportion of total mutations in the tumor with each of the 96 substitution types shown on the x-axis.
+  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
   \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
-  \textbf{(B)} The observed mutation profile of a breast tumor we used in our real data analysis \cite{Shi2018}.
-  \textbf{(C)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(D)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(E)} Equal proportions of signatures 1, 3, \& 5.
-  \textbf{(F)}--\textbf{(G)} The observed mutation profile of two real breast tumors obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
+  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
+  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
+  \textbf{()} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{()} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{()} Equal proportions of signatures 1, 3, \& 5.
+  Move simulated profiles to own figure.
    }
@@
 The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
+Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
 At the focal site, a mutation $m$ can be to any of three alternate alleles.
@@
 
-\rngcomment{Got here...}
-\bkmcomment{No currently published paper has numbered equations, and it is not mentioned in the guidelines.
-I am using their template for numbered equations and made all equations numbered for consistency.}
-We model the observed mutation profile $S$ (Figure~\ref{NAR-sigfig}\textbf{A} throught \textbf{C}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-sigfig}\textbf{A}-\textbf{E}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
 Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
@@
 But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
-We thus model $\mathrm{p}(\mathrm{S})$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
+We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
 \begin{equation}
@@
 These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
-Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 with the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
+Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 without the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
 The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
@@
 \end{equation}
-
-Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
+Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
 \begin{equation}
   \label{eqn:post_pred}
-  $$
   \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
-  $$
 \end{equation}
-The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
+The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$ is then
 \begin{equation}
   \label{eqn:to_allele}
-  $$
    \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
-  $$
 \end{equation}
-The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
-We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
-The R implementation for \batcave ships with pre-computed tables for both whole exomes and whole genomes.
+The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
+The R implementation of \batcave ships with pre-computed tables for both whole exomes and whole genomes.
 
-\subsection{Estimation of the mutation rate.}
-The final piece of equation \ref{eqn:detailed_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which we specify as the per-base mutation rate $\mu$.
-In an exponentially growing and neutrally or nearly neutrally evolving tumor, Williams et al.
-2018~\citep{Williams2018} show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
+\subsection{Estimation of the mutation rate}
+The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base mutation rate $\mu$.
+In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
 \begin{equation}
   \label{eqn:mut_rate}
-$$
   \mathrm{M}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
-$$
+\end{equation}
+The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
+The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
+We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
+We can then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
+We can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
+In the R implementation of \batcave, $f_{min}$ is a free parameter.
+For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
+%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
+%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
+
+\subsection{Likelihood function}
+The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq.
+\ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
+We used MuTect 1.1.7 for all analyses in this paper, so we have
+\begin{equation}
+  \label{eqn:tlod}
+    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
+\end{equation}
+The log posterior odds is the log likelihood ratio (\textrm{TLOD}) plus the log prior odds, so the posterior odds in favor of the alternate hypothesis for a given substitution type is
+\begin{equation}
+  \label{eqn:computed_posterior}
+  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))}.
+\end{equation}
+Here $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in Eq.~\ref{eqn:detailed_prior} and specified in Eq.~\ref{eqn:post_pred}-\ref{eqn:mut_rate}.
+When comparing our posterior odds to those of MuTect, we assume a uniform per-base probability of mutation of $3\cdot10^{-6}$ \cite{Cibulskis2013}, so
+\begin{equation}  \label{eqn:mutect_posterior}
+  \frac{\mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})}{1 - \mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6}.
 \end{equation}
 
-The number of bases $N$ is fixed at $3\cdot10^9$ for a whole genome sequence and $3\cdot10^7$ for a whole exome.
-Returning to the set of high confidence mutations $\mathrm{D}$, we can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, $f_{min}$ as the lowest, and count the number of mutations in $D$ falling between the two.
-In practice it is usually best to select a conservative value for $f_{min}$, which is strongly dependent on sequencing depth, and we include this as a parameter in the R implementation of \batcave.
-In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
-A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
-The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
-We make the simplifying assumption that there is no cell death, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
-Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
-
-\subsection{The likelihood function}
-We designed \batcave to be used as a post-calling evaluation tool with MuTect.
-MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in equation \ref{eqn:hypothesis} as either \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
-We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with
-
-\begin{equation}
-  \label{eqn:tlod}
-    $$
-    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
-    $$
-\end{equation}
-
-The log posterior odds is the log likelihood ratio (\textrm{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as
-
-\begin{equation}
-  \label{eqn:computed_posterior}
-  $$
-  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))},
-  $$
-\end{equation}
-where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation~\ref{eqn:detailed_prior} and specified in equations~\ref{eqn:post_pred}--\ref{eqn:mut_rate}.
-When comparing our posterior odds to those of MuTect in this paper, we following Cibulskis et al., 2013~\citep{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
-We compute the posterior odds for Mutect as
-\begin{equation}
-  \label{eqn:mutect_posterior}
-  $$
-  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
-  $$
-\end{equation}
-since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$.
+\subsection{Implementation}
+\rngcomment{Add section about code implementation. What libraries are using, etc.?}
 
 \subsection{Tumor simulations.}
-We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
-Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (version 2), downloaded from the COSMIC website.
-Each simulated profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
-The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
-The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
-The synthetic tumor mutation profiles we generated are designed to capture the observed diversity in mutation profiles, and assess the performance of \batcave across a spectrum of reasonable mutation profiles.
-The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
-Variants were selected from a combination of the Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
-Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
-Variants were spiked to create tumors with Bamsurgeon with default parameters~\citep{Ewing2015a},
-and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
-
-\begin{tiny}
+We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
+Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
+\rngcomment{Put COSMIC website as a citation.}
+Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
+The "Concentrated" profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
+The "Intermediate" profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
+The "Diffuse" profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
+\rngcomment{Citations for these signature combinations.}
+Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
+Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
+Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
+\begin{scriptsize}
 \begin{verbatim}
-
-  java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect --reference_sequence $ref_path \
+java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
+        --reference_sequence $ref_path \
         --dbsnp $db_snp \
@@
         --coverage_file $out_path/$chr.cov
+\end{verbatim}
+\end{scriptsize}
+Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
+Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.
 
-\end{verbatim}
-\end{tiny}
-Variants identified by MuTect are labelled as to whether they pass all MuTect filters, pass all filters *other* than the evidence threshold \textrm{tlod\_f\_star}, or fail to pass any filter other than \textrm{tlod\_f\_star}.
-Variants that pass all filters or fail only \textrm{tlod\_f\_star} are then passed to \batcave for prior estimation and rescoring.
+\subsection{Real Data}
+We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
+We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
+Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
+We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
+Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
+We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
+
+\subsection{Calibration metric}
+The quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
+Briefly, a loess-smoothed regression was fit by regressing the binary (True=1,False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
+For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$, and the Integrated Calibration Index is the average distance between the fit and that line.
+
+\section{RESULTS}
+We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
+\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites in to high- and low-confidence groups.
+The mutation profile is estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
+The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500x exome and 2,000 seconds for a 100x whole genome.
+
+To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
+The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{XXX}\textbf{A-C}).
+We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffiths2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
+In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
 
@@
   \begin{center}
-  \includegraphics[width=.35\textwidth]{figures/kl_only.pdf}
+  \includegraphics{figures/kl_only.pdf}
   \end{center}
-  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
-  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
+  \caption{Convergence of the site-specific prior to the data generating distribution.
+Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
+  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}
 \label{NAR-kl_fig}
@@
 
-\subsection{Real Data}
-We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer~\cite{Griffith2015,Shi2018}.
-We downloaded the normal and primary tumor bam files generated for the whole genome sequences in Griffith et al., 2015~\citep{Griffith2015} from dbGaP accession number phs000159.v8.p4.
-Griffith et al., 2015~\citep{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
-The whole exome sequencing bam files generated by Shi et al., 2015~\citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
-Shi et al., 2018~\citep{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
-We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.
+\subsection{Convergence of the prior}
+To improve performance, the context-dependent prior probability of mutation must converge to a good representation of the data generating distribution within the set of high-confidence mutations.
+When applied to simulated whole exome and whole genome data, the prior converges within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
+For comparison, in our simulated data sets the number of high-confidence mutations ranges between 1500 and 5000, and for a real acute myeloid leukemia it is over 17,000~\cite{Griffith2015}.
 
-\subsection{Calibration metric}
-
-	We use the Integrated Calibration Index to quantify the difference between MuTect and \batcave \cite{Austin2019}.
-	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and \batcave (Figure~\ref{NAR-wgs_fig}\textbf{A}--\textbf{B} and Table~\ref{table:01}).
-	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.
+%We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
+%In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
+%For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
 
@@
     \end{center}
-    \caption{Model performance.
-    \textbf{(A)} Precision-recall curves and \textbf{(B)} Receiver operating characteristic curves for 3 mutation signatures.
-    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.
-    A perfectly calibrated model would generate the diagonal line. 
-    \textbf{(E)}--\textbf{(F)} Precision-recall curves for real data.}
+    \caption{Variant-calling performance on both simulated and real data.
+    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation signatures.
+    \rngcomment{Make labels for \batcave and MuTect in A much bigger. Add similar labels to B.}
+    \rngcomment{Move "Synthetic 500x whole exomes" to above A\&C. Add similar label for B\&D}
+    \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated.
+For a perfectly calibrated caller, that curve would match the dashed y=x line.
+    \rngcomment{Remove dark edges of distributions. Make diagonal line dashed.}
+    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed
+    \rngcomment{Label plots with what type of cancer they are, plus depth.}.}
   \label{NAR-wes_fig}
@@
 
-\section{RESULTS}
-We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-The algorithm is effecient, processing 22,000 variants per second on a typical iMac desktop computer, which corresponds to 100 seconds to process a 500X exome and 2000 seconds for a 100X whole genome.
-MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
-\batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
-The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
+\begin{table*}[b]
+  \tableparts{%
+  \caption{Performance characteristics as measured for all data sets.}
+  \label{table:01}%
+  }{%
+  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
+  \toprule
+  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
+  \\
+  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
+  \\
+  \colrule
+  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
+  \\
+  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
+  \\
+  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
+  \\
+  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
+  \\
+  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
+  \\
+  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
+  \\
+  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+  \\
+  real breast \cite{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
+  \\
+  \botrule
+  \end{tabular*}%
+  }
+  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
+ \end{table*}
 
-\subsection{Convergence of the site-specific prior to the data generating distribution}
-The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
-We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
-
-\subsection{Simulated data}
-We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
-For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}\textbf{C,D,E}.
-These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.
-
-We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
-By both metrics \batcave outperforms MuTect (Figure~\ref{NAR-wes_fig}\textbf{A}--\textbf{B}, and Table~\ref{table:01}).
-The extent to which \batcave outperforms is dependent on both the sequencing depth and the level of concentration of the mutation signature.
+\subsection{Tests using simulated data}
+We assessed classification performance using the areas under both the receiver operating characteristic and the precision-recall curves, because the classes are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
+By both metrics \batcave outperforms MuTect (Fig.~\ref{NAR-wes_fig}\textbf{A}\&\textbf{B} and Table~\ref{table:01}).
+The extent of the performance difference is dependent on both the sequencing depth and the concentration of the mutation signature.
 Deeper sequencing and more concentrated mutation signatures increase the performance advantage of \batcave.
 
-The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table~\ref{table:01}).
-We note that the effective mutation rate is lower than the simulated rate of $3\cdot10^{-6}$.
-In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.
+For all simulated tumors, the estimated mutation rate is approximately $3\cdot10^{-7}$ (Table~\ref{table:01}), which is lower than the simulated rate of $3\cdot10^{-6}$.
+This is due to restrictions with BAMSurgeon, such as sequencing depth and quality, that prevent 100\% of simulated variants from being inserted into the reads.
 
-We measured overall calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
-The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
-This is useful in our setting because the density of posterior probabilities tends to be strongly bi-modal.
-A large fraction of true negative variants have posterior probabilities of less than $10^{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
-As we can see from the comparison in Figure~\ref{NAR-wes_fig}\textbf{C}--\textbf{D}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
-For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
-This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.
-As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.
+We also assessed the calibration of \batcave, assessing the likelihood that a potential variant with a given posterior probability is actually a true variant.
+We measured overall calibration performance using the Integrated Calibration Index (ICI) \cite{Austin2019}, which measures the difference between predicted and observed probabilities, weighted by the density of the predicted probabilities.
+This metric is particularly useful in our case, because the density of posterior probabilities is bi-modal (Fig.~\ref{NAR-wes_fig}C\&D).
+A large fraction of true negative variants have posterior probabilities less than $10^{-4}$, far below any meaningful threshold, so we evaluate calibration only on potential variants with posterior probability greater than 0.01.
+For these potential variants, \batcave tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D) while decreasing probabilities of low probability but false variants.
+For 500X tumors, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
+This is because that there are few low probability true positive variants in tumors sequenced to 100X depth.
+As with the other metrics, the advantage of \batcave increases with the concentration of the mutation profile the sequencing depth.
 
-Figure~\ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration for 500X whole exomes in the portion of the posterior probability distribution relevant to clinicians and researchers.
-For any probability between 60 and 90\% \batcave is better calibrated than Mutect.
-Essentially, for any posterior probability above 70\% MuTect has a false positive rate around 8\%, while \batcave has a false positive rate that increases as posterior probability decreases.
-The cost of this compression is recall, which is 3-5\% better than MuTect at 95\% posterior probability, and 3-7\% better at 90\%, depending on muation profile concentration.
+In practice, accurate calibration is most important for variants with high posterior probability.
+For variants with posterior probability between 60 and 90\%, \batcave is substantially better calibrated than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as posterior probability increases.
+The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
 At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
-Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
-Note that recall never improves because the uncalled true positive variants are below the threshold of detection for any reasonable level of false positive calls.
+%Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
 
-\subsection{Real tumor data.}
-We utilized two real data sets.
-In each, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
-Griffith et al., 2015~\citep{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
-We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
-The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{E} and Table \ref{table:01}).
+\subsection{Tests using real tumor data}
+We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
+We considered only precision-recall comparisons for these data, because the validation procedure yields true positives but not comprehensive true negatives. \rngcomment{Is  a correct statement?}
 
-Shi et al., 2018~\citep{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
-We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
+Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which would be considered normal in AML. \rngcomment{Really need a citation for this!}
+For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
 
-We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent \cite{Griffith2015,Shi2018}.
-These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
-At the same time, there is no way to compute true and false negative fractions for these data sets.
-As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
+Shi et al.\ performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$ and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
+%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$. \rngcomment{Any comment on this result?}
+and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
+For the validated variants, MuTect and \batcave yield almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
 
@@
   \begin{center}
-  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
+  \includegraphics{figures/ppv_wes.pdf}
   \end{center}
@@
 
+These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
+At the same time, there is no way to compute true and false negative fractions for these data sets.
+As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
+
 We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
+MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
 However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
@@
 This work was supported by the National Science Foundation via Graduate Research Fellowship award number DGE-1143953 to BKM and by the National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM127348 to RNG.
-We thank Professor Edward J. Bedrick for fruitful discussions about the statistical model.
+We thank Prof.\ Edward J. Bedrick for fruitful discussions about the statistical model.
 
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/fig_wgs.png}
+  \includegraphics{figures/fig_wgs.pdf}
   \end{center}
@@
 
-\section{Supplememtary Table}
-
-\begin{table}[b]
-  \tableparts{%
-  \caption{Performance characteristics as measured for all data sets.}
-  \label{table:01}%
-  }{%
-  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
-  \toprule
-  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
-  \\
-  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
-  \\
-  \colrule
-  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
-  \\
-  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
-  \\
-  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
-  \\
-  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
-  \\
-  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
-  \\
-  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
-  \\
-  \citet{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
-  \\
-  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
-  \\
-  \botrule
-  \end{tabular*}%
-  }
-  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
- \end{table}
-
 \end{document}

### 2019-09-17 NAR_DRAFT.tex

@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure \ref{NAR-sigfig}\textbf{A}--\textbf{C}, shows mutation profiles from several tumors.
-\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
-\bkmcomment{Fixed in portions of text. need to change all figures to have capital letter panel labels. A and so on.}
+For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
+These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
@@
 Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=.35\textwidth]{figures/kl_only.pdf}
+  \end{center}
+  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
+  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
+\label{NAR-kl_fig}
+\end{figure}
 

### 2019-09-17 NAR_DRAFT.tex
Fix some merge screw ups

@@
   \textbf{()} Equal proportions of signatures 1, 3, \& 5.
-  Move simulated profiles to own figure.
+  \rngcomment{Move simulated profiles to own figure.}
    }
@@
 
-\subsection{Tumor simulations.}
+\subsection{Tumor simulations}
 We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
@@
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The "Concentrated" profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
-The "Intermediate" profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
-The "Diffuse" profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
+The ``Concentrated'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
+The ``Intermediate'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
+The ``Diffuse'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
 \rngcomment{Citations for these signature combinations.}
@@
 
-\begin{figure}
-  \begin{center}
-  \includegraphics[width=.35\textwidth]{figures/kl_only.pdf}
-  \end{center}
-  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
-  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
-\label{NAR-kl_fig}
-\end{figure}
-
-\subsection{Real Data}
+\subsection{Real data}
 We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.

### 2019-09-18 NAR_DRAFT.tex
incorporate changes
notes to myself

@@
 \subsection{Implementation}
+\bkmcomment{Used packages from bioconductor(name them), to leverage. Follow mapex example.}
 \rngcomment{Add section about code implementation. What libraries are using, etc.?}
@@
 \rngcomment{Citations for these signature combinations.}
+\bkmcomment{These are really from the text of the cosmic website, the citation is probably Alexandrov2015 or Alexandrov2013a}
 Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
@@
 Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
-  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}
+  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}\bkmcomment{just saw it somewhere, change to comma.}
 \label{NAR-kl_fig}
@@
 We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-We considered only precision-recall comparisons for these data, because the validation procedure yields true positives but not comprehensive true negatives. \rngcomment{Is  a correct statement?}
+We considered only precision-recall comparisons for these data, because the validation procedure yields true positives but not comprehensive true negatives.
+\rngcomment{Is  a correct statement?}\bkmcomment{Yes, but examine aml31 closely to make sure, and then rearrange this section to explain why.}
 
 Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which would be considered normal in AML. \rngcomment{Really need a citation for this!}
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which would be considered normal in AML. \rngcomment{Really need a citation for this!}\bkmcomment{I have one, find it.}
 For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
@@
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$. \rngcomment{Any comment on this result?}
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$.
+\rngcomment{Any comment on this result?}\bkmcomment{not terribly out of line with other breast cancers.
+find a citation.
+other possibilities: selection?;all lumped together?}
 and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).

### 2019-09-18 NAR_DRAFT.tex
added second signature figure and fixed several things in KL figure

@@
   \begin{center}
-  \includegraphics{figures/signatures_only.pdf}
+  \includegraphics{figures/real_signatures_only.pdf}
   \end{center}
-  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
   \caption{Tumor mutation profiles.
@@
   \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
-  \textbf{()} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{()} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{()} Equal proportions of signatures 1, 3, \& 5.
-  \rngcomment{Move simulated profiles to own figure.}
    }
-  \label{NAR-sigfig}
+  \label{NAR-realsigfig}
  \end{figure}
@@
 
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/sim_signatures_only.pdf}
+  \end{center}
+  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
+  \caption{Tumor mutation profiles.
+  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
+  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
+   }
+  \label{NAR-simsigfig}
+ \end{figure}
 \subsection{Tumor simulations}
@@
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The ``Concentrated'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
-The ``Intermediate'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
-The ``Diffuse'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
+The ``Concentrated'' profile (Fig.~\ref{simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
+The ``Intermediate'' profile (Fig.~\ref{simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
+The ``Diffuse'' profile (Fig.~\ref{simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
 \rngcomment{Citations for these signature combinations.}
@@
 Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
-  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}\bkmcomment{just saw it somewhere, change to comma.}
+  }
 \label{NAR-kl_fig}

### 2019-09-18 NAR_DRAFT.tex
mutation rates investigation

@@
 We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-We considered only precision-recall comparisons for these data, because the validation procedure yields true positives but not comprehensive true negatives.
-\rngcomment{Is  a correct statement?}\bkmcomment{Yes, but examine aml31 closely to make sure, and then rearrange this section to explain why.}
+Both data sets were validated by deep targeted sequencing, yielding high quality false positive data.
+However, \texttt{only} variants called by at least one variant caller were validated.
+As a result, there is no validated measure of true/false negative calls, so we considered only precision-recall comparisons for these data.
+\rngcomment{Is  a correct statement?}\bkmcomment{Yes, definitely true.}
 
@@
 We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$.
-\rngcomment{Any comment on this result?}\bkmcomment{not terribly out of line with other breast cancers.
-find a citation.
-other possibilities: selection?;all lumped together?}
+\rngcomment{Any comment on this result?}\bkmcomment{Noone, as far as I can tell has done a systematic calculation for TCGA.
+\cite{Williams2018} is only out in 2018.
+On the basis of tumor mutation burden studies there breast cancer is one of the slowest mutators, but there is no way to go from TMB to mutation rate since it depends on depth/frequency.}
 and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).

### 2019-09-18 NAR_DRAFT.tex
mutation burden/rate arguing with myself

@@
 \cite{Williams2018} is only out in 2018.
-On the basis of tumor mutation burden studies there breast cancer is one of the slowest mutators, but there is no way to go from TMB to mutation rate since it depends on depth/frequency.}
+On the basis of tumor mutation burden studies breast cancer is one of the slowest mutators, but there is no way to go from TMB to mutation rate since it depends on depth/frequency.
+The one breast cancer studied in Williams 2018 the estimate at just over 1e-7.
+that is a whole genome whereas ours is an exome, and theirs is deeper.
+My argument would be that it is not out of line with previous estimates for breast cancer and it is what the data indicates.}
 and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).

### 2019-09-18 NAR_DRAFT.tex
citations for simulated spectra

@@
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The ``Concentrated'' profile (Fig.~\ref{simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
-The ``Intermediate'' profile (Fig.~\ref{simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
-The ``Diffuse'' profile (Fig.~\ref{simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
+The ``Concentrated'' profile (Fig.~\ref{simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
+The ``Intermediate'' profile (Fig.~\ref{simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
+The ``Diffuse'' profile (Fig.~\ref{simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
 \rngcomment{Citations for these signature combinations.}

### 2019-09-18 NAR_DRAFT.tex
citations

--- 2019-09-18 NAR_DRAFT.tex
@@
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The ``Concentrated'' profile (Fig.~\ref{simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
-The ``Intermediate'' profile (Fig.~\ref{simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
-The ``Diffuse'' profile (Fig.~\ref{simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
-\rngcomment{Citations for these signature combinations.}
-\bkmcomment{These are really from the text of the cosmic website, the citation is probably Alexandrov2015 or Alexandrov2013a}
+The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
+The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
+The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
 Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.

### 2019-09-19 NAR_DRAFT.tex
citations in intro

@@
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
+joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
 These methods vary in their complexity and specific focus.
@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
+For example, Figure~\ref{NAR-realsigfig}, shows mutation profiles from several tumors.
 These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
@@
 \end{equation}
-The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
+The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
 %Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
 
-We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-sigfig}\textbf{A}-\textbf{E}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}\textbf{A}-\textbf{D}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
 Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
@@
 We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
-Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
+Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) %\cite{XXX}.
 \rngcomment{Put COSMIC website as a citation.}
@@
 To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
-The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{XXX}\textbf{A-C}).
-We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffiths2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
+The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NA-realsigfig}\textbf{A-C}).
+We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
 In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.

### 2019-09-19 NAR_DRAFT.tex
aml citation

@@
 Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which would be considered normal in AML. \rngcomment{Really need a citation for this!}\bkmcomment{I have one, find it.}
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is inline with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
 For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

### 2019-09-19 NAR_DRAFT.tex
a stab at it anyway
moved some figure labels around
getting there

@@
   \end{center}
+  \bkmcomment{tumor in C is probably normal breast plus APOBEC. Mention somewhere?}
   \caption{Tumor mutation profiles.
@@
   \end{center}
-  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
   \caption{Tumor mutation profiles.
@@
     \begin{center}
-    \includegraphics{figures/fig_wes.pdf}
+    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
     \end{center}
@@
     \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation signatures.
-    \rngcomment{Make labels for \batcave and MuTect in A much bigger. Add similar labels to B.}
-    \rngcomment{Move "Synthetic 500x whole exomes" to above A\&C. Add similar label for B\&D}
     \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated.
 For a perfectly calibrated caller, that curve would match the dashed y=x line.
-    \rngcomment{Remove dark edges of distributions. Make diagonal line dashed.}
-    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed
-    \rngcomment{Label plots with what type of cancer they are, plus depth.}.}
+    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.}
+
   \label{NAR-wes_fig}

### 2019-09-20 NAR_DRAFT.tex
make ppv plot bigger
fix figure 2 caption
implementation

@@
 \subsection{Implementation}
-\bkmcomment{Used packages from bioconductor(name them), to leverage. Follow mapex example.}
-\rngcomment{Add section about code implementation. What libraries are using, etc.?}
+We have implemented the \batcave algorithm as an R package (batcaver).
+The package leverages the Bioconductor packages \texttt{BSgenome}, \texttt{GenomicAlignments}, \texttt{VariantAnnotation}, and \texttt{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
+Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
 
@@
  \end{figure}
+
 \subsection{Tumor simulations}

### 2019-09-20 NAR_DRAFT.tex
figuring out how to get a web site as a citation

@@
 We have implemented the \batcave algorithm as an R package (batcaver).
-The package leverages the Bioconductor packages \texttt{BSgenome}, \texttt{GenomicAlignments}, \texttt{VariantAnnotation}, and \texttt{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
+The package leverages the Bioconductor packages \textrm{BSgenome}, \textrm{GenomicAlignments}, \textrm{VariantAnnotation}, and \textrm{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
 Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
@@
 We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
-Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) %\cite{XXX}.
-\rngcomment{Put COSMIC website as a citation.}
+Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
@@
   \begin{center}
-  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
+  \includegraphics[width=.45\textwidth]{figures/ppv_wes.pdf}
   \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
+  \caption{Model performance on 500X simulated whole exomes. 
+  Positive predictive value vs. Recall for five different probability cutoffs. 
   Note the restricted axes to improve readability of points}

### 2019-09-20 NAR_DRAFT.tex
comments and web citation for mutation signatures
fix sim figure reference

@@
   \end{center}
+  \bkmcomment{This can be smaller. regenerate with its own base\_height}
   \caption{Tumor mutation profiles.
@@
 To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
-The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NA-realsigfig}\textbf{A-C}).
+The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}\textbf{A-C}).
 We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
@@
   \end{center}
+  \bkmcomment{this would probably look OK shorter too}
   \caption{Convergence of the site-specific prior to the data generating distribution.

### 2019-09-23 NAR_DRAFT.tex
more figure comments
figure fixes and remove comments
Revisions to dicussion

--- 2019-09-20 NAR_DRAFT.tex
@@
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
+joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
 These methods vary in their complexity and specific focus.
@@
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure~\ref{NAR-realsigfig}, shows mutation profiles from several tumors.
+For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
 These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
@@
   \begin{center}
-  \includegraphics{figures/real_signatures_only.pdf}
+  \includegraphics{figures/signatures_only.pdf}
   \end{center}
-  \bkmcomment{tumor in C is probably normal breast plus APOBEC. Mention somewhere?}
+  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
   \caption{Tumor mutation profiles.
@@
   \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
+  \textbf{()} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  \textbf{()} Equal proportions of signatures 1, 4, \& 5.
+  \textbf{()} Equal proportions of signatures 1, 3, \& 5.
+  \rngcomment{Move simulated profiles to own figure.}
    }
-  \label{NAR-realsigfig}
+  \label{NAR-sigfig}
  \end{figure}
@@
 \end{equation}
-The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
+The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
 %Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
 
-We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}\textbf{A}-\textbf{D}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-sigfig}\textbf{A}-\textbf{E}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
 Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
@@
 \subsection{Implementation}
-We have implemented the \batcave algorithm as an R package (batcaver).
-The package leverages the Bioconductor packages \texttt{BSgenome}, \texttt{GenomicAlignments}, \texttt{VariantAnnotation}, and \texttt{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
-Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
-
-\begin{figure}
-  \begin{center}
-  \includegraphics{figures/sim_signatures_only.pdf}
-  \end{center}
-  \caption{Tumor mutation profiles.
-  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
-  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
-   }
-  \label{NAR-simsigfig}
- \end{figure}
+\rngcomment{Add section about code implementation. What libraries are using, etc.?}
 
@@
 We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
-Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) %\cite{XXX}.
+Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
 \rngcomment{Put COSMIC website as a citation.}
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
-The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
-The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
+The ``Concentrated'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
+The ``Intermediate'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
+The ``Diffuse'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
+\rngcomment{Citations for these signature combinations.}
 Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
@@
 We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
+\rngcomment{Did MuTect fail to call anything we know is there?}
 
@@
 To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
-The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NA-realsigfig}\textbf{A-C}).
-We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
+The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{XXX}\textbf{A-C}).
+We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffiths2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
 In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
@@
 Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
-  }
+  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}
 \label{NAR-kl_fig}
@@
     \begin{center}
-    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
+    \includegraphics{figures/fig_wes.pdf}
     \end{center}
@@
     \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation signatures.
+    \rngcomment{Make labels for \batcave and MuTect in A much bigger. Add similar labels to B.}
+    \rngcomment{Move "Synthetic 500x whole exomes" to above A\&C. Add similar label for B\&D}
     \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated.
 For a perfectly calibrated caller, that curve would match the dashed y=x line.
-    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.}
-
+    \rngcomment{Remove dark edges of distributions. Make diagonal line dashed.}
+    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed
+    \rngcomment{Label plots with what type of cancer they are, plus depth.}.}
   \label{NAR-wes_fig}
@@
 
+\begin{figure}
+  \begin{center}
+  \includegraphics{figures/ppv_wes.pdf}
+  \end{center}
+  \caption{Calibration of posterior probability. 
+  For realistic thresholds, posterior probabilities are more accurately calibrated for \batcave than MuTect, no matter how concentration the mutation profile is.}
+\label{NAR-ppv_fig}
+\end{figure}
+
 In practice, accurate calibration is most important for variants with high posterior probability.
@@
 We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-Both data sets were validated by deep targeted sequencing, yielding high quality false positive data.
-However, \texttt{only} variants called by at least one variant caller were validated.
-As a result, there is no validated measure of true/false negative calls, so we considered only precision-recall comparisons for these data.
-\rngcomment{Is  a correct statement?}\bkmcomment{Yes, definitely true.}
+We considered only precision-recall comparisons for these data, because the validation procedure yields true positives but not comprehensive true negatives. \rngcomment{Is  a correct statement?}
 
 Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is inline with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which would be considered normal in AML. \rngcomment{Really need a citation for this!}
 For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
@@
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$.
-\rngcomment{Any comment on this result?}\bkmcomment{Noone, as far as I can tell has done a systematic calculation for TCGA.
-\cite{Williams2018} is only out in 2018.
-On the basis of tumor mutation burden studies breast cancer is one of the slowest mutators, but there is no way to go from TMB to mutation rate since it depends on depth/frequency.
-The one breast cancer studied in Williams 2018 the estimate at just over 1e-7.
-that is a whole genome whereas ours is an exome, and theirs is deeper.
-My argument would be that it is not out of line with previous estimates for breast cancer and it is what the data indicates.}
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$. \rngcomment{Any comment on this result?}
 and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
@@
 
-\begin{figure}
-  \begin{center}
-  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
-  \end{center}
-  \caption{Model performance on 500X simulated whole exome. 
-  Positive predictive values vs. Recall for three different probability cutoffs. 
-  Note the restricted axes to improve readability of points}
-\label{NAR-ppv_fig}
-\end{figure}
-
 \section{DISCUSSION}
 
-Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \citep{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.
-As Griffith et al., 2015~\citep{Griffith2015} show, whole genome sequencing at approx.
-350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing.
-We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
+\rngcomment{I worry that this bit about very high depth sequencing is going to be problematic in review, because it's not a problem we're solving. We're helping, but not making huge changes.}
+%Deep sequencing is necessary to measure intra-tumor heterogeneity \citep{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.
+%Current approaches to very high depth next-generation sequencing \cite{Griffith2015}, however, require combining calls from multiple variant calling algorithms and then validating these calls via ultra-deep sequencing.
+%We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
 
-\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
-We demonstrated here that \batcave improves classification accuracy in synthetic data, leading to better more accurate variant calls at any posterior probability threshold.
-Importantly, \batcave is better calibrated than MuTect at relevant probability thresholds, allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
-We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
-Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
+Our current implementation is built on MuTect, one of the most widely used somatic variant callers.
+\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
+Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (\ref{NAR-ppv_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
+For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F).
+The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.
+
+The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
+Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
+There are, however, many other aspects of genomic context that can affect local mutation rates, including replication timing, expression level, and chromatin organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
+\rngcomment{Move individual citations to the factors they support.}
+These additional forms of context could also be incorporated into the \batcave mutation prior.
+In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors. \rngcomment{Can we back this assertion up?} 
+In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.
+
+\rngcomment{Got here...}
 
@@
 
-The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
-There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
-\bkmcomment{what to say here? Are we acknowledging a weakness, or just observing there are other types?}
-These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.
-
 These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
@@
 As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
+%We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
+%Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
 

### 2019-09-23 NAR_DRAFT.tex

--- 2019-09-20 NAR_DRAFT.tex
@@
   \end{center}
-  \bkmcomment{tumor in C is probably normal breast plus APOBEC. Mention somewhere?}
+  \bkmcomment{tumor in C is probably normal breast plus APOBEC. Mention somewhere? Fix new typo in panel B.}
   \caption{Tumor mutation profiles.
@@
   \end{center}
-  \bkmcomment{This can be smaller. regenerate with its own base\_height}
   \caption{Tumor mutation profiles.
@@
   \end{center}
-  \bkmcomment{this would probably look OK shorter too}
   \caption{Convergence of the site-specific prior to the data generating distribution.
@@
 
-\begin{figure}
-  \begin{center}
-  \includegraphics[width=.45\textwidth]{figures/ppv_wes.pdf}
-  \end{center}
-  \caption{Model performance on 500X simulated whole exomes. 
-  Positive predictive value vs. Recall for five different probability cutoffs. 
-  Note the restricted axes to improve readability of points}
-\label{NAR-ppv_fig}
-\end{figure}
-
 \section{DISCUSSION}
 
-Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \citep{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.
-As Griffith et al., 2015~\citep{Griffith2015} show, whole genome sequencing at approx.
-350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing.
-We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
+\rngcomment{I worry that this bit about very high depth sequencing is going to be problematic in review, because it's not a problem we're solving. We're helping, but not making huge changes.}
+%Deep sequencing is necessary to measure intra-tumor heterogeneity \citep{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.
+%Current approaches to very high depth next-generation sequencing \cite{Griffith2015}, however, require combining calls from multiple variant calling algorithms and then validating these calls via ultra-deep sequencing.
+%We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
 
-\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
-We demonstrated here that \batcave improves classification accuracy in synthetic data, leading to better more accurate variant calls at any posterior probability threshold.
-Importantly, \batcave is better calibrated than MuTect at relevant probability thresholds, allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
-We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
-Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
+\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
+Our current implementation is built on MuTect, one of the most widely used somatic variant callers.
+\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
+Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (\ref{NAR-ppv_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
+For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F).
+The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.
+
+The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
+Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
+There are, however, many other aspects of genomic context that can affect local mutation rates, including replication timing, expression level, and chromatin organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
+\rngcomment{Move individual citations to the factors they support.}
+These additional forms of context could also be incorporated into the \batcave mutation prior.
+In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors. \rngcomment{Can we back this assertion up?} 
+In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.
+
+\rngcomment{Got here...}
 

### 2019-09-23 NAR_DRAFT.tex
adding notes to self
Finish pass through Discussion

@@
 
-\rngcomment{Got here...}
+\batcave divides the data into two classes: high- and low-confidence variants.
+The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
+Statistically, this is an empirical Bayesian approach~\cite{XXX}. \rngcomment{Add citations, and sharpen discussion if necessary.}
+In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
+A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
+Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{TrackSigpaper}, although those differences are relatively small.
+A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
+This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
 
-\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
-In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
-Early mutations are classified with low error, while later mutations are classified with less certainty.
-We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
-Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
-Thus, \batcave is best described as an algorithm trained on early mutations.
+%\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
+%In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
+%Early mutations are classified with low error, while later mutations are classified with less certainty.
+%We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
+%Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
+%Thus, \batcave is best described as an algorithm trained on early mutations.
 
-\bkmcomment{this needs more work}
-We simulate tumors with three mutation profiles, and evaluate two real tumor data sets.
-Among these performance is best with the most concentrated simulatin profile.
-The intermediate and diffuse mutation profiles are designed to represent baseline or minimal profiles of Lung and Breast tumors respectively.
-In fact, lung and breast tumors have heterogeneous mutation profiles with a wide range of levels of concentration.
-The Acute Myeloid Leukemia we evaluated has a particularly diffuse profile, which is common for these tumors \cite{Alexandrov2019}.
+We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
+The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
+But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
+The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
+They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
+A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
+This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
+Moreover, few true negative sites are validated, so specificity is impossible to calculate. \rngcomment{Double-check my logic here...}
 
-These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
-At the same time, there is no way to compute true and false negative fractions for these data sets.
-As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
 %We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
@@
 
-We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
-MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
-However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
-For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
-Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
+Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
+We chose to build \batcave off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
+But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
+For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a uniform probability of somatic mutation at each genomic site \cite{Kim2018}.
+Replacing that uniform probability with a context-aware prior would require editing the Strelka2 source code, but is conceptually straightforward.
+\rngcomment{Is this correct, or does Strelka2 output enough for us to do it post-hoc?}
 
-The R implementation \texttt{batcaver} adds 1 second per 22,000 variants evaluated to a standard GATK best practices variant calling pipeline. 
-The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
-Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.
+Our current R implementation, \texttt{batcaver}, is computationally inexpensive, adding 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
+The majority of the computational costs is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
+Since most callers are already walking the reference genome during the calling process, extracting the trinucleotide context simultaneously would virtually eliminate the computational cost of implementing a mutational context prior.
 
-We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications.
-One size does not fit all with respect to the evidence threshold. 
-Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of precision and recall.
-We present \batcave, and the R package \texttt{batcaver} to provide the user with a more accurate sense of these trade-offs.
+Our results show that adding a mutational context prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
+Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
+Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
+Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
+Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
+
+\rngcomment{Need to rearrange Discussion logically...}
 

### 2019-09-24 NAR_DRAFT.tex
Add note about calibration paragraph in dicussion
strelka edit and citations

@@
 \rngcomment{Did MuTect fail to call anything we know is there?}
+\bkmcomment{Depends on the cutoff. This is the precision-recall curve. I can answer this for the mutect threshold in MuTect 1.1.7. Could probably answer it for the Mutect2 threshold as well.}
 
@@
 Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
-There are, however, many other aspects of genomic context that can affect local mutation rates, including replication timing, expression level, and chromatin organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
+There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}.
 \rngcomment{Move individual citations to the factors they support.}
@@
 But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
-For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a uniform probability of somatic mutation at each genomic site \cite{Kim2018}.
-Replacing that uniform probability with a context-aware prior would require editing the Strelka2 source code, but is conceptually straightforward.
+For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a constant somatic mutation probability (somatic mutation state prior) at each genomic site \cite{Kim2018}.
+Replacing that constant probability with a context-aware prior would require a more complicated manipulation of the quality scores output by Strelka than we perform for MuTect, but is conceptually straightforward.
 \rngcomment{Is this correct, or does Strelka2 output enough for us to do it post-hoc?}
+
+\rngcomment{Need paragraph in discussion emphasizing importance of calibration.}
 

### 2019-09-24 NAR_DRAFT.tex
bib and citation and text changes

@@
 There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}.
-\rngcomment{Move individual citations to the factors they support.}
 These additional forms of context could also be incorporated into the \batcave mutation prior.
-In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors. \rngcomment{Can we back this assertion up?} 
+In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors.
+\rngcomment{Can we back this assertion up?} \bkmcomment{Possibly not in this strong form.
+Most of these citations for non tri-nucleotide context are from germline, and represent things that will happen in all replications, not just tumors.
+Hodgkinson and Eyre-Walker might be a usable citation here \citep{Hodgkinson2011}, but wording is awkward.
+May not be an assertion worth making.}
 In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.
@@
 Replacing that constant probability with a context-aware prior would require a more complicated manipulation of the quality scores output by Strelka than we perform for MuTect, but is conceptually straightforward.
-\rngcomment{Is this correct, or does Strelka2 output enough for us to do it post-hoc?}
 

### 2019-09-24 NAR_DRAFT.tex
citations and things

--- 2019-09-24 NAR_DRAFT.tex
@@
 \batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
-Our current implementation is built on MuTect, one of the most widely used somatic variant callers.
+Our implementation is built on MuTect, one of the most widely used somatic variant callers.
 \batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
@@
 There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}.
-\rngcomment{Move individual citations to the factors they support.}
-These additional forms of context could also be incorporated into the \batcave mutation prior.
-In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors. \rngcomment{Can we back this assertion up?} 
+Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutation prior by observing the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
+Others, such as expression level, could be tumor-specific, but would require information not available in the variant calls to compute.
 In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.
@@
 The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
-Statistically, this is an empirical Bayesian approach~\cite{XXX}. \rngcomment{Add citations, and sharpen discussion if necessary.}
+Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}. \rngcomment{Add citations, and sharpen discussion if necessary.}
 In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
 A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
-Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{TrackSigpaper}, although those differences are relatively small.
+Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
 A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.

### 2019-09-24 NAR_DRAFT.tex
little things
empirical bayes citations

@@
 \rngcomment{Did MuTect fail to call anything we know is there?}
-\bkmcomment{Depends on the cutoff. This is the precision-recall curve. I can answer this for the mutect threshold in MuTect 1.1.7. Could probably answer it for the Mutect2 threshold as well.}
+\bkmcomment{Depends on the cutoff. This is the precision-recall curve.}
 
@@
 The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
-Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}. \rngcomment{Add citations, and sharpen discussion if necessary.}
+Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments\citep{Morris1983,Efron2014}. 
 In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.

### 2019-09-24 NAR_DRAFT.tex
correct, no true negatives
add space

@@
 The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
-Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments\citep{Morris1983,Efron2014}. 
+Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \citep{Morris1983,Efron2014}. 
 In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
@@
 This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
-Moreover, few true negative sites are validated, so specificity is impossible to calculate. \rngcomment{Double-check my logic here...}
+Moreover, few true negative sites are validated, so specificity is impossible to calculate.
 

### 2019-09-24 NAR_DRAFT.tex
remove cruft

@@
 A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
-This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
-
-%\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
-%In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
-%Early mutations are classified with low error, while later mutations are classified with less certainty.
-%We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
-%Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
-%Thus, \batcave is best described as an algorithm trained on early mutations.
+This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
 

### 2019-09-24 NAR_DRAFT.tex
Figure comments, some formatting tweaks.

--- 2019-09-24 NAR_DRAFT.tex
@@
 
+\usepackage{xspace}
 \newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
 \newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
-\newcommand{\batcave}{BATCAVE }
+\newcommand{\batcave}{BATCAVE\xspace}
 \newcommand{\beginsupplement}{%
@@
   \end{center}
-  \bkmcomment{tumor in C is probably normal breast plus APOBEC. Mention somewhere? Fix new typo in panel B.}
   \caption{Tumor mutation profiles.
@@
   \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
+  \rngcomment{Label cut off in panel A. Since this figure is now all real data, we can remove "real" from each internal label.}
    }
@@
 Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
-\begin{scriptsize}
-\begin{verbatim}
-java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
-        --reference_sequence $ref_path \
-        --dbsnp $db_snp \
-        --enable_extended_output \
-        --fraction_contamination 0.00 \
-        --tumor_f_pretest 0.00 \
-        --initial_tumor_lod -10.00 \
-        --required_maximum_alt_allele_mapping_quality_score 1 \
-        --input_file:normal $tmp_normal \
-        --input_file:tumor $tmp_tumor \
-        --out $out_path/$chr.txt \
-        --coverage_file $out_path/$chr.cov
-\end{verbatim}
-\end{scriptsize}
+\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
+        --reference\_sequence \$ref\_path 
+        --dbsnp \$db\_snp 
+        --enable\_extended\_output 
+        --fraction\_contamination 0.00 
+        --tumor\_f\_pretest 0.00 
+        --initial\_tumor\_lod -10.00 
+        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
+        --input\_file:normal \$tmp\_normal 
+        --input\_file:tumor \$tmp\_tumor 
+        --out \$out\_path/\$chr.txt 
+        --coverage\_file \$out\_path/\$chr.cov}.
 Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
@@
 For a perfectly calibrated caller, that curve would match the dashed y=x line.
-    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.}
-
+    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.
+    \rngcomment{Wasting a ton of vertical space now. Remove extraneous whitespace throughout. For panels A,B,E\&F, make \batcave curve thicker, so it's visible from behind the MuTect curve.}
+}
   \label{NAR-wes_fig}
@@
 We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-Both data sets were validated by deep targeted sequencing, yielding high quality false positive data.
-However, \texttt{only} variants called by at least one variant caller were validated.
-As a result, there is no validated measure of true/false negative calls, so we considered only precision-recall comparisons for these data.
-\rngcomment{Is  a correct statement?}\bkmcomment{Yes, definitely true.}
+Both data sets were validated by deep targeted sequencing, yielding high quality true and false positive data.
+However, only variants called by at least one variant caller were validated.
+As a result, there is no validated measure of true or false negative calls, so we considered only precision-recall comparisons for these data.
 

### 2019-09-24 NAR_DRAFT.tex
Minor introduction edits

@@
 Many methods have been developed for calling somatic mutations from NGS data.
-The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
+The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
 joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
+\rngcomment{Connect citations to individual tools. (See MuTect1 and Varscan2 examples earlier in paragraph.}
 These methods vary in their complexity and specific focus.
-But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.
+But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.
 
@@
 %Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
-The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
+If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
 Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, Figure~\ref{NAR-realsigfig}, shows mutation profiles from several tumors.
-These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
+For example, the mutation profiles differs substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
@@
 Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
-Our algorithm estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
-Our implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
+Our algorithm first estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
+Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
 Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
-Our algorithm is simple, computationally inexpensive, and can be integrated into numerous other variant callers.
+Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
 Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.
@@
   \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
-  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
+  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
   \rngcomment{Label cut off in panel A. Since this figure is now all real data, we can remove "real" from each internal label.}

### 2019-09-24 NAR_DRAFT.tex
some calibration stuff

--- 2019-09-24 NAR_DRAFT.tex
@@
 
+We focus on the calibration benefits of \batcave because calibration is important for both clinical and research communities.
+\begin{itemize}
+  \item Research/Discovery
+  \begin{itemize}
+    \item Manual review is time consuming, and the number of mutations can be overwhelming at lower frequencies. Calibration reduces wasted effort.
+    \item Drug discovery is expensive and effort must be focused.
+Finding common low frequency alleles is important, and having a calibrated probability for each tumor a variant is found in would be useful.
+  \end{itemize}
+  \item clinical
+  \begin{itemize}
+    \item many resistance mutations are present at low frequencies in primary tumor.
+If a tumor is sequenced and you can assign a probability to the mutation, it may be possible to indicate treatment with the drug for the resistance mutation simultaneous to the first line treatment.
+For instance if the mutation has a 80\% probabiliity of being present, yes, but if 20\% no.
+Balance of harm from treating for a mutation that isn't there.
+  \end{itemize}
+
+\end{itemize}
+
 Our current R implementation, \texttt{batcaver}, is computationally inexpensive, adding 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.

### 2019-09-24 NAR_DRAFT.tex
comment something
More minor revisions and comments.

--- 2019-09-24 NAR_DRAFT.tex
@@
 Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
-Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
+Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
 Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
@@
   \label{eqn:detailed_prior}
-  \begin{array}{rcl}
-  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
-                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
-  \end{array}
+  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
 \end{equation}
@@
 
-We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}\textbf{A}-\textbf{D}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
 Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
@@
 \end{equation}
-In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.
+In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.
 
@@
 \begin{equation}
-    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
+    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\},
 \end{equation}
+and $I$ is the indicator function.
 Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
@@
 \end{equation}
+
 The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
-The R implementation of \batcave ships with pre-computed tables for both whole exomes and whole genomes.
+The R implementation of \batcave ships with pre-computed tables for both human whole exomes and whole genomes.
 
 \subsection{Estimation of the mutation rate}
-The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base mutation rate $\mu$.
-In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
+The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
+In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
 \begin{equation}
   \label{eqn:mut_rate}
-  \mathrm{M}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
+  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
 \end{equation}
@@
 We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
-We can then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
-We can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
+We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
+We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
 In the R implementation of \batcave, $f_{min}$ is a free parameter.
@@
 The package leverages the Bioconductor packages \textrm{BSgenome}, \textrm{GenomicAlignments}, \textrm{VariantAnnotation}, and \textrm{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
-Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
+Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.  \rngcomment{Add citations to all tools that have them.}
 
@@
 \subsection{Tumor simulations}
-We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
+We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of alleles.
 Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
 Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
-The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
-The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
+The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
+The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
+The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
 Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
@@
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
+\rngcomment{What are we doing about the multi-regional nature of this data? Are we combining across regions somehow?}
 We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
-\rngcomment{Did MuTect fail to call anything we know is there?}
-\bkmcomment{Depends on the cutoff. This is the precision-recall curve.}
 
 \subsection{Calibration metric}
-The quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
-Briefly, a loess-smoothed regression was fit by regressing the binary (True=1,False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
-For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$, and the Integrated Calibration Index is the average distance between the fit and that line.
+To quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
+Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
+For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
+The Integrated Calibration Index is the average distance between the fit and that line.
+\rngcomment{This can't be true, otherwise perfect calibration would be ICI=1... What is it?}
 
@@
 We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites in to high- and low-confidence groups.
+\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
 The mutation profile is estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
-The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500x exome and 2,000 seconds for a 100x whole genome.
+The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.
 
-To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
+To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
 The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}\textbf{A-C}).
@@
 In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
+Because our focus is on evaluating the statistical calling model, we compute all metrics using only those potential variants that pass MuTect's heuristic filters.
 
@@
   \caption{Convergence of the site-specific prior to the data generating distribution.
-Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
+Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations for whole exomes.
+Convergence for whole genomes is similar.
+  \rngcomment{Omit panel B and remove A label. These are similar enough we don't need to show both.}
   }
@@
 
+% XXX: RNG stopped here...
+
 \subsection{Convergence of the prior}
 To improve performance, the context-dependent prior probability of mutation must converge to a good representation of the data generating distribution within the set of high-confidence mutations.
-When applied to simulated whole exome and whole genome data, the prior converges within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
-For comparison, in our simulated data sets the number of high-confidence mutations ranges between 1500 and 5000, and for a real acute myeloid leukemia it is over 17,000~\cite{Griffith2015}.
+When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
+For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.
 

### 2019-09-24 NAR_DRAFT.tex

@@
 \rngcomment{Need paragraph in discussion emphasizing importance of calibration.}
+
+We focus on the calibration benefits of \batcave because calibration is important for both clinical and research communities.
+\begin{itemize}
+  \item Research/Discovery
+  \begin{itemize}
+    \item Manual review is time consuming, and the number of mutations can be overwhelming at lower frequencies. Calibration reduces wasted effort.
+    \item Drug discovery is expensive and effort must be focused.
+Finding common low frequency alleles is important, and having a calibrated probability for each tumor a variant is found in would be useful.
+  \end{itemize}
+  \item clinical
+  \begin{itemize}
+    \item many resistance mutations are present at low frequencies in primary tumor.
+If a tumor is sequenced and you can assign a probability to the mutation, it may be possible to indicate treatment with the drug for the resistance mutation simultaneous to the first line treatment.
+For instance if the mutation has a 80\% probabiliity of being present, yes, but if 20\% no.
+Balance of harm from treating for a mutation that isn't there.
+  \end{itemize}
+
+\end{itemize}
+\bkmcomment{begin needed paragraph above}
 

### 2019-09-25 NAR_DRAFT.tex
kl divergence is not symmetric

@@
   \caption{Convergence of the site-specific prior to the data generating distribution.
-Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations for whole exomes.
+Plotted is the Kullback-Leibler divergence between the simulated and estimated profiles versus number of incorporated mutations for whole exomes.
 Convergence for whole genomes is similar.
-  \rngcomment{Omit panel B and remove A label. These are similar enough we don't need to show both.}
   }
@@
     \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.
-    \rngcomment{Wasting a ton of vertical space now. Remove extraneous whitespace throughout. For panels A,B,E\&F, make \batcave curve thicker, so it's visible from behind the MuTect curve.}
 }

### 2019-09-25 NAR_DRAFT.tex
explain real breast cancer mor

@@
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-\rngcomment{What are we doing about the multi-regional nature of this data? Are we combining across regions somehow?}
+We ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
 We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

### 2019-09-25 NAR_DRAFT.tex
precise definition of the ICI

@@
 For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
-The Integrated Calibration Index is the average distance between the fit and that line.
-\rngcomment{This can't be true, otherwise perfect calibration would be ICI=1... What is it?}
+The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.
 

### 2019-09-25 NAR_DRAFT.tex
fix problems with figs 1 and 2

@@
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$.
-\rngcomment{Any comment on this result?}\bkmcomment{Noone, as far as I can tell has done a systematic calculation for TCGA.
-\cite{Williams2018} is only out in 2018.
-On the basis of tumor mutation burden studies breast cancer is one of the slowest mutators, but there is no way to go from TMB to mutation rate since it depends on depth/frequency.
-The one breast cancer studied in Williams 2018 the estimate at just over 1e-7.
-that is a whole genome whereas ours is an exome, and theirs is deeper.
-My argument would be that it is not out of line with previous estimates for breast cancer and it is what the data indicates.}
+We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is within the range of observed mutation rates for breast cancers \citep{Alexandrov2019}, and in line with the low number of validated somatic mutations present in these tumors.
 and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).

### 2019-09-25 NAR_DRAFT.tex
fix intro citations

@@
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
-joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
-\rngcomment{Connect citations to individual tools. (See MuTect1 and Varscan2 examples earlier in paragraph.}
+joint genotype analysis (SomaticSniper \citep{Larson2012}, JointSNVMix2 \citep{Roth2012a}, Seurat \citep{Christoforides2013}, CaVEMan \citep{Jones2016}, and MuClone \citep{Dorri2019}), allele frequency-based analysis (Strelka \citep{Saunders2012}, LoFreq \citep{Wilm2012}, EBCall \citep{Shiraishi2013b}, deepSNV \citep{Gerstung2012}, LoLoPicker \citep{Carrot-Zhang2017}, and MuSE \citep{Fan2016}), and ensemble and deep learning methods (MutationSeq \citep{Ding2012}, BAYSIC \citep{Cantarel2014}, SomaticSeq \citep{Fang2015}, and SNooPer \citep{Spinella2016}).
 These methods vary in their complexity and specific focus.

### 2019-09-25 NAR_DRAFT.tex
wrong citations in fig 1 fixed
citations for bioconductor packages

@@
   \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
-  \rngcomment{Label cut off in panel A. Since this figure is now all real data, we can remove "real" from each internal label.}
    }
@@
 We have implemented the \batcave algorithm as an R package (batcaver).
-The package leverages the Bioconductor packages \textrm{BSgenome}, \textrm{GenomicAlignments}, \textrm{VariantAnnotation}, and \textrm{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
-Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.  \rngcomment{Add citations to all tools that have them.}
+The package leverages the Bioconductor packages \textrm{BSgenome} \citep{Pages2019}, \textrm{GenomicAlignments} \citep{Lawrence2013}, \textrm{VariantAnnotation} \citep{Obenchain2014}, and \textrm{SomaticSignatures} \citep{Gehring2015} fast and memory-efficient variant annotation and genomic context identification.
+Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
 

### 2019-09-26 NAR_DRAFT.tex
paragraph on calibration in discussion

@@
 
-We focus on the calibration benefits of \batcave because calibration is important for both clinical and research communities.
-\begin{itemize}
-  \item Research/Discovery
-  \begin{itemize}
-    \item Manual review is time consuming, and the number of mutations can be overwhelming at lower frequencies. Calibration reduces wasted effort.
-    \item Drug discovery is expensive and effort must be focused.
-Finding common low frequency alleles is important, and having a calibrated probability for each tumor a variant is found in would be useful.
-  \end{itemize}
-  \item clinical
-  \begin{itemize}
-    \item many resistance mutations are present at low frequencies in primary tumor.
-If a tumor is sequenced and you can assign a probability to the mutation, it may be possible to indicate treatment with the drug for the resistance mutation simultaneous to the first line treatment.
-For instance if the mutation has a 80\% probabiliity of being present, yes, but if 20\% no.
-Balance of harm from treating for a mutation that isn't there.
-  \end{itemize}
+The improved calibration of \batcave provides several advantages.
+Cancer genomics for research and drug discovery is an important yet costly endeavour, and manual review of variants is often required to limit the number of false leads followed.
+Improved calibration allows researchers to select a tolerable false positive rate, and focus efforts on a set of variants with known probability of being present.
+In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs, and well calibrated probabilities facilitate the use of risk models in the choice of treatment rather than an all or nothing approach.
+\bkmcomment{I mean by "risk models", if you can assign a probability of the mutation being present you can make informed tradeoffs between likelihood of benefit and risk of harm for a particular targeted therapy.}
+A properly calibrated model implies a known rate of false positives.
+One reason why tumors are not routinely sequenced to high depth is that as sequencing depth increases, false positive rate increases without bound \citep{Cibulskis2013}.
+Proper calibration will allow deeper sequencing without ballooning false positive rates.
 
-\end{itemize}
-\bkmcomment{begin needed paragraph above}
+\bkmcomment{begin needed paragraph above.}
 

### 2019-09-26 NAR_DRAFT.tex
calculation

@@
 One reason why tumors are not routinely sequenced to high depth is that as sequencing depth increases, false positive rate increases without bound \citep{Cibulskis2013}.
+The International Cancer Genome Consortium recommends that catalogues of somatic mutation target a true positive rate of 95\%, and an acquisition rate of 80\% of all mutations present in a sample \texttt{https://icgc.org/icgc/goals-structure-policies-guidelines/e8-genome-analyses}.
+Assuming a sample size of 1 million cells, and a per-base mutation rate of $3 \dot 10^{-8}$, implies 10 million mutations per sample in a neutrally evolving tumor.
+A naive application of Eqn.
+\ref{eqn:mut_rate} implies a minimum variant allele frequency of \bkmcomment{calculate this, will be very low.
+First pass is vaf >= $10^{-5}$}, requiring sequencing depths far in excess of current standards.
 Proper calibration will allow deeper sequencing without ballooning false positive rates.

### 2019-09-26 NAR_DRAFT.tex
calibration

@@
 A naive application of Eqn.
-\ref{eqn:mut_rate} implies a minimum variant allele frequency of \bkmcomment{calculate this, will be very low.
-First pass is vaf >= $10^{-5}$}, requiring sequencing depths far in excess of current standards.
+\ref{eqn:mut_rate} implies a minimum variant allele frequency of \bkmcomment{calculate this, will be very low.}, requiring sequencing depths far in excess of current standards.
 Proper calibration will allow deeper sequencing without ballooning false positive rates.

### 2019-09-26 NAR_DRAFT.tex
Results tweaking

--- 2019-09-26 NAR_DRAFT.tex
@@
 To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
-The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}\textbf{A-C}).
+The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}).
 We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
 In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-Because our focus is on evaluating the statistical calling model, we compute all metrics using only those potential variants that pass MuTect's heuristic filters.
+Because our focus is on evaluating the statistical calling model, we computed all test metrics using only those potential variants that passed MuTect's heuristic filters and entered the statistical model.
 
@@
 
-% XXX: RNG stopped here...
-
 \subsection{Convergence of the prior}
-To improve performance, the context-dependent prior probability of mutation must converge to a good representation of the data generating distribution within the set of high-confidence mutations.
+To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
 When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
@@
     \caption{Variant-calling performance on both simulated and real data.
-    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation signatures.
+    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation profiles.
     \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated.
@@
     \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.
+    \rngcomment{Make \batcave lines thin, MuTect lines thick, and put MuTect in back in all comparison plots.}
 }
@@
   \tableparts{%
-  \caption{Performance characteristics as measured for all data sets.}
+  \caption{Variant calling metrics for all data sets.}
   \label{table:01}%
   }{%
-  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
+  \begin{tabular}{lllllllllll}
   \toprule
-  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
+  Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
   \\
@@
   \botrule
-  \end{tabular*}%
+  \end{tabular}%
   }
-  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
+  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
  \end{table*}
@@
 We assessed classification performance using the areas under both the receiver operating characteristic and the precision-recall curves, because the classes are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
-By both metrics \batcave outperforms MuTect (Fig.~\ref{NAR-wes_fig}\textbf{A}\&\textbf{B} and Table~\ref{table:01}).
-The extent of the performance difference is dependent on both the sequencing depth and the concentration of the mutation signature.
-Deeper sequencing and more concentrated mutation signatures increase the performance advantage of \batcave.
+By both metrics \batcave outperforms MuTect (Fig.~\ref{NAR-wes_fig}A\&B, Fig.~\ref{NAR-wgs_fig}A\&B, and Table~\ref{table:01}).
+The extent of the performance difference is dependent on both the sequencing depth and the concentration of the mutation profile.
+Deeper sequencing and more concentrated mutation profiles increase the performance advantage of \batcave.
 
 For all simulated tumors, the estimated mutation rate is approximately $3\cdot10^{-7}$ (Table~\ref{table:01}), which is lower than the simulated rate of $3\cdot10^{-6}$.
-This is due to restrictions with BAMSurgeon, such as sequencing depth and quality, that prevent 100\% of simulated variants from being inserted into the reads.
+This is likely due to restrictions within BAMSurgeon, such as sequencing depth and quality, that prevent 100\% of simulated variants from being inserted into the reads.
 
-We also assessed the calibration of \batcave, assessing the likelihood that a potential variant with a given posterior probability is actually a true variant.
-We measured overall calibration performance using the Integrated Calibration Index (ICI) \cite{Austin2019}, which measures the difference between predicted and observed probabilities, weighted by the density of the predicted probabilities.
-This metric is particularly useful in our case, because the density of posterior probabilities is bi-modal (Fig.~\ref{NAR-wes_fig}C\&D).
-A large fraction of true negative variants have posterior probabilities less than $10^{-4}$, far below any meaningful threshold, so we evaluate calibration only on potential variants with posterior probability greater than 0.01.
-For these potential variants, \batcave tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D) while decreasing probabilities of low probability but false variants.
-For 500X tumors, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
-This is because that there are few low probability true positive variants in tumors sequenced to 100X depth.
-As with the other metrics, the advantage of \batcave increases with the concentration of the mutation profile the sequencing depth.
+We also assessed calibration, the likelihood that a potential variant with a given posterior probability is actually a true variant.
+We measured overall calibration performance using the Integrated Calibration Index (ICI) \cite{Austin2019}, which integrates the difference between predicted and observed probabilities, weighted by the density of the predicted probabilities.
+This metric is particularly useful in our case, because the density of posterior probabilities is bi-modal (Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D).
+A large fraction of true negative variants have posterior probabilities less than $10^{-4}$, far below any meaningful threshold, so we evaluated calibration only on potential variants with posterior probability greater than 0.01.
+For these potential variants, \batcave tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D) while decreasing probabilities of low probability but false variants.
+For 500X exomes, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities (Figure \ref{NAR-wes_fig} and Table \ref{table:01}).
+For 100X whole genomes, the calibration of \batcave is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}), likely because there are few low probability true positive variants in tumors sequenced to 100X depth.
+As with the other metrics, the advantage of \batcave increases with the concentration of the mutation profile and the sequencing depth.
 
@@
   \end{center}
-  \caption{Calibration of posterior probability. 
-  For realistic thresholds, posterior probabilities are more accurately calibrated for \batcave than MuTect, no matter how concentration the mutation profile is.}
+  \caption{Calibration of posterior probability for 500X exomes. 
+  For realistic thresholds, posterior probabilities are more accurately calibrated for \batcave than MuTect, no matter the concentration of the mutation profile.
+  \rngcomment{Need to be careful about what is represented here. Here precision and recall are for all variants \textbf{above} the threshold, correct?}
+  }
 \label{NAR-ppv_fig}
@@
 For variants with posterior probability between 60 and 90\%, \batcave is substantially better calibrated than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as posterior probability increases.
+For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the posterior probability increases.
 The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
-At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
+At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect. \rngcomment{What does this mean?}
 %Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
@@
 \subsection{Tests using real tumor data}
-We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-Both data sets were validated by deep targeted sequencing, yielding high quality true and false positive data.
+We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the express purpose of evaluating tumor variant calling pipelines, yielding high quality true and false positive data~\cite{Griffith2015,Shi2018}.
 However, only variants called by at least one variant caller were validated.
-As a result, there is no validated measure of true or false negative calls, so we considered only precision-recall comparisons for these data.
+As a result, there are no validated true or false negative calls, so we considered only precision-recall comparisons for these data.
 
 Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is inline with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
+We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
 For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
 
-Shi et al.\ performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$ and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
+Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is within the range of observed mutation rates for breast cancers \citep{Alexandrov2019}, and in line with the low number of validated somatic mutations present in these tumors.
-and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
+We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \citep{Alexandrov2019} and with the low number of validated somatic mutations.
 For the validated variants, MuTect and \batcave yield almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
@@
 
-\rngcomment{I worry that this bit about very high depth sequencing is going to be problematic in review, because it's not a problem we're solving. We're helping, but not making huge changes.}
 %Deep sequencing is necessary to measure intra-tumor heterogeneity \citep{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.
@@
 Our implementation is built on MuTect, one of the most widely used somatic variant callers.
-\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
-Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (\ref{NAR-ppv_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
-For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F).
+\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D,  \ref{NAR-wgs_fig}, and Table~\ref{table:01}) across the entire range of recall and specificity.
+Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (Fig.~\ref{NAR-ppv_fig} and~\ref{NAR-ppv_wgs_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
+For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F and Table~\ref{table:01}).
 The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.
@@
   \caption{Model performance on 100X simulated whole genome.
-  \textbf{(A)} Precision recall curves and \textbf{(B)} Reciever operating characteristic curves for 3 mutation signatures.
+  \textbf{(A)} Precision recall curves and \textbf{(B)} Reciever operating characteristic curves for 3 mutation profiles.
   Distribution of estimated true positive probabilities for true positive  variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.

### 2019-09-26 NAR_DRAFT.tex

@@
 One reason why tumors are not routinely sequenced to high depth is that as sequencing depth increases, false positive rate increases without bound \citep{Cibulskis2013}.
+The International Cancer Genome Consortium recommends that catalogues of somatic mutation target a true positive rate of 95\%, and an acquisition rate of 80\% of all mutations present in a sample \texttt{https://icgc.org/icgc/goals-structure-policies-guidelines/e8-genome-analyses}.
+Assuming a sample size of 1 million cells, and a per-base mutation rate of $3 \dot 10^{-8}$, implies 10 million mutations per sample in a neutrally evolving tumor.
+A naive application of Eqn.
+\ref{eqn:mut_rate} implies a minimum variant allele frequency of \bkmcomment{calculate this, will be very low.}, requiring sequencing depths far in excess of current standards.
 Proper calibration will allow deeper sequencing without ballooning false positive rates.

### 2019-09-26 NAR_DRAFT.tex
Discussion rearrangement.

@@
 
+We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
+The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively.
+And the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
+But mutational profiles are highly heterogeneous, so concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
+The two real data sets we considered are among the few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
+They happen, however, to have diffuse mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
+A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
+This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
+Moreover, few true negative sites are validated, so specificity and calibration are impossible to calculate.
+Deep sequencing experiments that validate random samples of uncalled potential variants would give much-needed insight into the differences among statistical models in variant calling.
+
+Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
+We chose to build \batcave off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
+But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
+For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a constant somatic mutation probability at each genomic site \cite{Kim2018}.
+Replacing that constant probability with a mutational prior would require a more complicated manipulation of the quality scores output by Strelka than for MuTect but is conceptually straightforward.
+
+The \batcave algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
+The majority of the computational costs is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
+Since most callers are already walking the reference genome during the calling process, extracting the trinucleotide context simultaneously would virtually eliminate the computational cost of implementing a mutational prior.
+
 The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
@@
 There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}.
-Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutation prior by observing the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
+Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutational prior using the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
 Others, such as expression level, could be tumor-specific, but would require information not available in the variant calls to compute.
-In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.
+In the long run, we believe that incorporating more tumor biology into variant calling models will continue to improve performance.
 
@@
 
-We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
-The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
-But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
-The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
-They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
-A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
-This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
-Moreover, few true negative sites are validated, so specificity is impossible to calculate.
-
-%We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
-%Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
-
-Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
-We chose to build \batcave off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
-But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
-For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a constant somatic mutation probability (somatic mutation state prior) at each genomic site \cite{Kim2018}.
-Replacing that constant probability with a context-aware prior would require a more complicated manipulation of the quality scores output by Strelka than we perform for MuTect, but is conceptually straightforward.
-
 \rngcomment{Need paragraph in discussion emphasizing importance of calibration.}
-
 The improved calibration of \batcave provides several advantages.
@@
 
-\bkmcomment{begin needed paragraph above.}
-
-Our current R implementation, \texttt{batcaver}, is computationally inexpensive, adding 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
-The majority of the computational costs is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
-Since most callers are already walking the reference genome during the calling process, extracting the trinucleotide context simultaneously would virtually eliminate the computational cost of implementing a mutational context prior.
-
 Our results show that adding a mutational context prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
@@
 Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
-
-\rngcomment{Need to rearrange Discussion logically...}
 

### 2019-09-30 NAR_DRAFT.tex
ick, but moving in the right direction

@@
 
-In practice, accurate calibration is most important for variants with high posterior probability.
+In practice, variant calling is a decision problem, in which a threshold must be selected above which we accept a variant call.
+The proper threshold is ideally a choice the user makes based on their particular needs.
 For variants with posterior probability between 60 and 90\%, \batcave is substantially better calibrated than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).

### 2019-09-30 NAR_DRAFT.tex
working on wording for calibration

--- 2019-09-26 NAR_DRAFT.tex
@@
   \end{center}
-  \caption{Calibration of posterior probability for 500X exomes. 
-  For realistic thresholds, posterior probabilities are more accurately calibrated for \batcave than MuTect, no matter the concentration of the mutation profile.
-  \rngcomment{Need to be careful about what is represented here. Here precision and recall are for all variants \textbf{above} the threshold, correct?}
+  \caption{Performance of various posterior probability thresholds in 500X exomes. 
+  For realistic posterior probability thresholds,  positive predictive values are greater for \batcave than MuTect, no matter the concentration of the mutation profile.
   }
@@
 
-In practice, accurate calibration is most important for variants with high posterior probability.
-For variants with posterior probability between 60 and 90\%, \batcave is substantially better calibrated than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+\bkmcomment{I think there is a reasonable wording here in terms of positive predictive value.
+Think of the threshold as the posterior class probability above which we set all probabilities to 1, and below to zero?
+Getting better below but I still need to think harder.}
+In practice, variant calling is a decision problem, in which a threshold score must be selected above which a variant call is made.
+The proper threshold is ideally a choice the user makes based on their particular needs, which often balance precision with recall.
+\batcave provides posterior probabilities averaged not only across the whole range of possible probabilities, but in particular at higher probability ranges applicable in most situations.
+For posterior probability thresholds between 60 and 90\%, \batcave exhibits substantially better average calibration than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+As a result, \batcave posterior probabilities provide a useful way to select tumor-specific thresholds. 
 For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the posterior probability increases.
-The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
-At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect. \rngcomment{What does this mean?}
+%The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
+At any posterior probability threshold \batcave has recall better than MuTect.
 %Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.

### 2019-10-01 NAR_DRAFT.tex
part of that was just nonsense

@@
 The proper threshold is ideally a choice the user makes based on their particular needs, which often balance precision with recall.
-\batcave provides posterior probabilities averaged not only across the whole range of possible probabilities, but in particular at higher probability ranges applicable in most situations.
+\batcave provides posterior probabilities that are well calibrated when averaged not only across the whole range of possible probabilities, but in particular at higher probability ranges applicable in most situations.
 For posterior probability thresholds between 60 and 90\%, \batcave exhibits substantially better average calibration than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).

### 2019-10-01 NAR_DRAFT.tex
new language in the results section

--- 2019-09-30 NAR_DRAFT.tex
@@
 
-\bkmcomment{I think there is a reasonable wording here in terms of positive predictive value.
-Think of the threshold as the posterior class probability above which we set all probabilities to 1, and below to zero?
-Getting better below but I still need to think harder.}
-In practice, variant calling is a decision problem, in which a threshold score must be selected above which a variant call is made.
+In practice, variant calling is a classification problem, in which a threshold score must be selected above which a variant call is made.
 The proper threshold is ideally a choice the user makes based on their particular needs, which often balance precision with recall.
-\batcave provides posterior probabilities averaged not only across the whole range of possible probabilities, but in particular at higher probability ranges applicable in most situations.
-For posterior probability thresholds between 60 and 90\%, \batcave exhibits substantially better average calibration than MuTect (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+The posterior probability estimates provided by \batcave allow the user to select a tumor-specific threshold.
+For posterior probability thresholds between 60 and 90\%, \batcave exhibits positive predictive values that are relatively close to the threshold (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+In other words, selecting a minimum posterior probability creates a classifier with positive predictive value close to that threshold.
+On the other hand, the posterior probabilities from MuTect provide a poor estimate of positive predictive value (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the posterior probability increases.
 As a result, \batcave posterior probabilities provide a useful way to select tumor-specific thresholds. 
-For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the posterior probability increases.
+At any posterior probability threshold \batcave has recall better than MuTect.
 %The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
-At any posterior probability threshold \batcave has recall better than MuTect.
 %Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.

### 2019-10-01 NAR_DRAFT.tex
change to discussion.

@@
 \rngcomment{Need paragraph in discussion emphasizing importance of calibration.}
-The improved calibration of \batcave provides several advantages.
+The improved calibration of \batcave provides several advantages, including a tumor-specific means to select a threshold for positive classification that exhibits interpretable positive predictive value properties.
 Cancer genomics for research and drug discovery is an important yet costly endeavour, and manual review of variants is often required to limit the number of false leads followed.
@@
 Proper calibration will allow deeper sequencing without ballooning false positive rates.
+\bkmcomment{we can talk about what should actually survive from this previous section. I like the specific example but may be too wordy here.}
 

### 2019-10-01 NAR_DRAFT.tex
comment in discussion

@@
 Proper calibration will allow deeper sequencing without ballooning false positive rates.
-\bkmcomment{we can talk about what should actually survive from this previous section. I like the specific example but may be too wordy here.}
+\bkmcomment{we can talk about what should actually survive from this previous section.
+I like the specific example but may be too wordy here.
+One of the things I have tried to do is establish that ppv qualities arise from superior calibration, and that allows the use of calibration as short hand here in the discussion.
+This may not be well done.}
 

### 2019-10-01 NAR_DRAFT.tex
stab at fixing figure legend

@@
   \caption{Performance of various posterior probability thresholds in 500X exomes. 
-  For realistic posterior probability thresholds,  positive predictive values are greater for \batcave than MuTect, no matter the concentration of the mutation profile.
+  For realistic posterior probability thresholds,  \batcave has a positive predictive value that is better calibrated than MuTect, no matter the concentration of the mutation profile.
   }

### 2019-10-01 NAR_DRAFT.tex
I put the frequency in there. check results

@@
 Assuming a sample size of 1 million cells, and a per-base mutation rate of $3 \dot 10^{-8}$, implies 10 million mutations per sample in a neutrally evolving tumor.
-A naive application of Eqn.
-\ref{eqn:mut_rate} implies a minimum variant allele frequency of \bkmcomment{calculate this, will be very low.}, requiring sequencing depths far in excess of current standards.
+A naive application of Eqn. \ref{eqn:mut_rate} implies a minimum variant allele frequency of around .00001, requiring sequencing depths far in excess of current practice.
 Proper calibration will allow deeper sequencing without ballooning false positive rates.

### 2019-10-02 NAR_DRAFT.tex
remove comment
Edits to new calibration paragraphs.

@@
 
-\subsection{Convergence of the prior}
+\subsection{Tests using simulated data}
+
 To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
@@
     \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.
-    \rngcomment{Make \batcave lines thin, MuTect lines thick, and put MuTect in back in all comparison plots.}
 }
@@
 
-\subsection{Tests using simulated data}
 We assessed classification performance using the areas under both the receiver operating characteristic and the precision-recall curves, because the classes are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
@@
 
-In practice, variant calling is a classification problem, in which a threshold score must be selected above which a variant call is made.
-The proper threshold is ideally a choice the user makes based on their particular needs, which often balance precision with recall.
-The posterior probability estimates provided by \batcave allow the user to select a tumor-specific threshold.
-For posterior probability thresholds between 60 and 90\%, \batcave exhibits positive predictive values that are relatively close to the threshold (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-In other words, selecting a minimum posterior probability creates a classifier with positive predictive value close to that threshold.
-On the other hand, the posterior probabilities from MuTect provide a poor estimate of positive predictive value (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-For any posterior probability above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the posterior probability increases.
-As a result, \batcave posterior probabilities provide a useful way to select tumor-specific thresholds. 
-At any posterior probability threshold \batcave has recall better than MuTect.
+In practice, variant callers are typically used with a threshold score above which a variant is called.
+The user's choice of threshold ideally meets their need to balance precision and recall; accurate posterior probability estimates enable an informed choice.
+For posterior probability thresholds between 60 and 90\%, the precision of \batcave calls is similar to the chosen threshold (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+For this range of thresholds, however, the posterior probabilities from MuTect poorly predict precision (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
+For any posterior probability threshold above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the threshold increases.
+The cost of MuTect's compressed range of posterior probabilities is recall; at any posterior probability threshold \batcave has recall better than MuTect.
+Consequently, \batcave posterior probabilities are more informative than MuTect's with regard to choosing a calling threshold.
 %The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
@@
 
+The improved calibration of \batcave posterior probabilities compared to MuTect provides several advantages.
+In practice, called variants are often manually reviewed to further reduce false positives \cite{XXX}. \rngcomment{Citations needed throughout.}
+Improved calibration enables users to focus review on the most questionable variants.
+In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs \cite{XXX}.
+Well-calibrated posterior probabilities facilitate the use of probabilistic risk models in the choice of treatment \cite{XXX}, rather than an all or nothing approach.
+For research purposes, the International Cancer Genome Consortium recommends that catalogs of somatic mutations target a precision of 95\% and a recall of 80\% \cite{XXX}.
+\rngcomment{Put in real citation... https://icgc.org/icgc/goals-structure-policies-guidelines/e8-genome-analyses}
+Achieving this goal while minimizing cost demands well-calibrated posterior probabilities.
+
 Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
@@
 This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
-
-\rngcomment{Need paragraph in discussion emphasizing importance of calibration.}
-The improved calibration of \batcave provides several advantages, including a tumor-specific means to select a threshold for positive classification that exhibits interpretable positive predictive value properties.
-Cancer genomics for research and drug discovery is an important yet costly endeavour, and manual review of variants is often required to limit the number of false leads followed.
-Improved calibration allows researchers to select a tolerable false positive rate, and focus efforts on a set of variants with known probability of being present.
-In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs, and well calibrated probabilities facilitate the use of risk models in the choice of treatment rather than an all or nothing approach.
-\bkmcomment{I mean by "risk models", if you can assign a probability of the mutation being present you can make informed tradeoffs between likelihood of benefit and risk of harm for a particular targeted therapy.}
-A properly calibrated model implies a known rate of false positives.
-One reason why tumors are not routinely sequenced to high depth is that as sequencing depth increases, false positive rate increases without bound \citep{Cibulskis2013}.
-The International Cancer Genome Consortium recommends that catalogues of somatic mutation target a true positive rate of 95\%, and an acquisition rate of 80\% of all mutations present in a sample \texttt{https://icgc.org/icgc/goals-structure-policies-guidelines/e8-genome-analyses}.
-Assuming a sample size of 1 million cells, and a per-base mutation rate of $3 \dot 10^{-8}$, implies 10 million mutations per sample in a neutrally evolving tumor.
-A naive application of Eqn. \ref{eqn:mut_rate} implies a minimum variant allele frequency of around .00001, requiring sequencing depths far in excess of current practice.
-Proper calibration will allow deeper sequencing without ballooning false positive rates.
-\bkmcomment{we can talk about what should actually survive from this previous section.
-I like the specific example but may be too wordy here.
-One of the things I have tried to do is establish that ppv qualities arise from superior calibration, and that allows the use of calibration as short hand here in the discussion.
-This may not be well done.}
 

### 2019-10-03 NAR_DRAFT.tex
supplementary figures
zenodo id

@@
 \section{Software Availability}
-\bkmcomment{working on this today}
 
-github for source code
-
-zenodo for figures, simulations
-
-zenodo for package at time of bioarxiv
+The \texttt{batcaver} R package can be downloaded or installed from \texttt{github.com/bmannakee/batcaver}
+The version of \texttt{batcaver} used to generate results, and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}
 
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/fig_wgs.png}
+  \includegraphics[width=\textwidth]{figures/fig_wgs.pdf}
   \end{center}
@@
   \caption{Model performance on 100X simulated whole genome. 
-  Positive predictive values vs. Recall for five different probability cutoffs. 
+  Precision vs. Recall for five different probability cutoffs. 
   Note the restricted axes to improve readability of points.}

### 2019-10-03 NAR_DRAFT.tex
citations and python zenodo

@@
 The improved calibration of \batcave posterior probabilities compared to MuTect provides several advantages.
-In practice, called variants are often manually reviewed to further reduce false positives \cite{XXX}. \rngcomment{Citations needed throughout.}
+In practice, called variants are often manually reviewed to further reduce false positives \cite{Barnell2019}.
 Improved calibration enables users to focus review on the most questionable variants.
-In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs \cite{XXX}.
-Well-calibrated posterior probabilities facilitate the use of probabilistic risk models in the choice of treatment \cite{XXX}, rather than an all or nothing approach.
-For research purposes, the International Cancer Genome Consortium recommends that catalogs of somatic mutations target a precision of 95\% and a recall of 80\% \cite{XXX}.
-\rngcomment{Put in real citation... https://icgc.org/icgc/goals-structure-policies-guidelines/e8-genome-analyses}
+In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs \cite{Boutros2015a}.
+%Well-calibrated posterior probabilities facilitate the use of probabilistic risk models in the choice of treatment \cite{XXX}, rather than an all or nothing approach.
+For research purposes, the International Cancer Genome Consortium recommends that catalogs of somatic mutations target a precision of 95\% and a recall of 80\% \cite{Icgc}.
 Achieving this goal while minimizing cost demands well-calibrated posterior probabilities.
@@
 The version of \texttt{batcaver} used to generate results, and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}
+Python code used to generate simulated tumors has been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471741}
 

### 2019-10-03 NAR_DRAFT.tex
add citation for calibration

@@
 In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs \cite{Boutros2015a}.
-%Well-calibrated posterior probabilities facilitate the use of probabilistic risk models in the choice of treatment \cite{XXX}, rather than an all or nothing approach.
+Well-calibrated posterior probabilities facilitate the use of probabilistic risk models in the choice of treatment \cite{Holmberg2013}, rather than an all or nothing approach.
 For research purposes, the International Cancer Genome Consortium recommends that catalogs of somatic mutations target a precision of 95\% and a recall of 80\% \cite{Icgc}.
@@
 
-The \texttt{batcaver} R package can be downloaded or installed from \texttt{github.com/bmannakee/batcaver}
+The \texttt{batcaver} R package can be downloaded or installed from \texttt{http://github.com/bmannakee/batcaver}
 The version of \texttt{batcaver} used to generate results, and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}

### 2019-10-04 NAR_DRAFT.tex
commmit to merge

--- 2019-10-03 NAR_DRAFT.tex
@@
 \section{Software Availability}
+\bkmcomment{working on this today}
 
-The \texttt{batcaver} R package can be downloaded or installed from \texttt{github.com/bmannakee/batcaver}
-The version of \texttt{batcaver} used to generate results, and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}
+github for source code
+
+zenodo for figures, simulations
+
+zenodo for package at time of bioarxiv
 
@@
 We thank Prof.\ Edward J. Bedrick for fruitful discussions about the statistical model.
+\bkmcomment{Acknowledge HPC.}
 
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{figures/fig_wgs.pdf}
+  \includegraphics[width=\textwidth]{figures/fig_wgs.png}
   \end{center}
@@
   \caption{Model performance on 100X simulated whole genome. 
-  Precision vs. Recall for five different probability cutoffs. 
+  Positive predictive values vs. Recall for five different probability cutoffs. 
   Note the restricted axes to improve readability of points.}

### 2019-10-08 NAR_DRAFT.tex
hpc acknowledgement
new supplementary figure
Update title & abstract

--- 2019-10-03 NAR_DRAFT.tex
@@
 
-\title{\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation}
+\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}
 
@@
 \begin{abstract}
-%The accumulation of somatic mutations is an inevitable consequence of tumor development.
-Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
-Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
-The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
-% This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
-Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
-\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
-We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
-The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and identification of lower frequency variants while providing well calibrated false positive rate expectations.
-We provide an R package implementing the \batcave algorithm for use with MuTect output.
+Identifying somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
+Mutations at low allelic frequency, which are carried by only a portion of cells within a tumor, are particularly difficult to detect.
+Many algorithms have been developed to identify such mutations, but none models a key aspect of tumor biology.
+Namely, every tumor has its own profile of mutation types that it tends to generate.
+We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and then uses it as a prior for evaluating potential mutations.
+We also present an R implementation of the algorithm that is built on the popular somatic variant caller MuTect.
+Using simulations, we show that adding the \batcave algorithm to MuTect improves variant identification.
+It also improves the calibration of posterior probabilities, enabling a more principled tradeoff between precision and recall.
+Lastly, we show that \batcave also performs well on real data.
+Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
+More broadly, the algorithm can be added to other variant callers, and it can be extend to include additional biological features that affect mutation generation.
 \end{abstract}

### 2019-10-08 NAR_DRAFT.tex
Figure caption tweaks

@@
 \begin{figure}
-  \begin{center}
+\centering
   \includegraphics{figures/real_signatures_only.pdf}
-  \end{center}
-  \caption{Tumor mutation profiles.
-  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
+  \caption{Real tumor mutation profiles.
+  In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
   \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
@@
 Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
-\begin{equation}
+\begin{gather}
   \label{eqn:hypothesis}
-  \begin{array}{l}
     \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
     \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f}.
-  \end{array}
-\end{equation}
+\end{gather}
 Here $m$ is any of the 3 possible alternate non-reference bases and $\nu$ is the variant allele frequency.
@@
 \begin{figure}
-  \begin{center}
+\centering
   \includegraphics{figures/sim_signatures_only.pdf}
-  \end{center}
-  \caption{Tumor mutation profiles.
-  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
+  \caption{Simulated tumor mutation profiles.
+As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
   \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
@@
 \begin{figure}
-  \begin{center}
+\centering
   \includegraphics{figures/kl_only.pdf}
-  \end{center}
-  \caption{Convergence of the site-specific prior to the data generating distribution.
+  \caption{Convergence of the mutational prior to the data generating distribution.
 Plotted is the Kullback-Leibler divergence between the simulated and estimated profiles versus number of incorporated mutations for whole exomes.
@@
   \begin{figure*}
-    \begin{center}
+\centering
     \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
-    \end{center}
-    \caption{Variant-calling performance on both simulated and real data.
+    \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and \batcave results with black lines.
     \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation profiles.
-    \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated.
-For a perfectly calibrated caller, that curve would match the dashed y=x line.
-    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.
+    \textbf{(C)} and \textbf{(D)} Calibration plots.
+Shaded regions show distributions of posterior probabilities for true positive variants, and smooth lines show loess-smoothed relationships, from which the Integrated Calibration Index is calculated.
+For a perfectly calibrated caller, those curves would match the dashed y=x line.
+    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial mutation validation was performed \cite{Griffith2015,Shi2018}.
 }
@@
 \begin{figure}
-  \begin{center}
+\centering
   \includegraphics{figures/ppv_wes.pdf}
-  \end{center}
-  \caption{Performance of various posterior probability thresholds in 500X exomes. 
-  For realistic posterior probability thresholds,  \batcave has a positive predictive value that is better calibrated than MuTect, no matter the concentration of the mutation profile.
+  \caption{Posterior probability calibration for realistic calling thresholds, for 500X exomes.
+  Plotted is precision and recall for variants identified using various realistic posterior probability thresholds.
+At these thresholds, the precision of \batcave is much closer to the given threshold than MuTect, no matter the concentration of the mutation profile.
   }
@@
 We thank Prof.\ Edward J. Bedrick for fruitful discussions about the statistical model.
+This material is based upon High Performance Computing (HPC) resources supported by the University of Arizona TRIF, UITS, and RDI and maintained by the UA Research Technologies department.
 
@@
 \begin{figure}[b]
-  \begin{center}
-  \includegraphics[width=\textwidth]{figures/fig_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 100X simulated whole genome.
-  \textbf{(A)} Precision recall curves and \textbf{(B)} Reciever operating characteristic curves for 3 mutation profiles.
-  Distribution of estimated true positive probabilities for true positive  variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.
-  A perfectly calibrated model would generate the diagonal line.}
+\centering
+  \includegraphics[width=\textwidth]{figures/fig_wgs.png}
+  \caption{Variant-calling performance on  simulated 100X whole genomes. As in Fig.~\ref{NAR-wes_fig}A-D, but for 100X whole genomes.}
 \label{NAR-wgs_fig}
@@
 \begin{figure}
-  \begin{center}
+\centering
   \includegraphics{figures/ppv_wgs.pdf}
-  \end{center}
-  \caption{Model performance on 100X simulated whole genome. 
-  Precision vs. Recall for five different probability cutoffs. 
-  Note the restricted axes to improve readability of points.}
+  \caption{Posterior probability calibration for realistic calling thresholds, for 100X whole genomes. As in Fig.~\ref{NAR-ppv_fig}, but for 100X whole genomes.}
 \label{NAR-ppv_wgs_fig}

### 2019-10-08 NAR_DRAFT.tex
Minor text updates.

@@
 Namely, every tumor has its own profile of mutation types that it tends to generate.
-We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and then uses it as a prior for evaluating potential mutations.
+We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and mutation rate then uses it as a prior for evaluating potential mutations.
 We also present an R implementation of the algorithm that is built on the popular somatic variant caller MuTect.
@@
 Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
-More broadly, the algorithm can be added to other variant callers, and it can be extend to include additional biological features that affect mutation generation.
+More broadly, the algorithm can be added to other variant callers, and it can be extended to include additional biological features that affect mutation generation.
 \end{abstract}
@@
 Understanding the evolutionary history of a tumor, including the mutations that drive its growth, the genetic diversity within it, and the accumulation of new mutations, requires accurate variant identification, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
-Accurate variant calling is also critical for optimizing the treatment of individual patients' disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
+Accurate variant calling is also critical for optimizing the treatment of individual patients' disease \cite{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
 Low frequency mutations challenge current variant calling methods, because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this challenge increases with sequencing depth.
@@
 The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
-Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
-joint genotype analysis (SomaticSniper \citep{Larson2012}, JointSNVMix2 \citep{Roth2012a}, Seurat \citep{Christoforides2013}, CaVEMan \citep{Jones2016}, and MuClone \citep{Dorri2019}), allele frequency-based analysis (Strelka \citep{Saunders2012}, LoFreq \citep{Wilm2012}, EBCall \citep{Shiraishi2013b}, deepSNV \citep{Gerstung2012}, LoLoPicker \citep{Carrot-Zhang2017}, and MuSE \citep{Fan2016}), and ensemble and deep learning methods (MutationSeq \citep{Ding2012}, BAYSIC \citep{Cantarel2014}, SomaticSeq \citep{Fang2015}, and SNooPer \citep{Spinella2016}).
+Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
+joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka \cite{Saunders2012}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
 These methods vary in their complexity and specific focus.
@@
 %Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
-If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
-Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
+If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
+Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
 The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, the mutation profiles differs substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
+For example, the mutation profiles differ substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
 %Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
@@
 Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
-Our algorithm first estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
+Our \batcave algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
 Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
@@
 
-At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is local read depth.
+At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls $b$ and their associated quality scores $q$, where $\mathrm{d}$ is local read depth.
 Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
@@
 \end{equation}
-where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads.
+where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads, and $p(m,\nu)$ is the prior.
 
@@
 Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
-Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
+Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
 The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
@@
 \end{equation}
-The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
+The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
 %Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
 
-We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
+We model the observed mutation profile $\mathrm{S}$ as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
 Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
@@
   \label{eqn:to_allele}
-   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
+   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,C=c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
 \end{equation}
@@
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
-%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
+%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\cite{Williams2016}.
 %Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
@@
 \subsection{Implementation}
-We have implemented the \batcave algorithm as an R package (batcaver).
-The package leverages the Bioconductor packages \textrm{BSgenome} \citep{Pages2019}, \textrm{GenomicAlignments} \citep{Lawrence2013}, \textrm{VariantAnnotation} \citep{Obenchain2014}, and \textrm{SomaticSignatures} \citep{Gehring2015} fast and memory-efficient variant annotation and genomic context identification.
+We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
+The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
 Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
@@
 \subsection{Tumor simulations}
-We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of alleles.
+We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
 Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
-Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
+Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
 The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
@@
 Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
-Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
+Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
 \texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
@@
         --out \$out\_path/\$chr.txt 
-        --coverage\_file \$out\_path/\$chr.cov}.
+        --coverage\_file \$out\_path/\$chr.cov}.\\
 Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
 Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.
+
+\subsection{Calibration metric}
+To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
+Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
+For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
+The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.
 
@@
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-We ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
+For these multi-region data, we ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
 We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
-
-\subsection{Calibration metric}
-To quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
-Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
-For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
-The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.
 
@@
 \batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
-The mutation profile is estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
+The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
 The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.
@@
 The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}).
-We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
+We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
 In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
@@
 %In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-%For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\citep{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
+%For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\cite{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
 
@@
 
-For all simulated tumors, the estimated mutation rate is approximately $3\cdot10^{-7}$ (Table~\ref{table:01}), which is lower than the simulated rate of $3\cdot10^{-6}$.
+For all simulated tumors, the estimated mutation rate was approximately $3\cdot10^{-7}$ (Table~\ref{table:01}), which is lower than the simulated rate of $3\cdot10^{-6}$.
 This is likely due to restrictions within BAMSurgeon, such as sequencing depth and quality, that prevent 100\% of simulated variants from being inserted into the reads.
@@
 For these potential variants, \batcave tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D) while decreasing probabilities of low probability but false variants.
-For 500X exomes, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities (Figure \ref{NAR-wes_fig} and Table \ref{table:01}).
-For 100X whole genomes, the calibration of \batcave is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}), likely because there are few low probability true positive variants in tumors sequenced to 100X depth.
+For 500X exomes, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
+For 100X whole genomes, the calibration of \batcave is slightly worse (Fig.~\ref{NAR-wgs_fig} and Table~\ref{table:01}), likely because there are few low probability true positive variants in tumors sequenced to 100X depth.
 As with the other metrics, the advantage of \batcave increases with the concentration of the mutation profile and the sequencing depth.
@@
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \citep{Alexandrov2019} and with the low number of validated somatic mutations.
-For the validated variants, MuTect and \batcave yield almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
+We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
+For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
 
@@
 
-%Deep sequencing is necessary to measure intra-tumor heterogeneity \citep{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.
+%Deep sequencing is necessary to measure intra-tumor heterogeneity \cite{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.
 %Current approaches to very high depth next-generation sequencing \cite{Griffith2015}, however, require combining calls from multiple variant calling algorithms and then validating these calls via ultra-deep sequencing.
@@
 A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
-This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
+This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall for variants that pass heuristic filters (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
 Moreover, few true negative sites are validated, so specificity and calibration are impossible to calculate.
@@
 For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a constant somatic mutation probability at each genomic site \cite{Kim2018}.
-Replacing that constant probability with a mutational prior would require a more complicated manipulation of the quality scores output by Strelka than for MuTect but is conceptually straightforward.
+Replacing that constant probability with a mutational prior would require a more complicated manipulation of the quality scores output by Strelka than for MuTect, but it is conceptually straightforward.
 
 The \batcave algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
-The majority of the computational costs is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
+The majority of the computational cost is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
 Since most callers are already walking the reference genome during the calling process, extracting the trinucleotide context simultaneously would virtually eliminate the computational cost of implementing a mutational prior.
@@
 Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
-There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}.
+There are, however, many other aspects of genomic context that can affect local mutation rates \cite{Buisson2019}, including replication timing \cite{Stamatoyannopoulos2009}, expression level \cite{Pleasance2010}, and chromatin organization \cite{Schuster-Bockler2012}.
 Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutational prior using the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
@@
 \batcave divides the data into two classes: high- and low-confidence variants.
-The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
-Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \citep{Morris1983,Efron2014}. 
+The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
+Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
 In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
-A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
-Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
-A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
+An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
+Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, although those differences are relatively small.
+A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
 This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
 
-Our results show that adding a mutational context prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
+Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
 Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
@@
 The \texttt{batcaver} R package can be downloaded or installed from \texttt{http://github.com/bmannakee/batcaver}
-The version of \texttt{batcaver} used to generate results, and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}
+The version of \texttt{batcaver} used to generate results and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}
 Python code used to generate simulated tumors has been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471741}
 
-\section{Data Availability}
-COSMIC mutation signatures version 2 is available at \footnotesize{https://cancer.sanger.ac.uk/cosmic/signatures\_v2}.
+%\section{Data Availability}
+%COSMIC mutation signatures version 2 is available at \footnotesize{https://cancer.sanger.ac.uk/cosmic/signatures\_v2}.
 
@@
 \beginsupplement
-\section{Supplememtary Figures}
+\section{Supplementary Figures}
 
-\begin{figure}[b]
+\begin{figure} [H]
 \centering
@@
 
-\clearpage
-\begin{figure}
+\begin{figure} [H]
 \centering

### 2019-10-08 NAR_DRAFT.tex
Shorten abstract for journal.

@@
 \begin{abstract}
-Identifying somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
-Mutations at low allelic frequency, which are carried by only a portion of cells within a tumor, are particularly difficult to detect.
-Many algorithms have been developed to identify such mutations, but none models a key aspect of tumor biology.
+Detecting somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
+Mutations at low allelic frequency, those present in only a small portion of tumor cells, are particularly difficult to detect.
+Many algorithms have been developed to detect such mutations, but none models a key aspect of tumor biology.
 Namely, every tumor has its own profile of mutation types that it tends to generate.
-We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and mutation rate then uses it as a prior for evaluating potential mutations.
-We also present an R implementation of the algorithm that is built on the popular somatic variant caller MuTect.
-Using simulations, we show that adding the \batcave algorithm to MuTect improves variant identification.
-It also improves the calibration of posterior probabilities, enabling a more principled tradeoff between precision and recall.
-Lastly, we show that \batcave also performs well on real data.
+We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the individual tumor mutational profile and mutation rate then uses them in a prior for evaluating potential mutations.
+We also present an R implementation of the algorithm, built on the popular caller MuTect.
+Using simulations, we show that adding the \batcave algorithm to MuTect improves variant detection.
+It also improves the calibration of posterior probabilities, enabling more principled tradeoff between precision and recall.
+We also show that \batcave performs well on real data.
 Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.

### 2019-10-08 NAR_DRAFT.tex
commit to merge

--- 2019-10-08 NAR_DRAFT.tex
@@
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
-joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka \cite{Saunders2012}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
+joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka2 \cite{Kim2018}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
 These methods vary in their complexity and specific focus.

### 2019-10-08 NAR_DRAFT.tex
commit to merge

@@
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
-joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka2 \cite{Kim2018}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
+joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka2 \cite{Kim2018}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
 These methods vary in their complexity and specific focus.

### 2019-10-08 bioRxiv_v1.tex
Version for bioRxiv

--- 2019-10-08 NAR_DRAFT.tex
@@
-\documentclass[a4,center,fleqn]{NAR}
+\documentclass[a4,center,fleqn]{article}
 \bibliographystyle{NAR-natbib}
@@
 % Enter dates of publication
-\copyrightyear{2008}
-\pubdate{31 July 2009}
-\pubyear{2009}
-\jvolume{37}
-\jissue{12}
+%\copyrightyear{2008}
+%\pubdate{31 July 2009}
+%\pubyear{2009}
+%\jvolume{37}
+%\jissue{12}
 
-\articlesubtype{Genomics and Bioinformatics}
+%\articlesubtype{Genomics and Bioinformatics}
+
+\usepackage{fullpage}
+\usepackage{amsmath}
 
@@
 
-\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}
+\title{\batcave: Calling somatic mutations with a\\tumor- and site-specific prior}
 
 \author{%
-Brian K. Mannakee\,$^{1}$ and
-Ryan N. Gutenkunst\,$^{2}$%
+Brian K. Mannakee$^{1}$ and
+Ryan N. Gutenkunst$^{2}$%
 \footnote{To whom correspondence should be addressed.
@@
 
-\address{%
-$^{1}$Mel and Enid Zuckerman College of Public Health, University of Arizona, Tucson AZ
-and
-$^{2}$Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ}
+%\address{%
+%$^{1}$Mel and Enid Zuckerman College of Public Health, University of Arizona, Tucson AZ
+%and
+%$^{2}$Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ}
 % Affiliation must include:
@@
 
-\history{%
-Received January 1, 2009;
-Revised February 1, 2009;
-Accepted March 1, 2009}
+%\history{%
+%Received January 1, 2009;
+%Revised February 1, 2009;
+%Accepted March 1, 2009}
 
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
-joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka2 \cite{Kim2018}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
+joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka \cite{Saunders2012}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
 These methods vary in their complexity and specific focus.
@@
 For example, the mutation profiles differ substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
-%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
+%Figure \ref{NAR-sigfig} {a-e} shows several examples of mutation profiles from both real and simulated tumors.
 
@@
   In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
-  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
-  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
-  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
+  (A) The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
+  (B) The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
+  (C)\&(D) The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
    }
@@
 
-\section{MATERIALS AND METHODS}
-\subsection{Somatic variant calling probability model}
+\section*{MATERIALS AND METHODS}
+\subsection*{Somatic variant calling probability model}
 
@@
 
-\subsection{Site-specific prior probability of mutation}
+\subsection*{Site-specific prior probability of mutation}
 The probability that we have denoted $\mathrm{p}(m)$ in Eq.~\ref{eqn:2} is more precisely the joint probability that a mutation has occurred $M$ and that it was to allele $m$, which we denote $\mathrm{p}(m,M)$.
@@
 
-\subsection{Estimation of the mutation profile}
+\subsection*{Estimation of the mutation profile}
 Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
@@
 
-\subsection{Estimation of the mutation rate}
+\subsection*{Estimation of the mutation rate}
 The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
@@
 
-\subsection{Likelihood function}
+\subsection*{Likelihood function}
 The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq.
@@
 
-\subsection{Implementation}
+\subsection*{Implementation}
 We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
@@
 As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
-  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
+  (A) A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
+  (B) Equal proportions of signatures 1, 4, \& 5.
+  (C) Equal proportions of signatures 1, 3, \& 5.
    }
@@
 
-\subsection{Tumor simulations}
+\subsection*{Tumor simulations}
 We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
@@
         --reference\_sequence \$ref\_path 
-        --dbsnp \$db\_snp 
+        --dbsnp \$db\_snp \\
         --enable\_extended\_output 
@@
 
-\subsection{Calibration metric}
+\subsection*{Calibration metric}
 To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
@@
 
-\subsection{Real data}
+\subsection*{Real data}
 We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
@@
 
-\section{RESULTS}
+\section*{RESULTS}
 We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
@@
 
-\subsection{Tests using simulated data}
+\subsection*{Tests using simulated data}
 
@@
     \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and \batcave results with black lines.
-    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation profiles.
-    \textbf{(C)} and \textbf{(D)} Calibration plots.
+    (A) Precision-recall curves and {(B)} receiver operating characteristic curves for different mutation profiles.
+    (C) and {(D)} Calibration plots.
 Shaded regions show distributions of posterior probabilities for true positive variants, and smooth lines show loess-smoothed relationships, from which the Integrated Calibration Index is calculated.
 For a perfectly calibrated caller, those curves would match the dashed y=x line.
-    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial mutation validation was performed \cite{Griffith2015,Shi2018}.
+    {(E)} and {(F)} Precision-recall curves for real data in which substantial mutation validation was performed \cite{Griffith2015,Shi2018}.
 }
@@
 \begin{table*}[b]
-  \tableparts{%
-  \caption{Variant calling metrics for all data sets.}
-  \label{table:01}%
-  }{%
+
+  \caption{Variant calling metrics for all data sets.\label{table:01}}
+\resizebox{\textwidth}{!}{
   \begin{tabular}{lllllllllll}
-  \toprule
   Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
@@
   \\
-  \colrule
+\hline
   100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
@@
   \\
-  \botrule
-  \end{tabular}%
-  }
-  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
+  \end{tabular}}
+  $\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.
+
  \end{table*}
@@
 
-\subsection{Tests using real tumor data}
+\subsection*{Tests using real tumor data}
 We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the express purpose of evaluating tumor variant calling pipelines, yielding high quality true and false positive data~\cite{Griffith2015,Shi2018}.
@@
 
-\section{DISCUSSION}
+\section*{DISCUSSION}
 
@@
 
-% \section{CONCLUSION}
+% \section*{CONCLUSION}
 % \bkmcomment{I looked at a bunch of NAR methods papers, and none include a conclusions section, so I moved it to the last paragraph of discussion.}
 
-\section{Software Availability}
+\section*{Software Availability}
 
@@
 
-%\section{Data Availability}
+%\section*{Data Availability}
 %COSMIC mutation signatures version 2 is available at \footnotesize{https://cancer.sanger.ac.uk/cosmic/signatures\_v2}.
 
-\section{ACKNOWLEDGEMENTS}
+\section*{ACKNOWLEDGEMENTS}
 
@@
 
-\subsubsection{Conflict of interest statement.} None declared.
+%\subsubsection*{Conflict of interest statement.} None declared.
 
@@
 \beginsupplement
-\section{Supplementary Figures}
+\section*{Supplementary Figures}
 
-\begin{figure} [H]
+\begin{figure} [h]
 \centering
@@
 
-\begin{figure} [H]
+\begin{figure} [h]
 \centering

### 2019-11-18 ResponseToReviews.tex
Initial response to NAR - Genomics and Bioinformatics

\documentclass[11pt]{article}

\usepackage{fullpage}
\usepackage{color}
\newcommand{\BKM}[1]{\textcolor{blue}{BKM: #1}}
\newcommand{\RNG}[1]{\textcolor{red}{RNG: #1}}

\usepackage[export]{adjustbox}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage}
\cfoot{\includegraphics[height=0.6in,valign=c]{LetterheadFooter}}
\renewcommand{\headrulewidth}{0pt}

\newenvironment{response}
{\begin{quote}\itshape}
{\end{quote}}

\setlength{\parindent}{0em}
\setlength{\parskip}{0.5em}

\usepackage{xr}
\externaldocument{NAR_DRAFT}

\begin{document}
\vspace*{-0.5in}\hspace*{-0.2in}\includegraphics{LetterheadHeader}\vspace*{\baselineskip}

\noindent Dear Professor Herrero,

We appreciate the opportunity to revise our manuscript for publication in \emph{NAR Genomics and Bioinformatics}, and we thank you and the reviewers for the quality of the reviews. Below we respond in italics to the reviewers' concerns.

\noindent Sincerely,\\
\includegraphics{signature}\\
Ryan Gutenkunst, PhD\\
Associate Professor and Associate Department Head\\
Department of Molecular and Cellular Biology\\
University of Arizona

\section*{Associate Editor}

Stressing on some of the points raised by the reviewers, I would suggest you include more details about the assumption and limitations of BATCAVE. In particular, the possibility of changes of mutational signatures over time can be quite striking in some cases, especially if the original tumour is driven by very specific mutagens like UV light or tobacco smoke. Ideally, the software should try to correct for these cases or at least warn the user about a possible shift in signatures/contexts.
\begin{response}
\RNG{Correcting for context shifts beyond the scope of the current manuscript. But warning about it is a very good idea. Simple check, do a goodness-of-fit test on called the low-confidence mutations. What is the p-value for their profile versus the high-confidence? (A $\chi^2$ test may not be appropriate here, since many entries may have small numbers. This preprint suggests potential improvements: https://arxiv.org/abs/1711.05524.)}
\end{response}

Related to this, there is probably a minimum number of high-confidence mutations that are required by BATCAVE to produce reliable results. As with the previous point, ideally the batcaver should address or warn the user if the assumptions are violated.
\begin{response}
\RNG{Good idea, should be an easy check. Easiest way would be to count high-confidence mutations. More sophisticated approach would be to evaluate the convergence of the prior, using something like K-L distance.}
\end{response}

Lastly, as correctly pointed out by the second reviewer, the sequencing coverage only makes sense in the context of the purity of the samples and this information should be provided to the reader.
\begin{response}
\RNG{Brian: Can you think about this in the context of what MuTect already does?}
\end{response}

\section*{Reviewer 1}

Mannakee and Gutenkunst present BATCAVE, an algorithm that first estimates the tumor’s mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants based on mutational signatures. The concept behind this work is interesting. However, the difference in accuracy between MuTect and BATCAVE is relatively small except for the case when one has particularly high depth sequencing (500x), limiting the utility of the method.

Still, the 500x whole exome case does indicate that BATCAVE is better at identifying low frequency variants and thus can better reveal intratumoral heterogeneity, an important area of inquiry in cancer. Interestingly, the AUROC values for both MuTect and BATCAVE have room for improvement for the 500x whole exome case (Table 1). I suggest the authors further investigate how to improve the AUROC values. For example, in equation 4, it is unclear why the authors assume $\nu$ be uniformly distributed. Although this is also an assumption of MuTect, the BATCAVE approach inherently distinguishes mutations by allele frequency – since log-likelihood is closely associated to AF.  Perhaps a model in which $\nu$ follows a decreasing distribution vs. AF (e.g. in concordance with a model of a growing tumor) would yield better AUROC values.  
\begin{response}
\RNG{We've thought about this... Didn't we decide that it didn't make sense, because we'd just be overweighting low-frequency variants? I can think more about it.}
\end{response}

In Rubanova et al ( https://www.biorxiv.org/content/10.1101/260471v4.full), it is argued that there are changes in mutational signatures over time in cancers. While Mannakee and Gutenkunst mention this work, they should further comment on whether such changes in mutational signatures over time are small enough to be explained by the BATCAVER concept of re-evaluating mutations based on p-value.
\begin{response}
\RNG{Going to be difficult, depending on biology to offer specific guidance about the size of the change. Rather, we now do a goodness-of-fit test and warn the user if it looks like it's violated.}
\end{response}

Minor
p.2: “Folding the central base to the pyrimidines”: The “folding” terminology is not standard and should be improved. 
\begin{response}
\RNG{Brian, figure out correct terminology.}
\end{response}

Equation 11: This equation is not correct if there is subclonal copy number evolution that changes the distribution of mutations above the allele frequency threshold. The authors should more clearly state this assumption.
\begin{response}
\RNG{Revise to include assumption, and think in general about copy number changes. How does a standard MuTect pipeline handle this?}
\end{response}

\section*{Reviewer 2}

Mannakee et al have developed a tool called BATCAVE for calling mutations from sequencing data. They claim to have used a novel approach of using tumor and site-specific priors for calling mutations. In my opinion authors have done a good job in terms of using a novel approach which harnesses patient/tumor specific information to call mutations. I agree with the authors that this approach was missing in other tools. I accept this study to be published given the authors address few points mentioned below:

1. Does BATCAVE accounts for tumor purity estimates while calling mutations?
\begin{response}
\RNG{Brian: How does MuTect handle this?}
\end{response}

2. In case of multi-region sequencing data, used for studying spatial intra-tumor heterogeneity, usually the existing tools estimate very high levels of heterogeneity. I suppose this is because they are being very stringent in terms of calling mutations especially the ones with low vafs. In that case the people end up using other tools like bamreadcount etc. to check status of these mutations in the bam files and re-estimate the extent of heterogeneity. It will be important to see how BATCAVE performs in such datasets.
\begin{response}
\RNG{Unclear to me whether we should dodge this. We could say explicitly accounting for multiregional samples is out-of-scope, and add it to the discussion. Although we did do something like that for our breast cancer analysis.}
\end{response}

3. Do the mutation signatures (example smoking signatures in lung cancer or UV signatures in melanoma) still appear in the more stringent BATCAVE set of mutations?
\begin{response}
\RNG{Not exactly sure what the reviewer is asking here, but should be easy to check. It will certainly still appear in simulations. Not sure about the real data analyses. But we can check when we add our test as to whether the distribution has changed.}
\end{response}

\end{document}

### 2019-11-19 ResponseToReviews.tex
my thoughts

@@
 This preprint suggests potential improvements: https://arxiv.org/abs/1711.05524.)}
+\BKM{This looks quite easy to implement.
+Testing is another question.
+One thing we don't have right now is simulations where the signature changes appreciably, but we do have the infrastructure to do it.}
 \end{response}
@@
 More sophisticated approach would be to evaluate the convergence of the prior, using something like K-L distance.}
+\BKM{This makes sense to me. We could pick some sensible threshold and note the allele frequency at which it is crossed.}
 \end{response}
@@
 \RNG{Brian: Can you think about this in the context of what MuTect already does?}
+\BKM{Mutect has a parameter for contamination/purity if known. I could add a parameter default=0, and write a message to stdout with the value used. There are various methods for estimating purity.}
 \end{response}
@@
 \RNG{We've thought about this... Didn't we decide that it didn't make sense, because we'd just be overweighting low-frequency variants? I can think more about it.}
+\BKM{We thought about it and I actually implemented a version.
+What happened is that for the distribution I think is indicated the correction was far too strong, like you said it overweights low frequency variants.
+I will also think about ways to do this.}
 \end{response}
@@
 \RNG{Going to be difficult, depending on biology to offer specific guidance about the size of the change. Rather, we now do a goodness-of-fit test and warn the user if it looks like it's violated.}
+\BKM{Not entirely sure I understand this.
+We could detect a change above.
+Whether it matters or not is a complex interplay between depth and selection.
+For late changes most mutations with that signature will be below the threshold of detection.
+For an early mutation all of the marginal (low frequency) variants will be from the signature.}
 \end{response}
@@
 \RNG{Brian, figure out correct terminology.}
+\BKM{Will do}
 \end{response}
@@
 \RNG{Revise to include assumption, and think in general about copy number changes. How does a standard MuTect pipeline handle this?}
+\BKM{Will do.
+MuTect doesn't make copy number calls, and doesn't adjust allele frequency I don't think.
+They basically ignore it.
+I think a change large enough to have a real effect here would be more of a change in ploidy.}
 \end{response}
@@
 \RNG{Brian: How does MuTect handle this?}
+\BKM{We do not. MuTect has a parameter for this, I will check exactly what it does.}
 \end{response}
@@
 Although we did do something like that for our breast cancer analysis.}
+\BKM{One of the reasons we made the algorithm is to try to provide a more sensitive means for identifying low frequency alleles so that these types of measures are more accurate.
+The multi-region sample we got simply isn't deep enough for us to figure out whether we would change the heterogeneity calls.
+There are not enough callable low frequency variants.
+Which is why our performance is identical.
+I am unaware of any deeper data sets, and think simulating them is beyond the scope.}
 \end{response}
@@
 But we can check when we add our test as to whether the distribution has changed.}
+\BKM{I am not entirely sure.
+Perhaps he is asking if we decomposed the final "prior" multinomial parameters to see if we get the same signatures back out?
+We don't because as we observed in the work I was doing with Setayesh that the tools like deconstructSigs are bad.
+We know that we converge to the simulated distribution and we know that the simulated distribution is a equal linear combination of the 3 signatures we describe.}
 \end{response}

### 2019-12-02 ResponseToReviews.tex
small edit

@@
 \RNG{Going to be difficult, depending on biology to offer specific guidance about the size of the change. Rather, we now do a goodness-of-fit test and warn the user if it looks like it's violated.}
-\BKM{Not entirely sure I understand this.
-We could detect a change above.
+\BKM{We could detect a change above.
 Whether it matters or not is a complex interplay between depth and selection.
@@
 \RNG{Brian, figure out correct terminology.}
-\BKM{Will do}
+\BKM{Will do.\\Language from Alexandrov is:\\Because there are six classes of base substitution—C>A, C>G, C>T, T>A, T>C, T>G (all substitutions are referred to by the pyrimidine of the mutated Watson–Crick base pair)—and as we incorporated information on the bases immediately 5′ and 3′ to each mutated base, there are 96 possible mutations in this classification}
 \end{response}

### 2019-12-09 NAR_DRAFT.tex
add text as a comment to address editor comment number 2.

@@
 An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
-Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, although those differences are relatively small.
+Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
+\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
+In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
 A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.

### 2019-12-09 NAR_DRAFT.tex
text as comment for editor comment number 3

@@
 The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
+\bkmcomment{We make two simplifying assumptions here.
+First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
+Second, we assume 100\% tumor purity.
+MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
+We encourage users to take advantage of the MuTect contamination correction where appropriate.
+In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
+}
 We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
@@
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
+
 %A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\cite{Williams2016}.
@@
 To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
-When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
+When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), \bkmcomment{and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.}.
 For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.

### 2019-12-09 NAR_DRAFT.tex
add language to remove folding

@@
 The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
-At the focal site, a mutation $m$ can be to any of three alternate alleles.
+\bkmcomment{There are six classes of base substitution -- C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all substitutions are referred to by the pyrimidine of the mutated Watson–Crick base pair).
+This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}
+%Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
+%At the focal site, a mutation $m$ can be to any of three alternate alleles.
 Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.

### 2019-12-09 NAR_DRAFT.tex
fix watsoncrick

@@
 The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-\bkmcomment{There are six classes of base substitution -- C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all substitutions are referred to by the pyrimidine of the mutated Watson–Crick base pair).
+\bkmcomment{There are six classes of base substitution -- C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all substitutions are referred to by the pyrimidine of the mutated Watson - Crick base pair).
 This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}

### 2019-12-09 NAR_DRAFT.tex
add thing about copy number calls in the discussion

@@
 }
-We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
+%We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
 We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
@@
 This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
-
+\bkmcomment{We use properties of the allele frequency distribution generated by mutation acquisition during exponential growth.
+Copy number alterations can distort the allele frequency distribution.
+While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}
 Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.

### 2019-12-09 NAR_DRAFT.tex
small changes to copy number caveat

@@
 This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
-\bkmcomment{We use properties of the allele frequency distribution generated by mutation acquisition during exponential growth.
-Copy number alterations can distort the allele frequency distribution.
+\bkmcomment{We use properties of the allele frequency distribution of high-confidence mutations to estimate the mutation rate in a tumor, and copy number alterations can distort the allele frequency distribution.
 While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}

### 2019-12-16 NAR_DRAFT.tex
Quick and dirty thoughts about uniform \nu
Commit folding language fix

--- 2019-12-09 NAR_DRAFT.tex
@@
 The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-\bkmcomment{There are six classes of base substitution -- C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all substitutions are referred to by the pyrimidine of the mutated Watson - Crick base pair).
+\rngcomment{There are six classes of base substitution: C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all referred to by the pyrimidine of the mutated Watson-Crick pair).
 This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}
@@
 
+\rngcomment{Assumption of uniform prior with $\nu$. Standard (cite MuTect). Also distinguish between calling individual mutations and inferring distribution of allele frequencies, like ANGSD.}
+
 The improved calibration of \batcave posterior probabilities compared to MuTect provides several advantages.

### 2019-12-16 ResponseToReviews.tex
Commit folding language fix
Response to uniform nu comment

--- 2019-11-19 ResponseToReviews.tex
@@
 \begin{response}
-\RNG{We've thought about this... Didn't we decide that it didn't make sense, because we'd just be overweighting low-frequency variants? I can think more about it.}
-\BKM{We thought about it and I actually implemented a version.
-What happened is that for the distribution I think is indicated the correction was far too strong, like you said it overweights low frequency variants.
-I will also think about ways to do this.}
+A uniform prior over $\nu$ is appropriate when the goal is to make the best possible call at each potential variant site.
+Conceptually, before examining the read data we have no information about the allele frequency at any individual site.
+Practically, a nonuniform prior would favor low-frequency and thus typically low-evidence sites, increasing the false positive rate.
+If our goal were to best estimate the distribution of variant allele frequencies in the tumor, then a nonuniform $\nu$ would indeed be important.
+
+We discuss this issue in a new paragraph in Discussion.
 \end{response}
@@
 \begin{response}
-\RNG{Brian, figure out correct terminology.}
-\BKM{Will do}
+Corrected
 \end{response}

### 2019-12-16 NAR_DRAFT.tex
Response to uniform nu comment

@@
 
-\rngcomment{Assumption of uniform prior with $\nu$. Standard (cite MuTect). Also distinguish between calling individual mutations and inferring distribution of allele frequencies, like ANGSD.}
-
 The improved calibration of \batcave posterior probabilities compared to MuTect provides several advantages.
@@
 Replacing that constant probability with a mutational prior would require a more complicated manipulation of the quality scores output by Strelka than for MuTect, but it is conceptually straightforward.
+
+\rngcomment{
+Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
+In general, low-frequency variants are expected to be more common than high-frequency variants.
+Nevertheless, the uniform frequency prior is appropriate when the goal is to make the best call possible at each potential variant site.
+By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
+Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate \cite{Nielsen2012}.
+}
 

### 2019-12-16 ResponseToReviews.tex
Response to prior convergence question.

@@
 \begin{response}
-\RNG{Good idea, should be an easy check.
-Easiest way would be to count high-confidence mutations.
-More sophisticated approach would be to evaluate the convergence of the prior, using something like K-L distance.}
-\BKM{This makes sense to me. We could pick some sensible threshold and note the allele frequency at which it is crossed.}
+Our simulation studies suggest that a few hundred high-confidence mutations are enough to reliably infer even a sharply peaked mutation profile (Fig. 3).
+We have modified \texttt{batcaver} so that it now emits a warning if there are fewer than 500 high-confidence mutations in a sample.
 \end{response}

### 2019-12-16 NAR_DRAFT.tex
Response to prior convergence question.

@@
 To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
-When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), \bkmcomment{and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.}.
+When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.
 For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.

### 2019-12-16 ResponseToReviews.tex
Add response regarding copy number changes

@@
 \begin{response}
-\RNG{Revise to include assumption, and think in general about copy number changes. How does a standard MuTect pipeline handle this?}
-\BKM{Will do.
-MuTect doesn't make copy number calls, and doesn't adjust allele frequency I don't think.
-They basically ignore it.
-I think a change large enough to have a real effect here would be more of a change in ploidy.}
+The reviewer is correct, although a large number of copy number changes would be necessary to substantially bias the estimated mutation rate.
+Potential future work could incorporate external copy number calls to improve the estimate of the point mutation rate.
+
+We discuss this issue in a new paragraph in Discussion.
 \end{response}

### 2019-12-16 NAR_DRAFT.tex
Add response regarding copy number changes

@@
 
+\rngcomment{
+To estimate the tumor mutation rate, \batcave uses a model of the distribution of variant allele frequencies  generated during exponential growth (Eq.~\ref{eqn:mut_rate}).
+Mutations that change copy number might distort the empirical allele frequency distribution, biasing the estimate.
+But substantially biasing the estimate would require copy number changes in a large number of genomic regions.
+Nevertheless, a potential extension of the \batcave algorithm would be to incorporate external copy number calls in order to adjust allele frequencies before estimating the mutation rate.
+}
+
 The \batcave algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.

### 2019-12-16 ResponseToReviews.tex
Address multi-region comments

@@
 \begin{response}
-\RNG{Unclear to me whether we should dodge this.
-We could say explicitly accounting for multiregional samples is out-of-scope, and add it to the discussion.
-Although we did do something like that for our breast cancer analysis.}
-\BKM{One of the reasons we made the algorithm is to try to provide a more sensitive means for identifying low frequency alleles so that these types of measures are more accurate.
-The multi-region sample we got simply isn't deep enough for us to figure out whether we would change the heterogeneity calls.
-There are not enough callable low frequency variants.
-Which is why our performance is identical.
-I am unaware of any deeper data sets, and think simulating them is beyond the scope.}
+We agree that multi-region data are increasingly important.
+In fact, the breast cancer data we test on come from a multi-region study. 
+Because we built BATCAVE on top of the single-sample caller MuTect, we called variants independently across the samples in that study, and BATCAVE performed well.
+Unfortunately, the sequencing in that study is not deep enough to rigorously evaluate how estimates of heterogeneity would change with and without BATCAVE.
+
+More broadly, the BATCAVE algorithm could be implemented in a multi-regional caller that takes explicit advantage of the relationship between samples.
+But such an implementation is beyond the scope of the current manuscript.
+
+We have added a new paragraph in Discussion regarding multi-region samples, to address these issues.
 \end{response}

### 2019-12-16 NAR_DRAFT.tex
Address multi-region comments

@@
 By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
-Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate \cite{Nielsen2012}.
+Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate~\cite{Nielsen2012}.
 }
@@
 Nevertheless, a potential extension of the \batcave algorithm would be to incorporate external copy number calls in order to adjust allele frequencies before estimating the mutation rate.
+}
+
+\rngcomment{
+Sequencing multiple regions of the same tumor is of growing interest for understanding tumor heterogeneity and evolution \cite{McGranahan2017}.
+In fact, the breast cancer data we analyze came from a multi-region study~\cite{Shi2018}.
+We called variants independently across samples, because MuTect only considers a single sample at a time.
+But calling variants jointly across samples could substantially increase sensitivity.
+In the future, the \batcave algorithm could be integrated into a multi-sample caller such as MultiSNV~\cite{Josephidou2015}.
+In that case, an optimal implementation would likely exploit the relationship between mutational priors in related samples.
 }
@@
 This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
-\bkmcomment{We use properties of the allele frequency distribution generated by mutation acquisition during exponential growth.
-Copy number alterations can distort the allele frequency distribution.
-While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}
+
 Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.

### 2019-12-16 ResponseToReviews.tex
Respond to decomposition question.

@@
 \begin{response}
-\RNG{Not exactly sure what the reviewer is asking here, but should be easy to check.
-It will certainly still appear in simulations.
-Not sure about the real data analyses.
-But we can check when we add our test as to whether the distribution has changed.}
-\BKM{I am not entirely sure.
-Perhaps he is asking if we decomposed the final "prior" multinomial parameters to see if we get the same signatures back out?
-We don't because as we observed in the work I was doing with Setayesh that the tools like deconstructSigs are bad.
-We know that we converge to the simulated distribution and we know that the simulated distribution is a equal linear combination of the 3 signatures we describe.}
+In our real and simulated test cases, the mutational profile changes little between high- and low-confidence sets of mutations.
+In fact, we have added a statistical test for profile change, in response to other review comments.
+
+Approaches such as deconstructSigs can decompose an individual sample's mutational profile into a set of mutation signatures corresponding to different processes.
+In practice, however, we have found these approaches to be relatively unstable.
+Therefore, we do not believe it will be illuminating to apply them in the present manuscript.
 \end{response}

### 2019-12-16 NAR_DRAFT.tex
push addition of tumor purity

--- 2019-12-16 NAR_DRAFT.tex
@@
 The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-\rngcomment{There are six classes of base substitution: C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all referred to by the pyrimidine of the mutated Watson-Crick pair).
+\bkmcomment{There are six classes of base substitution -- C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all substitutions are referred to by the pyrimidine of the mutated Watson - Crick base pair).
 This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}
@@
 The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
-\bkmcomment{We make two simplifying assumptions here.
-First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
-Second, we assume 100\% tumor purity.
-MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
-We encourage users to take advantage of the MuTect contamination correction where appropriate.
-In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
-}
-%We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
+We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
+\bkmcomment{Low tumor purity can result in a distribution of reported allele frequencies that differs from the true distribution.
+We suggest in cases where an estimate of tumor purity is known, such as when using the GATK 4 best practices pipeline, providing it to batcaver.}
 We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
 We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
-In the R implementation of \batcave, $f_{min}$ is a free parameter.
+In the R implementation of \batcave, $f_{min}$ and tumor_purity are free parameters.
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
@@
 To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
-When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.
+When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), \bkmcomment{and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.}.
 For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.
@@
 
-\rngcomment{
-Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
-In general, low-frequency variants are expected to be more common than high-frequency variants.
-Nevertheless, the uniform frequency prior is appropriate when the goal is to make the best call possible at each potential variant site.
-By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
-Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate \cite{Nielsen2012}.
-}
-
 The \batcave algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.

### 2019-12-19 ResponseToReviews.tex
Add draft responses to profile change comments.

@@
 \begin{response}
-\RNG{Correcting for context shifts beyond the scope of the current manuscript.
-But warning about it is a very good idea.
-Simple check, do a goodness-of-fit test on called the low-confidence mutations.
-What is the p-value for their profile versus the high-confidence?
-(A $\chi^2$ test may not be appropriate here, since many entries may have small numbers.
-This preprint suggests potential improvements: https://arxiv.org/abs/1711.05524.)}
-\BKM{This looks quite easy to implement.
-Testing is another question.
-One thing we don't have right now is simulations where the signature changes appreciably, but we do have the infrastructure to do it.}
+As we noted in the Discussion, BATCAVE does assume that the mutation profile is similar between mutations that are high-confidence (tend to be earlier) and low-confidence (tend to be later).
+Modeling potential changes in the profile is beyond the scope of the current manuscript, but warning the user if such a shift exists is a great idea.
+We have thus implemented a statistical test as to whether the mutation profile from the high-confidence mutations differs significantly from the low-confidence mutations.
+If this test detects a difference, BATCAVE warns the user and outputs both profiles, so the user can assess the degree of difference and follow-up on any potential biological implications.
+
+When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data.
+Note that even with this shift, BATCAVE still performs better than MuTect on the validated mutations (although both perform extremely well).
+\RNG{The biology of the shift is...}
 \end{response}
@@
 \begin{response}
-\RNG{Going to be difficult, depending on biology to offer specific guidance about the size of the change. Rather, we now do a goodness-of-fit test and warn the user if it looks like it's violated.}
-\BKM{We could detect a change above.
-Whether it matters or not is a complex interplay between depth and selection.
-For late changes most mutations with that signature will be below the threshold of detection.
-For an early mutation all of the marginal (low frequency) variants will be from the signature.}
+We cannot comment generally on the magnitude of mutation profile changes, because this is an developing area of research.
+But to protect users, we have implemented a new statistical test to detect changes in mutation profiles between high- and low-confidence mutations.
+See our response to the Associate Editor for more details.
 \end{response}

### 2019-12-19 NAR_DRAFT.tex
Add draft responses to profile change comments.

--- 2019-12-16 NAR_DRAFT.tex
@@
 \subsection{Implementation}
-We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
+We have implemented the \batcave algorithm as an R package called \texttt{batcaver}.
 The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
 Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
+
+\rngcomment{
+An assumption of the \batcave algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
+To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
+If the two profiles differ at a significance threshold of $\alpha=XXX$, \texttt{batcaver} issues a warning and outputs the profiles.
+%The user can then assess them and follow-up on any biologically interesting differences.
+}
 
@@
 We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
-For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
+\rngcomment{
+For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
+The low-confidence mutation profile contains additional spikes of mutations corresponding to XXX (Fig.~\ref{fig:AML_profiles}).
+}
+Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
 
 Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
+\rngcomment{Brian: Add purity estimate, if in paper.}
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
@@
 In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
-An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
+
+An  implicit assumption of \batcave is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
 Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
-\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
-In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
+\rngcomment{Our \texttt{batcaver} software warns the user if the high- and low-confidence mutational profiles are statistically different.
+It also outputs those profiles, so the user can assess whether the difference will substantially affect their results.} 
 A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
@@
 Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
-
-% \section{CONCLUSION}
-% \bkmcomment{I looked at a bunch of NAR methods papers, and none include a conclusions section, so I moved it to the last paragraph of discussion.}
 
@@
 
+\begin{figure} [H]
+\centering
+  %\includegraphics{XXX}
+  \caption{Mutation profiles of high-confidence and low-confidence mutations in the AML data.}
+\label{fig:AML_profiles}
+\end{figure}
+
 \end{document}

### 2019-12-19 ResponseToReviews.tex
Add tumor purity draft responses. Will need to rerun analyses before finalizing.

@@
 \begin{response}
-\RNG{Brian: Can you think about this in the context of what MuTect already does?}
-\BKM{Mutect has a parameter for contamination/purity if known. I could add a parameter default=0, and write a message to stdout with the value used. There are various methods for estimating purity.}
+The direct effect of contamination by normal cells on the BATCAVE algorithm is to bias the estimated allele frequencies that are inputs to Eq.~11 for estimating the overall mutation rate.
+This bias can be simply corrected given an estimate of tumor purity.
+We have thus added a \texttt{--tumor\_purity} option to \texttt{batcaver}.
+When this is enabled, the estimated allele frequencies will be appropriately adjusted before applying Eq.~11.
+See new text in Methods for details.
+
+We also reanalyzed our real data examples, using purity estimates from the original publications.
+\RNG{
+The effect of accounting for tumor purity was to...
+We have updated the text, Table~1, and Fig.~4E\&F to reflect these new results.
+}
+
 \end{response}
@@
 \begin{response}
-\RNG{Brian: How does MuTect handle this?}
-\BKM{We do not. MuTect has a parameter for this, I will check exactly what it does.}
+We neglected to account for the effect of tumor purity on mutation rate estimation from Eq.~11.
+We have corrected this omission.
+Please see our response to the Associate Editor's comment for details.
 \end{response}

### 2019-12-19 NAR_DRAFT.tex
Add tumor purity draft responses. Will need to rerun analyses before finalizing.

@@
 The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
-\bkmcomment{We make two simplifying assumptions here.
-First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
-Second, we assume 100\% tumor purity.
-MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
-We encourage users to take advantage of the MuTect contamination correction where appropriate.
-In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
-}
-%We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
+We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
 We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
@@
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
+\rngcomment{Note that sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
+To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p$.
+All estimated allele frequencies are then multiplied by a factor of $1/p$ before being used in Eq.~\ref{eqn:mut_rate}.
+}
 
@@
 We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
+\rngcomment{Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.}
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-For these multi-region data, we ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
+For these multi-region data, we \rngcomment{used the XXX samples} and ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
+\rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
 We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
@@
 Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
-\rngcomment{Brian: Add purity estimate, if in paper.}
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.

### 2019-12-20 NAR_DRAFT.tex
commit to merge
check in changed Shi mutation rate

@@
 We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
-In the R implementation of \batcave, $f_{min}$ is a free parameter.
+In the R implementation of \batcave, $f_{min}$ and tumor_purity are free parameters.
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
@@
 \rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
-We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
+We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCrh37, so we used that reference.
 
@@
   \\
-  real breast \cite{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
+  real breast \cite{Shi2018} & Actual & 7e-7 & -- & -- & .972 & .972 & -- & --
   \\

### 2019-12-20 NAR_DRAFT.tex
add table for shi et al. needs mutation rates when they finish

@@
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-For these multi-region data, we \rngcomment{used the XXX samples} and ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
+For these multi-region data, we analyzed the three biological replicates from each of six tumors, running \batcave separately on each of the eighteen, and combined results to generate precision-recall curves.
 \rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
@@
 
+\section{Supplementary tables}
+
+\begin{table*}[b]
+  \tableparts{%
+  \caption{Sample summary for Shi et al.}
+  \label{table:01}%
+  }{%
+  \begin{tabular}{lll}
+  \toprule
+  Sample & Purity estimate (\%) & Mutation rate estimate 
+  \\
+  \colrule
+  Case 1 biorep A & 53.8 & 3.6e-7 
+  \\
+  Case 1 biorep B & 46.3 & 3.2e-7 
+  \\
+  Case 1 biorep C & 26.0 & 3.2e-7 
+  \\
+  Case 2 biorep A & 40.0 & 3.6e-7 
+  \\
+  Case 2 biorep B & 32.3 & 3.6e-7 
+  \\
+  Case 2 biorep C & .4(*) & 3.6e-7 & 
+  \\
+  Case 3 biorep A & 56.4 & 3.6e-8 
+  \\
+  Case 3 biorep B & 72.8 & 7e-7 
+  \\
+  Case 3 biorep C & 64.9 & 7e-7 
+  \\
+  Case 4 biorep A & 60.4 & 7e-7 
+  \\
+  Case 4 biorep B & 61.7 & 7e-7 
+  \\
+  Case 4 biorep C & 57.6 & 7e-7 
+  \\
+  Case 5 biorep A & 78.7 & 7e-7 
+  \\
+  Case 5 biorep B & 76.7 & 7e-7 
+  \\
+  Case 5 biorep C & 80.8 & 7e-7 
+  \\
+  Case 6 biorep A & 47.8 & 7e-7 
+  \\
+  Case 6 biorep B & 49.4 & 7e-7 
+  \\
+  Case 6 biorep C & 52.9 & 7e-7 
+  \\
+  \botrule
+  \end{tabular}%
+  }
+  {Shi et al. did not provide an estimate of purity for Case 2 biorep C. We chose .4 because it is similar to the other biological replicates from this sample.}
+\end{table*}
+
 \end{document}

### 2019-12-20 NAR_DRAFT.tex
add rates to table

@@
 To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
-If the two profiles differ at a significance threshold of $\alpha=XXX$, \texttt{batcaver} issues a warning and outputs the profiles.
+If the two profiles differ at a significance threshold of $\alpha=0.05$, \texttt{batcaver} issues a warning and outputs the profiles.
 %The user can then assess them and follow-up on any biologically interesting differences.
@@
 For these multi-region data, we analyzed the three biological replicates from each of six tumors, running \batcave separately on each of the eighteen, and combined results to generate precision-recall curves.
-\rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
+\rngcomment{Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.}
 We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCrh37, so we used that reference.
@@
   \colrule
-  Case 1 biorep A & 53.8 & 3.6e-7 
+  Case 1 biorep A & 53.8 & 9.27e-7 
   \\
-  Case 1 biorep B & 46.3 & 3.2e-7 
+  Case 1 biorep B & 46.3 & 8.4e-7 
   \\
-  Case 1 biorep C & 26.0 & 3.2e-7 
+  Case 1 biorep C & 26.0 & 1.02e-6 
   \\
-  Case 2 biorep A & 40.0 & 3.6e-7 
+  Case 2 biorep A & 40.0 & 1.07e-6 
   \\
-  Case 2 biorep B & 32.3 & 3.6e-7 
+  Case 2 biorep B & 32.3 & 1.04e-7 
   \\
-  Case 2 biorep C & .4(*) & 3.6e-7 & 
+  Case 2 biorep C & 40(*) & 9.56e-7
   \\
-  Case 3 biorep A & 56.4 & 3.6e-8 
+  Case 3 biorep A & 56.4 & 8.7e-7
   \\
-  Case 3 biorep B & 72.8 & 7e-7 
+  Case 3 biorep B & 72.8 & 7.37e-7 
   \\
-  Case 3 biorep C & 64.9 & 7e-7 
+  Case 3 biorep C & 64.9 & 4.3e-7 
   \\
-  Case 4 biorep A & 60.4 & 7e-7 
+  Case 4 biorep A & 60.4 & 8.75e-7 
   \\
-  Case 4 biorep B & 61.7 & 7e-7 
+  Case 4 biorep B & 61.7 & 7.91e-7 
   \\
-  Case 4 biorep C & 57.6 & 7e-7 
+  Case 4 biorep C & 57.6 & 8.14e-7 
   \\
-  Case 5 biorep A & 78.7 & 7e-7 
+  Case 5 biorep A & 78.7 & 5.7e-7 
   \\
-  Case 5 biorep B & 76.7 & 7e-7 
+  Case 5 biorep B & 76.7 & 5.8e-7 
   \\
-  Case 5 biorep C & 80.8 & 7e-7 
+  Case 5 biorep C & 80.8 & 1.8e-7 
   \\
-  Case 6 biorep A & 47.8 & 7e-7 
+  Case 6 biorep A & 47.8 & 2.77e-7 
   \\
-  Case 6 biorep B & 49.4 & 7e-7 
+  Case 6 biorep B & 49.4 & 4.08e-7 
   \\
-  Case 6 biorep C & 52.9 & 7e-7 
+  Case 6 biorep C & 52.9 & 4.37e-7 
   \\

### 2019-12-20 NAR_DRAFT.tex
Text updates regarding purity.

@@
 We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
+\rngcomment{Sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
+To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p_{ur}$.
+All estimated allele frequencies are then multiplied by a factor of $1/p_{ur}$ before being used in Eq.~\ref{eqn:mut_rate}.}
 We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
-In the R implementation of \batcave, $f_{min}$ and tumor_purity are free parameters.
+In the R implementation of \batcave, $f_{min}$ and $p_{ur}$ are free parameters.
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
-\rngcomment{Note that sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
-To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p$.
-All estimated allele frequencies are then multiplied by a factor of $1/p$ before being used in Eq.~\ref{eqn:mut_rate}.
-}
+For our simulated data analyses, we set $p_{ur}=1$, and for our real data analyses, we set $p_{ur}$ to the estimated purity.
 
@@
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-For these multi-region data, we analyzed the three biological replicates from each of six tumors, running \batcave separately on each of the eighteen, and combined results to generate precision-recall curves.
+For these multi-region data, we analyzed the three biological replicates from each of six tumors (Table~\ref{table:01}), running \batcave separately on each of the eighteen samples, and we combined results to generate precision-recall curves.
 \rngcomment{Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.}
@@
   \tableparts{%
-  \caption{Sample summary for Shi et al.}
+  \caption{Sample summary for Shi et al. (2018)}
   \label{table:01}%

### 2019-12-20 NAR_DRAFT.tex
Update significant figures in purity estimate tables.

@@
   \\
-  real breast \cite{Shi2018} & Actual & 7e-7 & -- & -- & .972 & .972 & -- & --
+  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .972 & .972 & -- & --
   \\
@@
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
+\rngcomment{Need to update!}
 We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
@@
   \colrule
-  Case 1 biorep A & 53.8 & 9.27e-7 
+  Case 1 biorep A & 53.8 & 9.3e-7 
   \\
@@
   \\
-  Case 1 biorep C & 26.0 & 1.02e-6 
+  Case 1 biorep C & 26.0 & 1.0e-6 
   \\
-  Case 2 biorep A & 40.0 & 1.07e-6 
+  Case 2 biorep A & 40.0 & 1.1e-6 
   \\
-  Case 2 biorep B & 32.3 & 1.04e-7 
+  Case 2 biorep B & 32.3 & 1.0e-7 
   \\
-  Case 2 biorep C & 40(*) & 9.56e-7
+  Case 2 biorep C & 40(*) & 1.0e-6
   \\
@@
   \\
-  Case 3 biorep B & 72.8 & 7.37e-7 
+  Case 3 biorep B & 72.8 & 7.4e-7 
   \\
@@
   \\
-  Case 4 biorep A & 60.4 & 8.75e-7 
+  Case 4 biorep A & 60.4 & 8.8e-7 
   \\
-  Case 4 biorep B & 61.7 & 7.91e-7 
+  Case 4 biorep B & 61.7 & 7.9e-7 
   \\
-  Case 4 biorep C & 57.6 & 8.14e-7 
+  Case 4 biorep C & 57.6 & 8.1e-7 
   \\
@@
   \\
-  Case 6 biorep A & 47.8 & 2.77e-7 
+  Case 6 biorep A & 47.8 & 2.8e-7 
   \\
-  Case 6 biorep B & 49.4 & 4.08e-7 
+  Case 6 biorep B & 49.4 & 4.1e-7 
   \\
-  Case 6 biorep C & 52.9 & 4.37e-7 
+  Case 6 biorep C & 52.9 & 4.4e-7 
   \\
@@
   }
-  {Shi et al. did not provide an estimate of purity for Case 2 biorep C. We chose .4 because it is similar to the other biological replicates from this sample.}
+  {Shi et al. did not estimate purity for Case 2 biorep C. We chose 40\% purity, based on the other biological replicates from this case.}
 \end{table*}

### 2020-01-10 ResponseToReviews.tex
Changes to text for updated results. Also some to Response.

@@
 When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data.
-Note that even with this shift, BATCAVE still performs better than MuTect on the validated mutations (although both perform extremely well).
-\RNG{The biology of the shift is...}
+Nevertheless, BATCAVE still performed extremely well.
 \end{response}
@@
 We also reanalyzed our real data examples, using purity estimates from the original publications.
-\RNG{
-The effect of accounting for tumor purity was to...
+Accounting for tumor purity changed the quantitative results only slightly, and did not change the qualitative conclusions.
 We have updated the text, Table~1, and Fig.~4E\&F to reflect these new results.
-}
 

### 2020-01-10 NAR_DRAFT.tex
add comparison figure
minor change to the main table. both aml and cell got slightly worse
fix supplementary figure width
Changes to text for updated results. Also some to Response.

@@
   \\
-  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
+  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .988 & .996 & -- & --
   \\
-  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .972 & .972 & -- & --
+  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .968 & .972 & -- & --
   \\
@@
 Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
-We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
+We estimated a per-base mutation rate for this tumor of $3.6\times10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
 \rngcomment{
 For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
-The low-confidence mutation profile contains additional spikes of mutations corresponding to XXX (Fig.~\ref{fig:AML_profiles}).
+The low-confidence mutation profile contains additional spikes of mutations corresponding to five substitution types (Fig.~\ref{fig:AML_profiles}).
 }
@@
 \rngcomment{Need to update!}
-We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
+We estimated an average per-base mutation rate for these tumor regions of $6.6\times10^{-7}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
 For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
@@
 \centering
-  %\includegraphics{XXX}
-  \caption{Mutation profiles of high-confidence and low-confidence mutations in the AML data.}
+  \includegraphics[width=\textwidth]{figures/aml_sig_comp.pdf}
+  \caption{Mutation profiles of A) high-confidence and B) low-confidence mutations in the AML data.}
 \label{fig:AML_profiles}
 \end{figure}
+
+\clearpage
 

### 2020-01-13 NAR_DRAFT.tex
fix plot and table
small fix to aucs in text

@@
 The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-\rngcomment{There are six classes of base substitution: C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all referred to by the pyrimidine of the mutated Watson-Crick pair).
-This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}
+There are six classes of base substitution: C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all referred to by the pyrimidine of the mutated Watson-Crick pair).
+This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.
 %Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
@@
 We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
-\rngcomment{Sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
+Sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
 To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p_{ur}$.
-All estimated allele frequencies are then multiplied by a factor of $1/p_{ur}$ before being used in Eq.~\ref{eqn:mut_rate}.}
+All estimated allele frequencies are then multiplied by a factor of $1/p_{ur}$ before being used in Eq.~\ref{eqn:mut_rate}.
 We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
@@
 
-\rngcomment{
 An assumption of the \batcave algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
@@
 %The user can then assess them and follow-up on any biologically interesting differences.
-}
 
@@
 We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
-\rngcomment{Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.}
+Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
 For these multi-region data, we analyzed the three biological replicates from each of six tumors (Table~\ref{table:01}), running \batcave separately on each of the eighteen samples, and we combined results to generate precision-recall curves.
-\rngcomment{Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.}
-We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCrh37, so we used that reference.
+Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.
+We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCh37, so we used that reference.
 
@@
   \\
-  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .988 & .996 & -- & --
+  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .996 & .988 & -- & --
   \\
-  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .968 & .972 & -- & --
+  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .972 & .968 & -- & --
   \\
@@
 We estimated a per-base mutation rate for this tumor of $3.6\times10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
-\rngcomment{
 For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
 The low-confidence mutation profile contains additional spikes of mutations corresponding to five substitution types (Fig.~\ref{fig:AML_profiles}).
-}
-Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
+Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.988 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
 
@@
 %We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-\rngcomment{Need to update!}
 We estimated an average per-base mutation rate for these tumor regions of $6.6\times10^{-7}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
@@
 
-\rngcomment{
 Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
@@
 Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate~\cite{Nielsen2012}.
-}
 
-\rngcomment{
 To estimate the tumor mutation rate, \batcave uses a model of the distribution of variant allele frequencies  generated during exponential growth (Eq.~\ref{eqn:mut_rate}).
@@
 Nevertheless, a potential extension of the \batcave algorithm would be to incorporate external copy number calls in order to adjust allele frequencies before estimating the mutation rate.
-}
 
-\rngcomment{
 Sequencing multiple regions of the same tumor is of growing interest for understanding tumor heterogeneity and evolution \cite{McGranahan2017}.
@@
 In that case, an optimal implementation would likely exploit the relationship between mutational priors in related samples.
-}
 
@@
 Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
-\rngcomment{Our \texttt{batcaver} software warns the user if the high- and low-confidence mutational profiles are statistically different.
-It also outputs those profiles, so the user can assess whether the difference will substantially affect their results.} 
+Our \texttt{batcaver} software warns the user if the high- and low-confidence mutational profiles are statistically different.
+It also outputs those profiles, so the user can assess whether the difference will substantially affect their results.
 A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
@@
 
-\begin{table*}[b]
+\begin{table*}[h]
   \tableparts{%

### 2020-01-13 ResponseToReviews.tex
Tweaks for submission of revision to NAR Bioinformatics and Genomics

@@
 We appreciate the opportunity to revise our manuscript for publication in \emph{NAR Genomics and Bioinformatics}, and we thank you and the reviewers for the quality of the reviews.
-Below we respond in italics to the reviewers' concerns.
+Below we respond in italics to the concerns of yourself and the reviewers.
+To aid re-review, we have included a manuscript with changes highlighted in the resubmission.
 
@@
 
-When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data.
-Nevertheless, BATCAVE still performed extremely well.
+When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data (Fig.~S3).
+Nevertheless, BATCAVE still performed extremely well (Table~1). 
 \end{response}
@@
 \begin{response}
-Our simulation studies suggest that a few hundred high-confidence mutations are enough to reliably infer even a sharply peaked mutation profile (Fig. 3).
+Our simulation studies suggest that a few hundred high-confidence mutations are enough to reliably infer even a sharply peaked mutation profile (Fig.~3).
 We have modified \texttt{batcaver} so that it now emits a warning if there are fewer than 500 high-confidence mutations in a sample.
@@
 
-We also reanalyzed our real data examples, using purity estimates from the original publications.
-Accounting for tumor purity changed the quantitative results only slightly, and did not change the qualitative conclusions.
+We also reanalyzed our real data examples, using purity estimates from the original publications (text and Table~S1).
+Accounting for tumor purity changed the quantitative results slightly but did not change the qualitative conclusions.
 We have updated the text, Table~1, and Fig.~4E\&F to reflect these new results.
@@
 A uniform prior over $\nu$ is appropriate when the goal is to make the best possible call at each potential variant site.
-Conceptually, before examining the read data we have no information about the allele frequency at any individual site.
+Conceptually, before examining the read data we have no information about the allele frequency at any individual site, hence a uniform prior.
 Practically, a nonuniform prior would favor low-frequency and thus typically low-evidence sites, increasing the false positive rate.
@@
 
-In Rubanova et al ( https://www.biorxiv.org/content/10.1101/260471v4.full), it is argued that there are changes in mutational signatures over time in cancers.
+In Rubanova et al (https://www.biorxiv.org/content/10.1101/260471v4.full), it is argued that there are changes in mutational signatures over time in cancers.
 While Mannakee and Gutenkunst mention this work, they should further comment on whether such changes in mutational signatures over time are small enough to be explained by the BATCAVER concept of re-evaluating mutations based on p-value.
@@
 We cannot comment generally on the magnitude of mutation profile changes, because this is an developing area of research.
-But to protect users, we have implemented a new statistical test to detect changes in mutation profiles between high- and low-confidence mutations.
-See our response to the Associate Editor for more details.
+But to protect users, we have implemented and applied a new statistical test to detect changes in mutation profiles between high- and low-confidence mutations.
+See our response to the Associate Editor for details.
 \end{response}
@@
 
-We discuss this issue in a new paragraph in Discussion.
+We discuss now this issue in a new paragraph in Discussion.
 \end{response}
@@
 \begin{response}
-In our real and simulated test cases, the mutational profile changes little between high- and low-confidence sets of mutations.
-In fact, we have added a statistical test for profile change, in response to other review comments.
+In response to this and other review comments, we have added a statistical test as to whether the mutation profile changes between high- and low-confidence sets of mutations.
+And in the AML data, we do see such a change (Fig.~S3).
 
-Approaches such as deconstructSigs can decompose an individual sample's mutational profile into a set of mutation signatures corresponding to different processes.
+Approaches such as deconstructSigs can decompose an individual sample's mutational profile into a set of mutation signatures corresponding to different processes.\footnote{Rosenthal et al.
+(2016) \emph{Genome Biology} 17:31}
 In practice, however, we have found these approaches to be relatively unstable.

### 2020-01-22 Mannakee_zip/NAR_DRAFT.tex
Clean up for final files to NAR Genomics and Bioinformatics

--- 2020-01-13 NAR_DRAFT.tex
@@
 
-\usepackage{xspace}
-\newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
-\newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
-\newcommand{\batcave}{BATCAVE\xspace}
-\newcommand{\beginsupplement}{%
-        \clearpage
-        \onecolumn
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{S\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}%
-     }
 % Enter dates of publication
-\copyrightyear{2008}
-\pubdate{31 July 2009}
-\pubyear{2009}
-\jvolume{37}
-\jissue{12}
+\copyrightyear{x}
+\pubdate{x}
+\pubyear{x}
+\jvolume{x}
+\jissue{x}
 
@@
 
-\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}
+\title{BATCAVE: Calling somatic mutations with a tumor- and site-specific prior}
 
@@
 Namely, every tumor has its own profile of mutation types that it tends to generate.
-We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the individual tumor mutational profile and mutation rate then uses them in a prior for evaluating potential mutations.
+We present BATCAVE (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the individual tumor mutational profile and mutation rate then uses them in a prior for evaluating potential mutations.
 We also present an R implementation of the algorithm, built on the popular caller MuTect.
-Using simulations, we show that adding the \batcave algorithm to MuTect improves variant detection.
+Using simulations, we show that adding the BATCAVE algorithm to MuTect improves variant detection.
 It also improves the calibration of posterior probabilities, enabling more principled tradeoff between precision and recall.
-We also show that \batcave performs well on real data.
+We also show that BATCAVE performs well on real data.
 Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
@@
 Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
-Our \batcave algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
+Our BATCAVE algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
 Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
@@
 \centering
-  \includegraphics{figures/real_signatures_only.pdf}
+  \includegraphics{real_signatures_only.pdf}
   \caption{Real tumor mutation profiles.
@@
 \end{equation}
-The focus of \batcave is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.
+The focus of BATCAVE is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.
 
@@
 \end{equation}
-In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.
+In BATCAVE we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.
 
@@
 The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
-The R implementation of \batcave ships with pre-computed tables for both human whole exomes and whole genomes.
+The R implementation of BATCAVE ships with pre-computed tables for both human whole exomes and whole genomes.
 
@@
 We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
-In the R implementation of \batcave, $f_{min}$ and $p_{ur}$ are free parameters.
+In the R implementation of BATCAVE, $f_{min}$ and $p_{ur}$ are free parameters.
 For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
@@
 \subsection{Likelihood function}
-The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq.
+The current implementation of BATCAVE builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq.
 \ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
@@
 \subsection{Implementation}
-We have implemented the \batcave algorithm as an R package called \texttt{batcaver}.
+We have implemented the BATCAVE algorithm as an R package called \texttt{batcaver}.
 The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
@@
 
-An assumption of the \batcave algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
+An assumption of the BATCAVE algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
 To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
@@
 \centering
-  \includegraphics{figures/sim_signatures_only.pdf}
+  \includegraphics{sim_signatures_only.pdf}
   \caption{Simulated tumor mutation profiles.
@@
 Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
-Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.
+Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to BATCAVE for prior estimation and rescoring.
 
 \subsection{Calibration metric}
-To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
-Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
+To quantify the difference in calibration between MuTect and BATCAVE, we used the Integrated Calibration Index \cite{Austin2019}.
+Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and BATCAVE.
 For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
@@
 Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-For these multi-region data, we analyzed the three biological replicates from each of six tumors (Table~\ref{table:01}), running \batcave separately on each of the eighteen samples, and we combined results to generate precision-recall curves.
+For these multi-region data, we analyzed the three biological replicates from each of six tumors (Table~\ref{table:01}), running BATCAVE separately on each of the eighteen samples, and we combined results to generate precision-recall curves.
 Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.
@@
 \section{RESULTS}
-We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
+We implemented BATCAVE as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
+BATCAVE extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
 The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
-The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.
+The BATCAVE algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.
 
-To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
+To test the performance of BATCAVE, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
 The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}).
-We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
+We also tested BATCAVE using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
 In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
@@
 \centering
-  \includegraphics{figures/kl_only.pdf}
+  \includegraphics{kl_only.pdf}
   \caption{Convergence of the mutational prior to the data generating distribution.
@@
 \centering
-    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
-    \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and \batcave results with black lines.
+    \includegraphics{fig_wes.pdf}
+    \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and BATCAVE results with black lines.
     \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation profiles.
@@
 We assessed classification performance using the areas under both the receiver operating characteristic and the precision-recall curves, because the classes are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
-By both metrics \batcave outperforms MuTect (Fig.~\ref{NAR-wes_fig}A\&B, Fig.~\ref{NAR-wgs_fig}A\&B, and Table~\ref{table:01}).
+By both metrics BATCAVE outperforms MuTect (Fig.~\ref{NAR-wes_fig}A\&B, Fig.~S1A\&B, and Table~\ref{table:01}).
 The extent of the performance difference is dependent on both the sequencing depth and the concentration of the mutation profile.
-Deeper sequencing and more concentrated mutation profiles increase the performance advantage of \batcave.
+Deeper sequencing and more concentrated mutation profiles increase the performance advantage of BATCAVE.
 
@@
 We measured overall calibration performance using the Integrated Calibration Index (ICI) \cite{Austin2019}, which integrates the difference between predicted and observed probabilities, weighted by the density of the predicted probabilities.
-This metric is particularly useful in our case, because the density of posterior probabilities is bi-modal (Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D).
+This metric is particularly useful in our case, because the density of posterior probabilities is bi-modal (Fig.~\ref{NAR-wes_fig}C\&D and~S1C\&D).
 A large fraction of true negative variants have posterior probabilities less than $10^{-4}$, far below any meaningful threshold, so we evaluated calibration only on potential variants with posterior probability greater than 0.01.
-For these potential variants, \batcave tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D) while decreasing probabilities of low probability but false variants.
-For 500X exomes, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
-For 100X whole genomes, the calibration of \batcave is slightly worse (Fig.~\ref{NAR-wgs_fig} and Table~\ref{table:01}), likely because there are few low probability true positive variants in tumors sequenced to 100X depth.
-As with the other metrics, the advantage of \batcave increases with the concentration of the mutation profile and the sequencing depth.
+For these potential variants, BATCAVE tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D and~S1C\&D) while decreasing probabilities of low probability but false variants.
+For 500X exomes, the calibration of BATCAVE is better than MuTect across the full spectrum of posterior probabilities (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
+For 100X whole genomes, the calibration of BATCAVE is slightly worse (Fig.~S1 and Table~\ref{table:01}), likely because there are few low probability true positive variants in tumors sequenced to 100X depth.
+As with the other metrics, the advantage of BATCAVE increases with the concentration of the mutation profile and the sequencing depth.
 
@@
 \centering
-  \includegraphics{figures/ppv_wes.pdf}
+  \includegraphics{ppv_wes.pdf}
   \caption{Posterior probability calibration for realistic calling thresholds, for 500X exomes.
   Plotted is precision and recall for variants identified using various realistic posterior probability thresholds.
-At these thresholds, the precision of \batcave is much closer to the given threshold than MuTect, no matter the concentration of the mutation profile.
+At these thresholds, the precision of BATCAVE is much closer to the given threshold than MuTect, no matter the concentration of the mutation profile.
   }
@@
 The user's choice of threshold ideally meets their need to balance precision and recall; accurate posterior probability estimates enable an informed choice.
-For posterior probability thresholds between 60 and 90\%, the precision of \batcave calls is similar to the chosen threshold (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-For this range of thresholds, however, the posterior probabilities from MuTect poorly predict precision (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-For any posterior probability threshold above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the threshold increases.
-The cost of MuTect's compressed range of posterior probabilities is recall; at any posterior probability threshold \batcave has recall better than MuTect.
-Consequently, \batcave posterior probabilities are more informative than MuTect's with regard to choosing a calling threshold.
-%The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
-%Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
+For posterior probability thresholds between 60 and 90\%, the precision of BATCAVE calls is similar to the chosen threshold (Fig.~\ref{NAR-ppv_fig}\&S2).
+For this range of thresholds, however, the posterior probabilities from MuTect poorly predict precision (Fig.~\ref{NAR-ppv_fig}\&S2).
+For any posterior probability threshold above 70\%, MuTect has a false positive rate of roughly 8\%, whereas BATCAVE has a false positive rate that decreases as the threshold increases.
+The cost of MuTect's compressed range of posterior probabilities is recall; at any posterior probability threshold BATCAVE has recall better than MuTect.
+Consequently, BATCAVE posterior probabilities are more informative than MuTect's with regard to choosing a calling threshold.
+%The cost of MuTect's compression is recall, which is 3-5\% worse than BATCAVE at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
+%Figure~S2 shows results for 100X whole genomes, where the results are similar for precision.
 
 \subsection{Tests using real tumor data}
-We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the express purpose of evaluating tumor variant calling pipelines, yielding high quality true and false positive data~\cite{Griffith2015,Shi2018}.
+We tested BATCAVE using two data sets for which deep sequencing and variant validation were performed with the express purpose of evaluating tumor variant calling pipelines, yielding high quality true and false positive data~\cite{Griffith2015,Shi2018}.
 However, only variants called by at least one variant caller were validated.
@@
 For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
-The low-confidence mutation profile contains additional spikes of mutations corresponding to five substitution types (Fig.~\ref{fig:AML_profiles}).
-Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.988 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
+The low-confidence mutation profile contains additional spikes of mutations corresponding to five substitution types (Fig.~S3).
+Nevertheless, for both MuTect and BATCAVE, the precision-recall curve is almost perfect for the validated variants (.988 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
 
@@
 We estimated an average per-base mutation rate for these tumor regions of $6.6\times10^{-7}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
-For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
+For the validated variants, MuTect and BATCAVE yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
 
@@
 
-\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
+BATCAVE is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
 Our implementation is built on MuTect, one of the most widely used somatic variant callers.
-\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D,  \ref{NAR-wgs_fig}, and Table~\ref{table:01}) across the entire range of recall and specificity.
-Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (Fig.~\ref{NAR-ppv_fig} and~\ref{NAR-ppv_wgs_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
-For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F and Table~\ref{table:01}).
-The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.
+BATCAVE improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D,  S1, and Table~\ref{table:01}) across the entire range of recall and specificity.
+Moreover, BATCAVE is better calibrated than MuTect at relevant posterior probability thresholds (Fig.~\ref{NAR-ppv_fig} and~S2), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
+For real data, testing on validated calls shows that BATCAVE does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F and Table~\ref{table:01}).
+The BATCAVE algorithm can thus be included in a wide variety of sequencing pipelines.
 
-We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
+We evaluated BATCAVE with simulated tumors with three different mutation profiles and two real tumors.
 The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively.
@@
 The two real data sets we considered are among the few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
-They happen, however, to have diffuse mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
+They happen, however, to have diffuse mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of BATCAVE over MuTect (Table~\ref{table:01}).
 A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
-This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall for variants that pass heuristic filters (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
+This subset tends to be relatively easy to call, so both MuTect and BATCAVE have almost perfect precision and recall for variants that pass heuristic filters (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
 Moreover, few true negative sites are validated, so specificity and calibration are impossible to calculate.
@@
 
-The improved calibration of \batcave posterior probabilities compared to MuTect provides several advantages.
+The improved calibration of BATCAVE posterior probabilities compared to MuTect provides several advantages.
 In practice, called variants are often manually reviewed to further reduce false positives \cite{Barnell2019}.
@@
 
-Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
-We chose to build \batcave off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
+Our current implementation of BATCAVE is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
+We chose to build BATCAVE off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
 But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
@@
 
-Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
+Following MuTect \cite{Cibulskis2013}, BATCAVE assumes a uniform prior on the variant allele frequency.
 In general, low-frequency variants are expected to be more common than high-frequency variants.
@@
 
-To estimate the tumor mutation rate, \batcave uses a model of the distribution of variant allele frequencies  generated during exponential growth (Eq.~\ref{eqn:mut_rate}).
+To estimate the tumor mutation rate, BATCAVE uses a model of the distribution of variant allele frequencies  generated during exponential growth (Eq.~\ref{eqn:mut_rate}).
 Mutations that change copy number might distort the empirical allele frequency distribution, biasing the estimate.
 But substantially biasing the estimate would require copy number changes in a large number of genomic regions.
-Nevertheless, a potential extension of the \batcave algorithm would be to incorporate external copy number calls in order to adjust allele frequencies before estimating the mutation rate.
+Nevertheless, a potential extension of the BATCAVE algorithm would be to incorporate external copy number calls in order to adjust allele frequencies before estimating the mutation rate.
 
@@
 But calling variants jointly across samples could substantially increase sensitivity.
-In the future, the \batcave algorithm could be integrated into a multi-sample caller such as MultiSNV~\cite{Josephidou2015}.
+In the future, the BATCAVE algorithm could be integrated into a multi-sample caller such as MultiSNV~\cite{Josephidou2015}.
 In that case, an optimal implementation would likely exploit the relationship between mutational priors in related samples.
 
-The \batcave algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
+The BATCAVE algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
 The majority of the computational cost is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
@@
 
-The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
+The BATCAVE algorithm incorporates genomic context into the probabilistic model for variant calling.
 Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
 There are, however, many other aspects of genomic context that can affect local mutation rates \cite{Buisson2019}, including replication timing \cite{Stamatoyannopoulos2009}, expression level \cite{Pleasance2010}, and chromatin organization \cite{Schuster-Bockler2012}.
-Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutational prior using the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
+Some of these, such as replication timing and chromatin organization, could be incorporated into the BATCAVE mutational prior using the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
 Others, such as expression level, could be tumor-specific, but would require information not available in the variant calls to compute.
@@
 
-\batcave divides the data into two classes: high- and low-confidence variants.
+BATCAVE divides the data into two classes: high- and low-confidence variants.
 The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
@@
 
-An  implicit assumption of \batcave is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
+An  implicit assumption of BATCAVE is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
 Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
@@
 It also outputs those profiles, so the user can assess whether the difference will substantially affect their results.
-A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
+A potential extension of the BATCAVE algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
 This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
@@
 Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
-Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
-Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
+Moreover, BATCAVE proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
+Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with BATCAVE.
 Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
@@
 
-\beginsupplement
-\section{Supplementary Figures}
-
-\begin{figure} [H]
-\centering
-  \includegraphics[width=\textwidth]{figures/fig_wgs.png}
-  \caption{Variant-calling performance on  simulated 100X whole genomes. As in Fig.~\ref{NAR-wes_fig}A-D, but for 100X whole genomes.}
-\label{NAR-wgs_fig}
-\end{figure}
-
-\begin{figure} [H]
-\centering
-  \includegraphics{figures/ppv_wgs.pdf}
-  \caption{Posterior probability calibration for realistic calling thresholds, for 100X whole genomes. As in Fig.~\ref{NAR-ppv_fig}, but for 100X whole genomes.}
-\label{NAR-ppv_wgs_fig}
-\end{figure}
-
-\begin{figure} [H]
-\centering
-  \includegraphics[width=\textwidth]{figures/aml_sig_comp.pdf}
-  \caption{Mutation profiles of A) high-confidence and B) low-confidence mutations in the AML data.}
-\label{fig:AML_profiles}
-\end{figure}
-
-\clearpage
-
-\section{Supplementary tables}
-
-\begin{table*}[h]
-  \tableparts{%
-  \caption{Sample summary for Shi et al. (2018)}
-  \label{table:01}%
-  }{%
-  \begin{tabular}{lll}
-  \toprule
-  Sample & Purity estimate (\%) & Mutation rate estimate 
-  \\
-  \colrule
-  Case 1 biorep A & 53.8 & 9.3e-7 
-  \\
-  Case 1 biorep B & 46.3 & 8.4e-7 
-  \\
-  Case 1 biorep C & 26.0 & 1.0e-6 
-  \\
-  Case 2 biorep A & 40.0 & 1.1e-6 
-  \\
-  Case 2 biorep B & 32.3 & 1.0e-7 
-  \\
-  Case 2 biorep C & 40(*) & 1.0e-6
-  \\
-  Case 3 biorep A & 56.4 & 8.7e-7
-  \\
-  Case 3 biorep B & 72.8 & 7.4e-7 
-  \\
-  Case 3 biorep C & 64.9 & 4.3e-7 
-  \\
-  Case 4 biorep A & 60.4 & 8.8e-7 
-  \\
-  Case 4 biorep B & 61.7 & 7.9e-7 
-  \\
-  Case 4 biorep C & 57.6 & 8.1e-7 
-  \\
-  Case 5 biorep A & 78.7 & 5.7e-7 
-  \\
-  Case 5 biorep B & 76.7 & 5.8e-7 
-  \\
-  Case 5 biorep C & 80.8 & 1.8e-7 
-  \\
-  Case 6 biorep A & 47.8 & 2.8e-7 
-  \\
-  Case 6 biorep B & 49.4 & 4.1e-7 
-  \\
-  Case 6 biorep C & 52.9 & 4.4e-7 
-  \\
-  \botrule
-  \end{tabular}%
-  }
-  {Shi et al. did not estimate purity for Case 2 biorep C. We chose 40\% purity, based on the other biological replicates from this case.}
-\end{table*}
-
 \end{document}

### 2020-01-22 suppl.tex
Clean up for final files to NAR Genomics and Bioinformatics

--- 2020-01-13 NAR_DRAFT.tex
@@
-\documentclass[a4,center,fleqn]{NAR}
-\bibliographystyle{NAR-natbib}
-\usepackage{color}
-\usepackage{xfrac}
+\documentclass{article}
 
-\usepackage{xspace}
-\newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
-\newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
-\newcommand{\batcave}{BATCAVE\xspace}
-\newcommand{\beginsupplement}{%
-        \clearpage
-        \onecolumn
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{S\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}%
-     }
-% Enter dates of publication
-\copyrightyear{2008}
-\pubdate{31 July 2009}
-\pubyear{2009}
-\jvolume{37}
-\jissue{12}
-
-\articlesubtype{Genomics and Bioinformatics}
+\usepackage{fullpage}
+\usepackage{float}
+\usepackage{placeins}
+\usepackage{graphicx}
 
@@
 
-\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
 
-\author{%
-Brian K. Mannakee\,$^{1}$ and
-Ryan N. Gutenkunst\,$^{2}$%
-\footnote{To whom correspondence should be addressed.
-Email: rgutenk@email.arizona.edu}}
+\title{Supplementary Material\\~\\BATCAVE: Calling somatic mutations with a tumor- and site-specific prior}
 
-\address{%
-$^{1}$Mel and Enid Zuckerman College of Public Health, University of Arizona, Tucson AZ
-and
-$^{2}$Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ}
-% Affiliation must include:
-% Department name, institution name, full road and district address,
-% state, Zip or postal code, country
-
-\history{%
-Received January 1, 2009;
-Revised February 1, 2009;
-Accepted March 1, 2009}
+\author{
+Brian K. Mannakee and
+Ryan N. Gutenkunst}
+\date{}
 
 \maketitle
-
-\begin{abstract}
-Detecting somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
-Mutations at low allelic frequency, those present in only a small portion of tumor cells, are particularly difficult to detect.
-Many algorithms have been developed to detect such mutations, but none models a key aspect of tumor biology.
-Namely, every tumor has its own profile of mutation types that it tends to generate.
-We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the individual tumor mutational profile and mutation rate then uses them in a prior for evaluating potential mutations.
-We also present an R implementation of the algorithm, built on the popular caller MuTect.
-Using simulations, we show that adding the \batcave algorithm to MuTect improves variant detection.
-It also improves the calibration of posterior probabilities, enabling more principled tradeoff between precision and recall.
-We also show that \batcave performs well on real data.
-Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
-More broadly, the algorithm can be added to other variant callers, and it can be extended to include additional biological features that affect mutation generation.
-\end{abstract}
-
-\section{Introduction}
-
-Cancer develops through the accumulation of somatic mutations and clonal selection of cells with mutations that confer an advantage.
-Understanding the evolutionary history of a tumor, including the mutations that drive its growth, the genetic diversity within it, and the accumulation of new mutations, requires accurate variant identification, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
-Accurate variant calling is also critical for optimizing the treatment of individual patients' disease \cite{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
-Low frequency mutations challenge current variant calling methods, because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this challenge increases with sequencing depth.
-
-Many methods have been developed for calling somatic mutations from NGS data.
-The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
-Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
-joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka2 \cite{Kim2018}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
-These methods vary in their complexity and specific focus.
-But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.
-
-The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
-%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
-If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
-Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
-The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
-For example, the mutation profiles differ substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
-%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
-
-Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
-Our \batcave algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
-Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
-Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
-Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
-Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.
-
-\begin{figure}
-\centering
-  \includegraphics{figures/real_signatures_only.pdf}
-  \caption{Real tumor mutation profiles.
-  In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
-  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
-  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
-  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
-   }
-  \label{NAR-realsigfig}
- \end{figure}
-
-\section{MATERIALS AND METHODS}
-\subsection{Somatic variant calling probability model}
-
-At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls $b$ and their associated quality scores $q$, where $\mathrm{d}$ is local read depth.
-Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
-\begin{gather}
-  \label{eqn:hypothesis}
-    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
-    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f}.
-\end{gather}
-Here $m$ is any of the 3 possible alternate non-reference bases and $\nu$ is the variant allele frequency.
-The maximum likelihood estimate of $\nu$ is simply $\hat{f}$, the number of variant reads divided by the local read depth.
-The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given that hypothesis and the prior probability of that hypothesis. 
-Assuming that reads are independent, this is
-\begin{equation}\label{eqn:1}
-  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
-\end{equation}
-where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads, and $p(m,\nu)$ is the prior.
-
-Assuming that the identity of the alternate allele and its allele frequency are independent and that $\nu$ is uniformly distributed, Eq.~\ref{eqn:1} becomes
-\begin{equation}  \label{eqn:2}
-  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
-\end{equation}
-The focus of \batcave is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.
-
-\subsection{Site-specific prior probability of mutation}
-The probability that we have denoted $\mathrm{p}(m)$ in Eq.~\ref{eqn:2} is more precisely the joint probability that a mutation has occurred $M$ and that it was to allele $m$, which we denote $\mathrm{p}(m,M)$.
-But $\mathrm{p}(m,M)$ is not uniform across the genome.
-Rather it depends on the local genomic context $C$, so its full form is $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
-Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' theorem to further decompose as
-\begin{equation}
-  \label{eqn:full_prior}
-  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
-\end{equation}
-We next show how to estimate the quantities in Eq.~\ref{eqn:full_prior}.
-
-\subsection{Estimation of the mutation profile}
-Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
-Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
-The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
-There are six classes of base substitution: C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all referred to by the pyrimidine of the mutated Watson-Crick pair).
-This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.
-%Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
-%At the focal site, a mutation $m$ can be to any of three alternate alleles.
-Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
-Eq.~\ref{eqn:full_prior} is then
-\begin{equation}
-  \label{eqn:detailed_prior}
-  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
-\end{equation}
-The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
-%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
-
-We model the observed mutation profile $\mathrm{S}$ as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
-Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
-In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
-But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
-We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
-\begin{equation}
-\begin{aligned}
-  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
-  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
-\end{aligned}
-\end{equation}
-In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.
-
-To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
-These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
-Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 without the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
-The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
-\begin{equation}
-    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\},
-\end{equation}
-and $I$ is the indicator function.
-Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
-\begin{equation}
-  \label{eqn:post_pred}
-  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
-\end{equation}
-The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$ is then
-\begin{equation}
-  \label{eqn:to_allele}
-   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,C=c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
-\end{equation}
-
-The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
-The R implementation of \batcave ships with pre-computed tables for both human whole exomes and whole genomes.
-
-\subsection{Estimation of the mutation rate}
-The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
-In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
-\begin{equation}
-  \label{eqn:mut_rate}
-  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
-\end{equation}
-The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
-The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
-We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
-We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
-Sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
-To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p_{ur}$.
-All estimated allele frequencies are then multiplied by a factor of $1/p_{ur}$ before being used in Eq.~\ref{eqn:mut_rate}.
-We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
-In the R implementation of \batcave, $f_{min}$ and $p_{ur}$ are free parameters.
-For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
-For our simulated data analyses, we set $p_{ur}=1$, and for our real data analyses, we set $p_{ur}$ to the estimated purity.
-
-%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\cite{Williams2016}.
-%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
-
-\subsection{Likelihood function}
-The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq.
-\ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
-We used MuTect 1.1.7 for all analyses in this paper, so we have
-\begin{equation}
-  \label{eqn:tlod}
-    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
-\end{equation}
-The log posterior odds is the log likelihood ratio (\textrm{TLOD}) plus the log prior odds, so the posterior odds in favor of the alternate hypothesis for a given substitution type is
-\begin{equation}
-  \label{eqn:computed_posterior}
-  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))}.
-\end{equation}
-Here $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in Eq.~\ref{eqn:detailed_prior} and specified in Eq.~\ref{eqn:post_pred}-\ref{eqn:mut_rate}.
-When comparing our posterior odds to those of MuTect, we assume a uniform per-base probability of mutation of $3\cdot10^{-6}$ \cite{Cibulskis2013}, so
-\begin{equation}  \label{eqn:mutect_posterior}
-  \frac{\mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})}{1 - \mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6}.
-\end{equation}
-
-\subsection{Implementation}
-We have implemented the \batcave algorithm as an R package called \texttt{batcaver}.
-The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
-Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.
-
-An assumption of the \batcave algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
-To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
-If the two profiles differ at a significance threshold of $\alpha=0.05$, \texttt{batcaver} issues a warning and outputs the profiles.
-%The user can then assess them and follow-up on any biologically interesting differences.
-
-\begin{figure}
-\centering
-  \includegraphics{figures/sim_signatures_only.pdf}
-  \caption{Simulated tumor mutation profiles.
-As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
-  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
-  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
-  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
-   }
-  \label{NAR-simsigfig}
- \end{figure}
-
-\subsection{Tumor simulations}
-We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
-Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
-Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
-The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
-The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
-The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
-Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
-Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
-Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
-\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
-        --reference\_sequence \$ref\_path 
-        --dbsnp \$db\_snp 
-        --enable\_extended\_output 
-        --fraction\_contamination 0.00 
-        --tumor\_f\_pretest 0.00 
-        --initial\_tumor\_lod -10.00 
-        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
-        --input\_file:normal \$tmp\_normal 
-        --input\_file:tumor \$tmp\_tumor 
-        --out \$out\_path/\$chr.txt 
-        --coverage\_file \$out\_path/\$chr.cov}.\\
-Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
-Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.
-
-\subsection{Calibration metric}
-To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
-Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
-For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
-The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.
-
-\subsection{Real data}
-We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
-We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
-Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
-We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
-Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.
-Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
-For these multi-region data, we analyzed the three biological replicates from each of six tumors (Table~\ref{table:01}), running \batcave separately on each of the eighteen samples, and we combined results to generate precision-recall curves.
-Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.
-We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCh37, so we used that reference.
-
-\section{RESULTS}
-We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
-\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
-The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
-The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.
-
-To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
-The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}).
-We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
-In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
-Because our focus is on evaluating the statistical calling model, we computed all test metrics using only those potential variants that passed MuTect's heuristic filters and entered the statistical model.
-
-\begin{figure}
-\centering
-  \includegraphics{figures/kl_only.pdf}
-  \caption{Convergence of the mutational prior to the data generating distribution.
-Plotted is the Kullback-Leibler divergence between the simulated and estimated profiles versus number of incorporated mutations for whole exomes.
-Convergence for whole genomes is similar.
-  }
-\label{NAR-kl_fig}
-\end{figure}
-
-\subsection{Tests using simulated data}
-
-To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
-When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.
-For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.
-
-%We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
-%In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
-%For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the leukemia~\cite{Griffith2015} is over 17000, and for our simulated data sets ranges between 1500 and 5000.
-
-  \begin{figure*}
-\centering
-    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
-    \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and \batcave results with black lines.
-    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation profiles.
-    \textbf{(C)} and \textbf{(D)} Calibration plots.
-Shaded regions show distributions of posterior probabilities for true positive variants, and smooth lines show loess-smoothed relationships, from which the Integrated Calibration Index is calculated.
-For a perfectly calibrated caller, those curves would match the dashed y=x line.
-    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial mutation validation was performed \cite{Griffith2015,Shi2018}.
-}
-  \label{NAR-wes_fig}
-  \end{figure*}
-
-\begin{table*}[b]
-  \tableparts{%
-  \caption{Variant calling metrics for all data sets.}
-  \label{table:01}%
-  }{%
-  \begin{tabular}{lllllllllll}
-  \toprule
-  Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
-  \\
-  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
-  \\
-  \colrule
-  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
-  \\
-  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
-  \\
-  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
-  \\
-  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
-  \\
-  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
-  \\
-  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
-  \\
-  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .996 & .988 & -- & --
-  \\
-  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .972 & .968 & -- & --
-  \\
-  \botrule
-  \end{tabular}%
-  }
-  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
- \end{table*}
-
-We assessed classification performance using the areas under both the receiver operating characteristic and the precision-recall curves, because the classes are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
-By both metrics \batcave outperforms MuTect (Fig.~\ref{NAR-wes_fig}A\&B, Fig.~\ref{NAR-wgs_fig}A\&B, and Table~\ref{table:01}).
-The extent of the performance difference is dependent on both the sequencing depth and the concentration of the mutation profile.
-Deeper sequencing and more concentrated mutation profiles increase the performance advantage of \batcave.
-
-For all simulated tumors, the estimated mutation rate was approximately $3\cdot10^{-7}$ (Table~\ref{table:01}), which is lower than the simulated rate of $3\cdot10^{-6}$.
-This is likely due to restrictions within BAMSurgeon, such as sequencing depth and quality, that prevent 100\% of simulated variants from being inserted into the reads.
-
-We also assessed calibration, the likelihood that a potential variant with a given posterior probability is actually a true variant.
-We measured overall calibration performance using the Integrated Calibration Index (ICI) \cite{Austin2019}, which integrates the difference between predicted and observed probabilities, weighted by the density of the predicted probabilities.
-This metric is particularly useful in our case, because the density of posterior probabilities is bi-modal (Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D).
-A large fraction of true negative variants have posterior probabilities less than $10^{-4}$, far below any meaningful threshold, so we evaluated calibration only on potential variants with posterior probability greater than 0.01.
-For these potential variants, \batcave tends to increase posterior probabilities of low probability but true variants (density curves in Fig.~\ref{NAR-wes_fig}C\&D and~\ref{NAR-wgs_fig}C\&D) while decreasing probabilities of low probability but false variants.
-For 500X exomes, the calibration of \batcave is better than MuTect across the full spectrum of posterior probabilities (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
-For 100X whole genomes, the calibration of \batcave is slightly worse (Fig.~\ref{NAR-wgs_fig} and Table~\ref{table:01}), likely because there are few low probability true positive variants in tumors sequenced to 100X depth.
-As with the other metrics, the advantage of \batcave increases with the concentration of the mutation profile and the sequencing depth.
-
-\begin{figure}
-\centering
-  \includegraphics{figures/ppv_wes.pdf}
-  \caption{Posterior probability calibration for realistic calling thresholds, for 500X exomes.
-  Plotted is precision and recall for variants identified using various realistic posterior probability thresholds.
-At these thresholds, the precision of \batcave is much closer to the given threshold than MuTect, no matter the concentration of the mutation profile.
-  }
-\label{NAR-ppv_fig}
-\end{figure}
-
-In practice, variant callers are typically used with a threshold score above which a variant is called.
-The user's choice of threshold ideally meets their need to balance precision and recall; accurate posterior probability estimates enable an informed choice.
-For posterior probability thresholds between 60 and 90\%, the precision of \batcave calls is similar to the chosen threshold (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-For this range of thresholds, however, the posterior probabilities from MuTect poorly predict precision (Fig.~\ref{NAR-ppv_fig}\&\ref{NAR-ppv_wgs_fig}).
-For any posterior probability threshold above 70\%, MuTect has a false positive rate of roughly 8\%, whereas \batcave has a false positive rate that decreases as the threshold increases.
-The cost of MuTect's compressed range of posterior probabilities is recall; at any posterior probability threshold \batcave has recall better than MuTect.
-Consequently, \batcave posterior probabilities are more informative than MuTect's with regard to choosing a calling threshold.
-%The cost of MuTect's compression is recall, which is 3-5\% worse than \batcave at 95\% posterior probability, and 3-7\% worse at 90\%, depending on mutation profile concentration.
-%Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
-
-\subsection{Tests using real tumor data}
-We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the express purpose of evaluating tumor variant calling pipelines, yielding high quality true and false positive data~\cite{Griffith2015,Shi2018}.
-However, only variants called by at least one variant caller were validated.
-As a result, there are no validated true or false negative calls, so we considered only precision-recall comparisons for these data.
-
-Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
-We estimated a per-base mutation rate for this tumor of $3.6\times10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
-For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
-The low-confidence mutation profile contains additional spikes of mutations corresponding to five substitution types (Fig.~\ref{fig:AML_profiles}).
-Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.988 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).
-
-Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
-%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
-We estimated an average per-base mutation rate for these tumor regions of $6.6\times10^{-7}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
-For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})
-
-\section{DISCUSSION}
-
-%Deep sequencing is necessary to measure intra-tumor heterogeneity \cite{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.
-%Current approaches to very high depth next-generation sequencing \cite{Griffith2015}, however, require combining calls from multiple variant calling algorithms and then validating these calls via ultra-deep sequencing.
-%We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
-
-\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
-Our implementation is built on MuTect, one of the most widely used somatic variant callers.
-\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D,  \ref{NAR-wgs_fig}, and Table~\ref{table:01}) across the entire range of recall and specificity.
-Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (Fig.~\ref{NAR-ppv_fig} and~\ref{NAR-ppv_wgs_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
-For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F and Table~\ref{table:01}).
-The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.
-
-We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
-The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively.
-And the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
-But mutational profiles are highly heterogeneous, so concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
-The two real data sets we considered are among the few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
-They happen, however, to have diffuse mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
-A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
-This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall for variants that pass heuristic filters (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
-Moreover, few true negative sites are validated, so specificity and calibration are impossible to calculate.
-Deep sequencing experiments that validate random samples of uncalled potential variants would give much-needed insight into the differences among statistical models in variant calling.
-
-The improved calibration of \batcave posterior probabilities compared to MuTect provides several advantages.
-In practice, called variants are often manually reviewed to further reduce false positives \cite{Barnell2019}.
-Improved calibration enables users to focus review on the most questionable variants.
-In the clinic, identified variants act as biomarkers for susceptibility to targeted drugs \cite{Boutros2015a}.
-Well-calibrated posterior probabilities facilitate the use of probabilistic risk models in the choice of treatment \cite{Holmberg2013}, rather than an all or nothing approach.
-For research purposes, the International Cancer Genome Consortium recommends that catalogs of somatic mutations target a precision of 95\% and a recall of 80\% \cite{Icgc}.
-Achieving this goal while minimizing cost demands well-calibrated posterior probabilities.
-
-Our current implementation of \batcave is as a post-calling algorithm for MuTect, but the algorithm is broadly applicable.
-We chose to build \batcave off MuTect because MuTect is widely used, has state-of-the-art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Cibulskis2013,Griffith2015}.
-But the mutational prior can be incorporated into almost any caller with an underlying probabilistic model.
-For example, Strelka2 computes a joint posterior probability over tumor and normal genotypes, assuming a constant somatic mutation probability at each genomic site \cite{Kim2018}.
-Replacing that constant probability with a mutational prior would require a more complicated manipulation of the quality scores output by Strelka than for MuTect, but it is conceptually straightforward.
-
-Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
-In general, low-frequency variants are expected to be more common than high-frequency variants.
-Nevertheless, the uniform frequency prior is appropriate when the goal is to make the best call possible at each potential variant site.
-By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
-Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate~\cite{Nielsen2012}.
-
-To estimate the tumor mutation rate, \batcave uses a model of the distribution of variant allele frequencies  generated during exponential growth (Eq.~\ref{eqn:mut_rate}).
-Mutations that change copy number might distort the empirical allele frequency distribution, biasing the estimate.
-But substantially biasing the estimate would require copy number changes in a large number of genomic regions.
-Nevertheless, a potential extension of the \batcave algorithm would be to incorporate external copy number calls in order to adjust allele frequencies before estimating the mutation rate.
-
-Sequencing multiple regions of the same tumor is of growing interest for understanding tumor heterogeneity and evolution \cite{McGranahan2017}.
-In fact, the breast cancer data we analyze came from a multi-region study~\cite{Shi2018}.
-We called variants independently across samples, because MuTect only considers a single sample at a time.
-But calling variants jointly across samples could substantially increase sensitivity.
-In the future, the \batcave algorithm could be integrated into a multi-sample caller such as MultiSNV~\cite{Josephidou2015}.
-In that case, an optimal implementation would likely exploit the relationship between mutational priors in related samples.
-
-The \batcave algorithm is computationally inexpensive; our current implementation adds 1 second per 22,000 variants evaluated to a standard GATK best-practices variant calling pipeline.
-The majority of the computational cost is associated with extracting the trinucleotide context for each potential variant site from the reference genome.
-Since most callers are already walking the reference genome during the calling process, extracting the trinucleotide context simultaneously would virtually eliminate the computational cost of implementing a mutational prior.
-
-The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
-Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
-There are, however, many other aspects of genomic context that can affect local mutation rates \cite{Buisson2019}, including replication timing \cite{Stamatoyannopoulos2009}, expression level \cite{Pleasance2010}, and chromatin organization \cite{Schuster-Bockler2012}.
-Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutational prior using the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
-Others, such as expression level, could be tumor-specific, but would require information not available in the variant calls to compute.
-In the long run, we believe that incorporating more tumor biology into variant calling models will continue to improve performance.
-
-\batcave divides the data into two classes: high- and low-confidence variants.
-The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
-Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
-In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
-
-An  implicit assumption of \batcave is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
-Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
-Our \texttt{batcaver} software warns the user if the high- and low-confidence mutational profiles are statistically different.
-It also outputs those profiles, so the user can assess whether the difference will substantially affect their results.
-A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
-This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
-
-Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
-Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
-Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
-Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
-Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
-
-\section{Software Availability}
-
-The \texttt{batcaver} R package can be downloaded or installed from \texttt{http://github.com/bmannakee/batcaver}
-The version of \texttt{batcaver} used to generate results and all analysis code have been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471715}
-Python code used to generate simulated tumors has been preserved on Zenodo \texttt{https://doi.org/10.5281/zenodo.3471741}
-
-%\section{Data Availability}
-%COSMIC mutation signatures version 2 is available at \footnotesize{https://cancer.sanger.ac.uk/cosmic/signatures\_v2}.
-
-\section{ACKNOWLEDGEMENTS}
-
-This work was supported by the National Science Foundation via Graduate Research Fellowship award number DGE-1143953 to BKM and by the National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM127348 to RNG.
-We thank Prof.\ Edward J. Bedrick for fruitful discussions about the statistical model.
-This material is based upon High Performance Computing (HPC) resources supported by the University of Arizona TRIF, UITS, and RDI and maintained by the UA Research Technologies department.
-
-\subsubsection{Conflict of interest statement.} None declared.
-
-\bibliography{caller_paper}
-
-\beginsupplement
-\section{Supplementary Figures}
 
@@
   \includegraphics[width=\textwidth]{figures/fig_wgs.png}
-  \caption{Variant-calling performance on  simulated 100X whole genomes. As in Fig.~\ref{NAR-wes_fig}A-D, but for 100X whole genomes.}
+  \caption{Variant-calling performance on  simulated 100X whole genomes. As in Fig.~4A-D, but for 100X whole genomes.}
 \label{NAR-wgs_fig}
@@
   \includegraphics{figures/ppv_wgs.pdf}
-  \caption{Posterior probability calibration for realistic calling thresholds, for 100X whole genomes. As in Fig.~\ref{NAR-ppv_fig}, but for 100X whole genomes.}
+  \caption{Posterior probability calibration for realistic calling thresholds, for 100X whole genomes. As in Fig.~5, but for 100X whole genomes.}
 \label{NAR-ppv_wgs_fig}
@@
 
-\clearpage
+\begin{table*}[h]
+  \caption{Sample summary for Shi et al. (2018).   }
 
-\section{Supplementary tables}
+ \centering
+  \label{table:01}%
+  \begin{tabular}{l|c|c}
+  Sample & Purity estimate (\%) & Mutation rate estimate
 
-\begin{table*}[h]
-  \tableparts{%
-  \caption{Sample summary for Shi et al. (2018)}
-  \label{table:01}%
-  }{%
-  \begin{tabular}{lll}
-  \toprule
-  Sample & Purity estimate (\%) & Mutation rate estimate 
   \\
-  \colrule
+  \hline
   Case 1 biorep A & 53.8 & 9.3e-7 
@@
   \\
-  Case 2 biorep C & 40(*) & 1.0e-6
+  Case 2 biorep C & 40$^*$ & 1.0e-6
   \\
@@
   \\
-  \botrule
-  \end{tabular}%
-  }
-  {Shi et al. did not estimate purity for Case 2 biorep C. We chose 40\% purity, based on the other biological replicates from this case.}
+  \end{tabular}
+
+{  \small$^*$Shi et al. did not estimate purity for Case 2 biorep C. We chose 40\% purity, based on the other biological replicates from this case.}
 \end{table*}
